### 2017-03-31 paper/paper.tex
Added paper to repo

\documentclass{article}
\usepackage{times}
\usepackage{graphicx}
\usepackage{subfigure} 
\usepackage{natbib}
\usepackage{algorithm}
\usepackage[nice]{nicefrac}
\usepackage{algorithmic}
\usepackage{hyperref}
\newcommand{\theHalgorithm}{\arabic{algorithm}}
\usepackage{icml2017}
%\usepackage[accepted]{icml2017}

\def\bsa{{\boldsymbol\alpha}}
\def\bse{{\boldsymbol\epsilon}}
\def\bsk{{\boldsymbol\kappa}}
\def\bsm{{\boldsymbol\mu}}
\def\bsl{{\boldsymbol\lambda}}
\def\bsph{{\boldsymbol\phi}}
\def\bsth{{\boldsymbol\theta}}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{amsfonts}
\usepackage{enumerate}
\usepackage{fullpage}
\usepackage{bbm}
\usepackage{caption}
\usepackage{mathrsfs}

\newcommand{\pa}[1]{ \left({#1}\right) }

\def \N {\mathbb{N}}
\def \Nbr {\mathcal{N}}
\def \mcL {\mathcal{L}}
\def \Q {\mathbb{Q}}
\def \F {\mathbb{F}}
\def \then {\implies &}
\def \oif {\Longleftrightarrow &\,}
\def \given {\text{Given }&}
\def \assume {\text{Assume }&}
\def \thfr {\therefore &\enskip}
\def \bij {\leftrightarrow}
\def \inj {\rightarrowtail}
\def \sur {\twoheadedrightarrow}
\def \Z {\mathbb{Z}}
\def \R {\mathbb{R}}
\def \C {\mathbb{C}}
\def \T {\mathbb{T}}
\def \iff {\Longleftrightarrow}
\def \kron {\boldsymbol\delta}
\def \indicator {\mathbbm{1}}

\def\Tx{\textbf{x}}
\def\TX{\textbf{X}}
\def\TU{\textbf{U}}
\def\Ty{\textbf{y}}
\def\tbr{\textbf{r}}
\def\Tz{\textbf{z}}
\def\quotient{\mathclose{}/\mathopen{}}
\def\Tf{\textbf{f}}
\def\Th{\textbf{h}}
\def\Tg{\textbf{g}}
\def\sumn{\sum_{n=0}^\infty}
\def\limn{\lim_{n\rightarrow\infty}}
\def\prodn{\prod_{n=0}^\infty}

\DeclareMathOperator\adj{adj}
\DeclareMathOperator\cov{cov}
\DeclareMathOperator\Unif{Unif}
\DeclareMathOperator\blockdiag{blockdiag}
\DeclareMathOperator\Pois{Pois}
\DeclareMathOperator\mathProb{\mathbb{P}}
\def\P{\mathProb} % need to overwrite stupid paragraph symbol
\DeclareMathOperator\mathExp{\mathbb{E}}
\def \E {\mathExp}

\newcommand{\stc}[1]{\widetilde{#1}}   
\newcommand{\set}[2]{ \left\{ #1 \,\middle|\, #2 \right\} }
\newcommand{\shift}[1]{&\quad & \text{#1}\\}
\newcommand{\lem}[1]{\text{\textbf{L.\ref{#1}}}}
\newcommand{\card}[1]{\left\vert{#1}\right\vert}
\newcommand{\Ps}[1]{\mathcal{P}\left({ #1 }\right)}
\newcommand{\colv}[1]{\begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\mat}[1]{\begin{pmatrix} #1 \end{pmatrix}}
\newcommand{\detmat}[1]{\begin{vmatrix} #1 \end{vmatrix}}
\newcommand{\spanb}[1]{\text{span}\{ #1 \}}
\newcommand{\abs}[1]{\left|#1\right|}
\newcommand{\Inner}[1]{\langle #1 \rangle}
\newcommand{\Innercpy}[1]{\langle #1, #1 \rangle}
\newcommand{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$

\def\eqd{\mathrel{\overset{\Delta}{=}}}

\DeclareMathOperator{\diag}{diag}
\DeclareMathOperator{\vcdim}{VC-dim}
\DeclareMathOperator*{\Err}{\text{err}}
\DeclareMathOperator*{\ErrE}{\mathbb{E}}
\DeclareMathOperator{\Tr}{tr}
\DeclareMathOperator{\Dim}{dim}
\DeclareMathOperator{\Rank}{rank}
\DeclareMathOperator*{\argmin}{argmin}
\DeclareMathOperator*{\proj}{proj}
\DeclareMathOperator{\Ker}{ker}
\DeclareMathOperator{\Diam}{diam}
\DeclareMathOperator{\Int}{int}
\DeclareMathOperator{\Clo}{clo}
\DeclareMathOperator{\sgn}{sgn}
\DeclareMathOperator{\MyRe}{Re}
\DeclareMathOperator{\MyIm}{Im}
\DeclareMathOperator{\Supp}{supp}
\DeclareMathOperator{\Var}{Var}

% probability stuff
\def \sa {{$\sigma$-algebra}}
\def\OR{{\overline{\R}}}
\def\OX{{\overline{X}}}

\def \mcX {\mathcal{X}}
\def \mcY {\mathcal{Y}}
\def \mcH {\mathcal{H}}
\def \mcD {\mathcal{D}}
\def \mcA {\mathcal{A}}
\def \mcK {\mathcal{K}}
\def\mcG{{\mathcal{G}}}
\def\mcH{{\mathcal{H}}}
\def\mcF{{\mathcal{F}}}
\def\mcB{{\mathcal{B}}}
\def\mcE{{\mathcal{E}}}
\def\mcP{{\mathcal{P}}}
\def\mcM{{\mathcal{M}}}
\def\mcC{{\mathcal{C}}}
\def\mcT{{\mathcal{T}}}

% \icmltitlerunning{Submission and Formatting Instructions for ICML 2017}

\begin{document}

\twocolumn[
\icmltitle{Large Linear Multi-output Gaussian Process Learning}
%\icmlsetsymbol{equal}{*}
\begin{icmlauthorlist}
  \icmlauthor{Vladimir Feinberg}{pu}
  \icmlauthor{Li-Fang Cheng}{pu}
  \icmlauthor{Barbara E. Engelhardt}{pu}
  \icmlauthor{Kai Li}{pu}
\end{icmlauthorlist}
\icmlaffiliation{pu}{Princeton University, Princeton, USA}
\icmlcorrespondingauthor{Vladimir Feinberg}{vyf@princeton.edu}

\icmlkeywords{multi-task, multi-output, Gaussian process, large scale}

\vskip 0.3in
]
\printAffiliationsAndNotice{}
%\printAffiliationsAndNotice{\icmlEqualContribution}

\begin{abstract}
Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}. Such models can learn correlations across the multiple outputs, such as temperature data from disparate regions, which can then be exploited to share knowledge across the multiple outputs. While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large $n$, the total number of observations across all outputs. In this work, we describe a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs. We show results of our approach applied to real time series data, and compare these results against related methods. Finally, we discuss extensions of our approach to multidimensional inputs.
\end{abstract}

\section{Introduction}
\label{introduction}

Gaussian processes (GPs) are a popular nonlinear regression method that innately capture function smoothness across inputs as defined by their covariance function \cite{williams1996gaussian}. GPs seamlessly extend to multi-output learning, picking up on latent cross-output processes, where in multi-output learning the objective is to build a probabilistic regression model over vector-valued observations.

Such an extension is especially important in time series data, where multi-output GPs have been applied to the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.

Exact GP inference comes at a cost: to compute the gradients for kernel hyperparameters, an exact computation requires an $O(n^3)$ decomposition of an $O(n^2)$ matrix in memory for $n$ samples \cite{gpml}. Each hyperparameter's scalar partial derivative then requires another $O(n^2)$ computation. While these operations may be performed in parallel, the memory use and runtime is prohibitive for problems with a large number of outputs, since the number of hyperparameters grows accordingly. For this reason, an accurate approximate approach has been an important goal of machine learning research.

For a single output, variational approaches that work on $m\ll n$ inducing points, or locations at which the true kernel is sampled, letting users trade off accuracy for efficiency \cite{hensman2013gaussian}.

Nguyen and Bonilla observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once \yrcite{nguyen2014collaborative}. We analogously exploit sharing in the covariance structure.

\subsection{Contributions}

Our approach makes the following contributions to approximate inference in multi-output GPs. First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid. Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}. Both of these structures coordinate for fast matrix-vector multiplication $K\Ty$ with the covariance matrix $K$.

When inputs do not lie on a uniform grid (i.e., for time series, the length of time between observations differs), we apply a technique proposed in Massively Scalable Gaussian Processes (MSGP) \cite{msgp}. For single-output kernels, MSGP leverages the structured kernel interpolation (SKI) framework to make inputs act as if they are on a grid \cite{kiss-gp}. In a single-output case, this opens up the possibility for a matrix-free GP, avoiding construction of the explicit Gram covariance matrix. Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI. While direct application of SKI is not possible, we can make an extension that makes matrix-free LMC kernel learning viable even on non-grid inputs.

Our paper is organized as follows. In Section~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs. Section~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs. Section~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion. Then, in Section~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Section~\ref{conclusion}.

\section{Background}
\label{sec:background}
\subsection{General Gaussian processes}

First, we give an overview of the theoretical foundation for GPs. A GP is a set of random variables (RVs) $\{f_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that for any finite collection $X\subset\mcX$, the RVs are jointly Gaussian with a prescribed mean function $\bsm:\mcX\rightarrow\R$ and covariance $K:\mcX^2\rightarrow\R$, i.e., $f_X\sim N\pa{\bsm_X, K_{X, X}}$ \cite{williams1996gaussian}. As usual, we drop the mean with $\Ty\triangleq f_X-\bsm_X$ and denote vectorization $f_X=\colv{f_{\Tx}}_{\Tx\in X}$ or matricization $K_{X,Z}=\mat{K(\Tx,\Tz)}_{\Tx\in X,\Tz\in Z}$ with subscripts.

Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by marginalization \cite{gpml}:
\begin{equation*}
y_*|\Ty\sim N\pa{K_{*, {X}}K^{-1}_{{X}, {X}}\Ty, K_{*, *}-K_{*, {X}}K^{-1}_{{X}, {X}}K_{{X}, *}}.
\end{equation*}
Accuracy is often sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
\begin{align}
  \mcL(\bsth)&=\log p(\Tf_{X}|X,\bsth)\label{datall}\\
             &=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.
\end{align}
As such, the heart of GP learning lies in optimization of $\mcL$. Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
\begin{align}
\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}}.\label{llgradient}
\end{align}

\subsection{Multi-output linear GPs}

We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2011kernels}.

Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$: an input is in a point from a shared domain coupled with an output tag. Then, if we make observations at $X_d\subset\mcX$ for output $d\in[D]$, we can set:
\begin{align*}
\TX=\set{(d, x)}{d\in[D],x\in X_d}\subset{\mcX'};\;\; n=\card{\TX}.
\end{align*}

An LMC kernel $K$ is of the form 
\begin{align}
K([i,\Tx_i],[j,\Tx_j])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx_i-\Tx_j})+\epsilon_i\delta_{ij},\label{lmcpointwise}
\end{align} 
where $k_q:\R\rightarrow\R$ is a stationary kernel on $\mcX$. Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\bsk_q I_D$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a pre-set rank. Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz. Settings of $B_q$ that arise in practice, where we wish to capture covariance across outputs as an arbitrary $D^2$-dimensional latent process,  prevent stationarity in $K$ and therefore prohibit direct application of SKI.

\subsection{Structured covariance matrices}

If we can identify structure in the covariance matrices $K$, then we can recover fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$.

The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$-th block $A_{ij}B$. We can represent it by keeping representations of $A$ and $B$ separately, rather than the product. Then, the corresponding MVMs can be computed in time $O(a\text{MVM}(B)+b\text{MVM}(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM. For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.

Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_j$, yielding an $O(n)$ representation. Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs. Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that:
\[
T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}
\]
Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs  $C\mat{\Tx\\\textbf{0}}=\mat{T\Tx\\\textbf{0}}$ in $O(n\log n)$ time. This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.

\section{Related Work}
\label{sec:related-work}
\subsection{Approximate inference methods}

To cope with the lack of tractable GP inference methods, inducing point approaches create a tractable model to use instead of the classic GP. Such approaches fix or estimate inducing points $\TU$ and claim that the data $f_X$ is conditionally deterministic (deterministic training conditional, or DTC), independent (fully independent training conditional, or FITC), or partially independent (partially independent training conditional, or PITC) given RVs $f_{\TU}$  \cite{quinonero2005unifying}. These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime. Setting the inducing points $\TU=\TX$ recovers the original exact GP.

In the context of LMC, a framework for variational DTC, DTCVAR, provided a matrix-free method where each iteration has complexity $O(n(Dm)^2)$, where $m$ is the average number of inducing points per output, $\card{\TU}/D$ \cite{alvarez2010efficient}.

By sharing the $m$ inducing points across all outputs, Collaborative Multi-output Gaussian Processes (COGP) improves runtime per iteration to $O(m^3)$. Moreover, COGP only requires $O(Dnm)$ memory \cite{nguyen2014collaborative}. To ensure that $O(m^3)$ is computationally tractable for COGP, $m$ must be sufficiently small. COGP makes large multi-output GP estimation feasible through a variational approximation, the evidence lower bound, to the log likelihood.

\subsection{Approaches for stationary kernels}

\subsubsection{Structured Kernel Interpolation (SKI)}\label{ski-section}

SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \cite{kiss-gp}. To do this efficiently, SKI relies on the differentiability of $K$. For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$. The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$. Importantly, $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.

In order to adapt SKI to our context of multiple outputs, we build grid $\TU\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with $\TU = [D]\times U$. Since the LMC kernel evaluated between two sets of outputs $K_{X_i,X_j}$ is differentiable, as long as $U$ contains each $\{X_d\}_{d\in[D]}$, the corresponding SKI approximation $K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top$ holds with the same asymptotic convergence cubic in $1/m$.

Massively Scalable Gaussian Processes (MSGP) observes that the kernel $K_{U,U}$ on a grid exhibits Kronecker and Toeplitz matrix structure \cite{msgp}. Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{datall}. In addition, \cite{wilson2014fast} mentions an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{datall}.

While evaluating $\log\abs{K_{|\bsth}}$ is not feasible in the LMC setting (because the LMC sum breaks Kronecker and Toeplitz structure), the general notion of creating structure with SKI carries over to LLGP.

\section{Matrix-free LMC Learning}
\label{sec:matrix-free}

We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches. Critically, the accuracy of the method need not be reduced by keeping $m$ low because its reliance on structure allows better asymptotic performance.
For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.

Although this is a method of optimizing the marginal log likelihood, it can easily be augmented with a hyperprior $\log p(\bsth)$ term to Eq.~\ref{datall}.% what does "augmented with a hyperprior $\log p(\bsth)$ term to Eq.~\ref{datall}" mean? make this explicit here. a log probabliity is not a hyper prior (generally)

\subsection{Matrix-free GP Learning}

Gibbs and MacKay describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix \yrcite{gibbs1996cient}. In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG). Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
\begin{align}
  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
\end{align}
For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$.

We depart from Gibbs and MacKay in selecting MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs. MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}. This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations (Fig.~\ref{fx2007-iterations}).

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{iterations.eps}}
\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Section~\ref{fx2007-results}. The iteration cutoff is the number of training points $n$ in the dataset.
} % TODO no time right now but later could compare to LCG MVMs.
\label{fx2007-iterations}
\end{center}
\vskip -0.2in
\end{figure}

If we sample $N_t$ instances of $\tbr$, every iteration needs to perform $N_t+1$ inversions including $\Ty$, if we reuse the same samples $K_{|\bsth}^{-1}\tbr$ between derivatives $\partial_{\theta_j}$. These inversions are done in parallel. Then, since $\partial_{\theta_j}K_{|\bsth}$ is both structured and smaller than $K_{|\bsth}$, each of the terms in the approximation Eq.~\ref{eq:trace} require $\tilde{O}(n)$ time to compute.

Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine \cite{zeiler2012adadelta}. We considered several stochastic approximations for finding $\log\abs{K_{|\bsth}}$~\cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.

In other words, all the work is done in the inversions $K_{|\bsth}^{-1}\tbr,K_{|\bsth}^{-1}\Ty$, which in turn rely on the MVM operation, discussed in Section~\ref{fast-mvm}. Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense. % this is a good transition paragraph, and motivates what follows well, but perhaps reworking the structure of the prior sections will make this (for lack of a better word) pipeline bottleneck mroe obvious? TODO any suggestions?

\subsection{Fast MVMs and parsimonious kernels}\label{fast-mvm}

When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$.
\begin{align}
  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}
\end{align}
The SLFM and COGP models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$. Moreover, we include $D$ additional kernels for the independent GP components, which can be incorporated using kernels $K_d$ where $A_d=0$ and $\bsk_d$ as the $d$th standard basis vector of $\R^D$ for $d\in[D]$. This shows a reduction from SLFM to LMC.

Importantly, the Kronecker structure of Eq.~\ref{lmcgrid} lets us re-use the same grid $K_q$ in a computational sense across the different outputs. Recalling our SKI extension to multiple outputs (Section~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
\begin{align}
  K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top+\bse.\label{lmcski}
\end{align}
We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel, so the SKI approximation fails. But this fact does not prevent efficient representation or multiplication since the matrix is diagonal. In particular, the MVM operation $K_{\TX,\TX}\Tx$ can be approximated by $WK_{\TU,\TU}W^\top\Tx+\bse\Tx$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.

We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$ (where we have computed $\Tz=W^\top\Tx$).

% TODO get rid of all the linear terms below; delay big-O discussion until "asymptotics"?

\subsubsection{\textsc{sum}: sum representation}

In \textsc{sum}, we represent $K_{\TU,\TU}$ with a $Q$-length list. At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, where only its top row needs to be represented to perform MVMs.

In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results, corresponding to . As described before, the Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$. In turn, the runtime for each of the $Q$ terms is $O(Dm\log m)$.

\subsubsection{\textsc{bt}: block-Toeplitz representation}

In \textsc{bt}, we notice that $K_{\TU,\TU}$ is a block matrix with blocks $T_{ij}$:
\begin{align*}
\sum_qB_q\otimes K_q =\mat{T_{ij}}_{i,j\in[D]^2},\;\; T_{ij}=\sum_qb_{ij}^{(q)}K_q.
\end{align*}
On a one-dimensional grid $U$, these matrices are Toeplitz since they are linear combinations of Toeplitz matrices. \textsc{bt} requires $D^2$ $m$-sized rows to represent each $T_{ij}$. Then, using usual block matrix multiplication, an MVM $K_{\TU,\TU}\Tz$ takes $O(D^2m\log m)$ time.

\subsubsection{\textsc{slfm}: SLFM representation}

\textsc{slfm} uses a rank-based representation. Let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average added rank, $R\le D$.

We first rewrite the grid kernel:
\begin{align*}
  K_{U,U}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
\end{align*}
Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank-1. Under some re-indexing $q'\in[RQ]$ which flattens the double sum such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
\begin{align*}
  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top;
\end{align*}
where the second simplification has $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}. Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.

Thus, the \textsc{slfm} representation writes $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix, so MVMs take $O((QR + D)m\log m)$ time.

\subsection{Asymptotic performance}

Because the runtimes of \textsc{bt} and \textsc{slfm} are complimentary in the sense that one performs better than the other when $D^2>QR$ and vice-versa, an algorithm that uses the aforementioned condition between to decide between which representation to use can minimize runtime. We also found that \textsc{sum} is efficient in practice for $Q=1$.

Using the switching condition, MVMs with the original matrix $K_{\TX,\TX}$ altogether have a space and time upper bound of $\tilde{O}(\min(QR,D^2)m+n)$, where the $\min$ is earned thanks to the choice between different representations. Every AdaDelta iteration then takes $\tilde{O}(\sqrt{\kappa_2})$ such matrix-vector products for machine-precision answers, with $\kappa_2$ the spectral condition number \cite{raykar2007fast}.

On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$. In this case, using \textsc{bt} alone leads to a faster algorithm --- applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.

\subsection{Stopping conditions}\label{stopping}

For a gradient-only stopping heuristic, we maintain the running maximum gradient $\infty$-norm. If gradient norms drop below a pre-set proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate. For example, when applied to the foreign exchange rate prediction (FX2007) dataset in Section~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{running_cutoff.eps}}
\caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
\label{fx2007-stop}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Prediction}

The predictive mean can be computed in $O(1)$ time as observed in \cite{msgp} by $K_{*,X}\bsa\approx W_{*,\TU}K_{\TU,\TU}\bsa$.

The full predictive covariance estimate requires finding a new term $K_{*,X}K_{X,X}^{-1}K_{X,*}$. This is done by solving the linear system in a matrix-free manner on-the-fly; in particular, $K_{X,X}^{-1}K_{X,*}$ is computed via MINRES for every new test point $K_{X,*}$. Over several test points, this is embarrassingly parallelizable.

A more efficient predictive variance algorithm is outside the scope of this paper: for a research setting, we expect training time to be the bottleneck. Moreover, one can couple LLGP learning with other prediction mechanisms. One example is the sampling-based approach proposed in \cite{papandreou2011efficient}, which extends to linear combinations of kernels that allow fast eigendecompositions.

\subsection{Hyperparameter Settings}

AdaDelta parameters were set to the following on all tests: rate $=1$, decay $=0.9$, momentum $=0.5$, offset $=0.0001$. The stopping criterion parameters permit the gradient $\infty$-norm to drop below a threshold of its maximum value so far a small, fixed number of times, 5. The maximum number of iterations was 100.

For learning, we initialize entries $A_q$ according to a unit normal and all $\bsk_q$ to $1$. Note that COGP initializes the coregionalization matrix to $1$ uniformly. Like COGP, we initialize noise to $0.1$ for all outputs.

\section{Results} % TODO read these over
\label{sec:results}
We evaluate the methods on held out data by using standardized mean square error (SMSE) of the test points with the predicted mean, and the negative log predictive density (NLPD) of the Gaussian likelihood of the inferred model. Notably, NLPD takes confidence into account, while SMSE only evaluates the mean prediction. In both cases, lower values represent better performance.

\subsection{Representation evaluation}\label{empirical-rep}

Here, experiments were carried out on ``random'' parameters with the same seed. % im not sure exactly what that means ("random parameters")? same seed? what is an experiment? might call it an application to simulated data? if we are desperate for space, make this a single sentence... instead of an itemized list?
In particular, we have the following parameters:
\begin{itemize}
\item $n$ total sample size across all outputs
\item $D$ number of outputs
\item $Q$ number of kernels $k_q$
\item $R$ average added rank
\item $\epsilon$ mean noise
\item \texttt{ktype} kernel type, one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions. \texttt{mix} refers to a mix of all three kernels.
Inverse length scales and periods for the kernels varied uniformly in log space from 1 to 10 with $Q$ samples.
\end{itemize}

Given this parameterization, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$. Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.

As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$ --- this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.

\begin{table}[!ht]
  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Section~\ref{empirical-rep} using MINRES for each of the kernel representations. For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system. We averaged over five runs. In every run, we use $n=5,000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
}
\label{complement}
\vskip 0.15in
\begin{center}
\begin{small}
  \begin{sc}
\begin{tabular}{|ccc|cccc|}
  \hline
  \abovespace\belowspace
  $D$ & $R$ & $Q$ & {chol} & {sum} & {bt} & {slfm}\\
\hline
  \abovespace
 $ 2 $ & $ 2 $ & $ 10 $ & $ 40.71 $ & $ 38.13 $ & $ \textbf{7.83} $ & $ 59.53 $ \\ 
 $ 10 $ & $ 1 $ & $ 10 $ & $ 37.69 $ & $ 33.15 $ & $ 16.39 $ & $ \textbf{11.59} $ \\ 
 $ 10 $ & $ 10 $ & $ 1 $ & $ 11.19 $ & $ \textbf{0.784} $ & $ 2.458 $ & $ 1.138 $

  \belowspace \\

  \hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

We next evaluated the accuracy of the gradients for $N_t=10$ trace samples. Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$. Since, for each partial derivative, LLGP requires only $N_t$ $n$-sized vector dot products (Eq.~\ref{eq:trace}), it generally runs faster than the exact approach (Fig.~\ref{fig:gradtime}), which must compute a matrix-matrix dot product (Eq.~\ref{llgradient}). The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}). Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor MINRES convergence in the inversions (Fig.~\ref{fig:inverr}). We will show that the stochastic gradients suffice for optimization in practical examples.

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relgrad_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and LLGP gradients. Higher is better, and extremal values are noted.  For each data point, the average was taken over five runs.}
\label{fig:relgrad}
\end{center}
\vskip -0.2in
\end{figure}

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{time_ratio.eps}}
\caption{Logarithm of the ratio of time required by the exact approach to that of LLGP for computing the gradient over all parameters from Eq.~\ref{llgradient}. Higher is better, and extremal values are noted. For each data point, the average was taken over five runs.}
\label{fig:gradtime}
\end{center}
\vskip -0.2in
\end{figure}

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{relalpha_l2.eps}}
\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and MINRES solutions to $K\Tx=\Ty$ for $\Tx$. For each data point, the average was taken over five runs.}
\label{fig:inverr}
\end{center}
\vskip -0.2in
\end{figure}

\subsection{Foreign exchange rate prediction (FX2007)}\label{fx2007-results}

We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance. The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \texttt{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below. The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP. COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP. All kernels are RBF.

The data used in this example are from 2007, and include $n=3,054$ training points and $150$ test points. The test points include 50 contiguous points extracted from each of the CAD, JPY, and AUD exchanges.

For this application, LLGP uses $m=n/D=238$ interpolating points. We use the COGP settings from the paper.\footnote{COGP hyperparameters for FX2007 were 100 inducing points, 500 iterations, 200 mini-batch size.} LLGP, for both LMC, outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).

\begin{table}[!ht]
  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset. Parenthesized values are standard error. LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points. COGP used a $Q=2$ kernel with $100$ inducing points}
\label{fx2007-tbl}
\vskip 0.15in
\begin{center}
\begin{small}
\begin{sc}
\begin{tabular}{|l|cc|}
\hline
  \abovespace\belowspace
  Metric & LLGP Q1R2 & COGP\\
\hline
\abovespace
  SMSE & $0.2107$ & $0.2444$ \\
  NLPD & $-3.6074$ & $10.9142$ \\
  Seconds & $65$ ($3$) & $291$ ($3$)
\belowspace\\
\hline
\end{tabular}
\end{sc}
\end{small}
\end{center}
\vskip -0.1in
\end{table}

\begin{figure*}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
\caption{Test outputs for the FX2007 dataset. COGP mean is black, with 95\% confidence intervals shaded in grey. LLGP mean is a solid red curve, with light green 95\% confidence intervals. Magenta points are in the training set, while blue ones are in the test set.} % TODO much better -- might also have 10 words on high level point: Predictions from LLGP on the test set are uniformly better across these three tasks. or something.
\label{fx2007-graph}
\end{center}
\vskip -0.2in
\end{figure*}

\subsection{Weather dataset}\label{large-bench}

Next, we replicate results from a weather dataset, a large time series used to validate COGP. In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure. Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15,789$ training measurements and $374$ test measurements.

We use the COGP parameters that were set by default in the code provided by the authors.\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}. The weather dataset was run on 1500 iterations, mini-batch size 1000.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model. We tested LLGP models on different numbers of interpolating points $m$. Both LLGP and COGP were run $10$ times, and average results are reported.

LLGP performed slightly worse than COGP in SMSE: its SMSE is $0.0907$ with compared to COGP's $0.0812$. However, both NLPD and runtime indicate significant improvements. The COGP result on $200$ inducing points had an NLPD of $99.7$ and runtime of $1342$ seconds. Varying the number of interpolating points $m$ from $400$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}). While NLPD improvement diminishes as $m$ increases, LLGP is still an improvement compared to COGP for a range of $m$ by an order of magnitude in runtime and almost two orders of magnitude for NLPD.

\begin{figure}[!ht]
\vskip 0.2in
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{m_time_nlpd.eps}}
\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 50 times.}
\label{fig:llgpweather}
\end{center}
\vskip -0.2in
\end{figure}
\section{Conclusion}\label{conclusion}

LLGP recovers speedups from SKI \cite{kiss-gp} for the problem of multi-output GP regression by recognizing structure unique to LMC kernels, and otherwise not necessarily recoverable in general multi-output kernels. This structure further enables a parsimonious representation that allows even large GPs to be learned without explicit construction of the covariance matrix.

LLGP provides a means to approximate the log-likelihood function gradients through interpolation. We have shown on several datasets that this can be done in a way that is faster and leads to more accurate results than variational approximations.

Future work would extend the inputs to accept multiple dimensions. This can be done without losing internal structure in the kernel \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB). The cubic interpolation requires and exponential number of terms, so projection into lower dimensions learned in a supervised manner would be essential. Another useful line for investigation would be more informed stopping heuristics. Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \cite{cutajar2016preconditioning}.

\bibliography{paper}
\bibliographystyle{icml2017} % TODO - follow blei notes for this

\end{document}

### 2017-03-31 benchmarks/benchlib/standard_tester.py
no newline in LaTeX table for dataset benchmarks

@@ -360,6 +360,9 @@ def latex_table(filename, cols, col_results):
         row = ' & '.join([metric] + row) + r'\\' + '\n'
         latex += row
 
+    # drop the last \\\n
+    latex = latex[:-3]
+
     colfmt = '|l|' + 'c' * ncols + '|'
     begin = (r'\begin{tabular}{' + colfmt + r'}'
              r'\hline\abovespace\belowspace' + '\n'

### 2017-03-31 benchmarks/representation-cmp/run.sh
latex no newline in representation-cmp + rerun

@@ -132,4 +132,4 @@ epilog='  \\belowspace \\\\
   \\hline
 \\end{tabular}
 '
-printf "${latex}\n${epilog}" > results.tex
+printf "${latex::${#latex}-6}\n${epilog}" > results.tex

### 2017-03-31 paper/paper.tex
added paper auxiliary files (now compiles)

@@
   \begin{small}
-  \begin{sc}
-\begin{tabular}{|ccc|cccc|}
-  \hline
-  \abovespace\belowspace
-  $D$ & $R$ & $Q$ & {chol} & {sum} & {bt} & {slfm}\\
-\hline
-  \abovespace
- $ 2 $ & $ 2 $ & $ 10 $ & $ 40.71 $ & $ 38.13 $ & $ \textbf{7.83} $ & $ 59.53 $ \\ 
- $ 10 $ & $ 1 $ & $ 10 $ & $ 37.69 $ & $ 33.15 $ & $ 16.39 $ & $ \textbf{11.59} $ \\ 
- $ 10 $ & $ 10 $ & $ 1 $ & $ 11.19 $ & $ \textbf{0.784} $ & $ 2.458 $ & $ 1.138 $
-
-  \belowspace \\
-
-  \hline
-\end{tabular}
-\end{sc}
+    \input{representation.tex}
 \end{small}
@@
     \begin{sc}
-\begin{tabular}{|l|cc|}
-\hline
-  \abovespace\belowspace
-  Metric & LLGP Q1R2 & COGP\\
-\hline
-\abovespace
-  SMSE & $0.2107$ & $0.2444$ \\
-  NLPD & $-3.6074$ & $10.9142$ \\
-  Seconds & $65$ ($3$) & $291$ ($3$)
-\belowspace\\
-\hline
-\end{tabular}
+      \input{results_fx2007.tex}
 \end{sc}
@@
 We tested LLGP models on different numbers of interpolating points $m$.
-Both LLGP and COGP were run $10$ times, and average results are reported.
 
-LLGP performed slightly worse than COGP in SMSE: its SMSE is $0.0907$ with compared to COGP's $0.0812$.
-However, both NLPD and runtime indicate significant improvements.
-The COGP result on $200$ inducing points had an NLPD of $99.7$ and runtime of $1342$ seconds.
-Varying the number of interpolating points $m$ from $400$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}).
+\begin{table}[!ht]
+  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the weather dataset.
+Parenthesized values are standard error.
+Both LLGP and COGP trained the SLFM model.
+We show LLGP with $500$ and $1000$ interpolating points and COGP with $200$ inducing points.}
+\label{weather-tbl}
+\vskip 0.15in
+\begin{center}
+  \begin{small}
+    \begin{sc}
+      \input{results_weather.tex}
+\end{sc}
+\end{small}
+\end{center}
+\vskip -0.1in
+\end{table}
+
+LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tbl.~\ref{weather-tbl}).
+Varying the number of interpolating points $m$ from $500$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}).
 While NLPD improvement diminishes as $m$ increases, LLGP is still an improvement compared to COGP for a range of $m$ by an order of magnitude in runtime and almost two orders of magnitude for NLPD.
@@
 \centerline{\includegraphics[width=\columnwidth]{m_time_nlpd.eps}}
-\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 50 times.}
+\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
 \label{fig:llgpweather}

### 2017-03-31 paper/copyresults.sh

@@ -1,0 +1,31 @@
+#!/bin/bash
+
+echo 'iterations.eps'
+cp ../benchmarks/picture-fx2007/out/iterations.eps .
+
+echo 'running_cutoff.eps'
+cp ../benchmarks/picture-fx2007/out/running_cutoff.eps .
+
+echo 'relgrad_l2.eps'
+cp ../benchmarks/grad-grid/out/relgrad_l2.eps .
+
+echo 'time_ratio.eps'
+cp ../benchmarks/grad-grid/out/time_ratio.eps .
+
+echo 'relalpha_l2.eps'
+cp ../benchmarks/grad-grid/out/relalpha_l2.eps .
+
+echo 'fx2007graph.pdf'
+cp ../benchmarks/picture-fx2007/out/fx2007graph.pdf .
+
+echo 'm_time_nlpd.eps'
+cp ../benchmarks/weather/out/m_time_nlpd.eps .
+
+echo 'representation-cmp/out/results.tex -> representation.tex'
+cp ../benchmarks/representation-cmp/out/results.tex representation.tex
+
+echo 'results_fx2007.tex'
+cp ../benchmarks/fx2007/out/results_fx2007.tex .
+
+echo 'results_weather.tex'
+cp ../benchmarks/weather/out/results_weather.tex .

### 2017-03-31 paper/paper.tex
minor paper updates

@@
 \begin{figure}[!ht]
-\vskip 0.2in
+\vskip 0.1in
 \begin{center}
@@
 The iteration cutoff is the number of training points $n$ in the dataset.
-} % TODO no time right now but later could compare to LCG MVMs.
+}
 \label{fx2007-iterations}
 \end{center}
-\vskip -0.2in
+\vskip -0.1in
 \end{figure}
@@
 Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense.
-% this is a good transition paragraph, and motivates what follows well, but perhaps reworking the structure of the prior sections will make this (for lack of a better word) pipeline bottleneck mroe obvious?
-TODO any suggestions?
 
@@
 We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$ (where we have computed $\Tz=W^\top\Tx$).
-
-% TODO get rid of all the linear terms below; delay big-O discussion until "asymptotics"?
 
@@
 On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm --- applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.
+In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.
 
@@
 \end{center}
-\vskip -0.2in
+\vskip -0.1in
 \end{figure}
@@
 
-Here, experiments were carried out on ``random'' parameters with the same seed.
-% im not sure exactly what that means ("random parameters")?
-same seed?
-what is an experiment?
-might call it an application to simulated data?
-if we are desperate for space, make this a single sentence...
-instead of an itemized list?
-In particular, we have the following parameters:
-\begin{itemize}
-\item $n$ total sample size across all outputs
-\item $D$ number of outputs
-\item $Q$ number of kernels $k_q$
-\item $R$ average added rank
-\item $\epsilon$ mean noise
-\item \texttt{ktype} kernel type, one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions.
+We evaluated the performance of the different kernel representations over various rank and parameterization settings.
+In particular, we have the following parameters: $n$ total sample size across all outputs, $D$ number of outputs, $Q$ number of kernels $k_q$, $R$ average added rank, $\epsilon$ mean noise, and \texttt{ktype} kernel type.
+Kernel type is one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions.
 \texttt{mix} refers to a mix of all three kernels.
-Inverse length scales and periods for the kernels varied uniformly in log space from 1 to 10 with $Q$ samples.
-\end{itemize}
 
-Given this parameterization, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
+Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
+Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
 Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
 
-As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$ --- this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
+As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$---this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
 For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
-
 \begin{table}[!ht]
@@
 We averaged over five runs.
-In every run, we use $n=5,000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
+In every run, we use $n=5000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
 }
@@
 We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance.
-The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \texttt{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below.
+The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below.
 The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP.
@@
 LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points.
-COGP used a $Q=2$ kernel with $100$ inducing points}
+COGP used a $Q=2$ kernel with $100$ inducing points.}
 \label{fx2007-tbl}
@@
 \begin{figure*}[!ht]
-\vskip 0.2in
+\vskip 0.1in
 \begin{center}
@@
 LLGP mean is a solid red curve, with light green 95\% confidence intervals.
-Magenta points are in the training set, while blue ones are in the test set.} % TODO much better -- might also have 10 words on high level point: Predictions from LLGP on the test set are uniformly better across these three tasks.
-or something.
+Magenta points are in the training set, while blue ones are in the test set.
+Notice LLGP variance corresponds to an appropriate level of uncertainty on the test set and certainty on the training set, as opposed to the uniform variance from COGP.}
 \label{fx2007-graph}
 \end{center}
-\vskip -0.2in
+\vskip -0.1in
 \end{figure*}
@@
 In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure.
-Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15,789$ training measurements and $374$ test measurements.
+Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15789$ training measurements and $374$ test measurements.
 

### 2017-03-31 paper/paper.tex
spell check

@@
 where $k_q:\R\rightarrow\R$ is a stationary kernel on $\mcX$.
-Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\bsk_q I_D$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a pre-set rank.
+Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\bsk_q I_D$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a preset rank.
 Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz.
@@
 For a gradient-only stopping heuristic, we maintain the running maximum gradient $\infty$-norm.
-If gradient norms drop below a pre-set proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate.
+If gradient norms drop below a preset proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate.
 For example, when applied to the foreign exchange rate prediction (FX2007) dataset in Section~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).
@@
 \end{document}
+
+%  LocalWords:  Feinberg pu Cheng Engelhardt Kai LLGP Bonilla COGP qA
+%  LocalWords:  hyperparameter's nguyen Semiparametric SLFM Scalable
+%  LocalWords:  MSGP vectorization matricization parameterization PSD

### 2017-04-04 benchmarks/benchlib/standard_tester.py
print latex table correctly for weather/fx2007 datasets

@@ -350,14 +350,19 @@ def latex_table(filename, cols, col_results):
     ncols = len(cols)
 
     for metric, fmt, results in zip(metrics, formatting, transposed):
-        best = np.argmin(results)
+        best = min(results, key=lambda r:r[0])
         if fmt == 'd':
             results = [tuple(map(int, r)) for r in results]
-        left = [r'\textbf{' if i == best else '' for i in range(ncols)]
-        right = ['}' if i == best else '' for i in range(ncols)]
-        center = [statprint(x, fmt, '$') for x in results]
-        row = [''.join(s) for s in zip(left, center, right)]
-        row = ' & '.join([metric] + row) + r'\\' + '\n'
+        stringified = [statprint(x, fmt, '$') for x in results]
+        for i, res in enumerate(results):
+            if res == best:
+                mean, space, var, end = stringified[i].split('$')
+                stringified[i] = r'$\textbf{' + mean + r'}$'
+                stringified[i] += space
+                stringified[i] += r'$\textbf{' + var + r'}$'
+                stringified[i] += end
+                break
+        row = ' & '.join([metric] + stringified) + r'\\' + '\n'
         latex += row
 
     # drop the last \\\n

### 2017-04-04 paper/paper.tex

@@
 
-We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2011kernels}.
+We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2012kernels}.
 

### 2017-04-04 benchmarks/benchlib/standard_tester.py
Fixed bug in latex table printing

@@ -350,18 +350,16 @@ def latex_table(filename, cols, col_results):
     ncols = len(cols)
 
     for metric, fmt, results in zip(metrics, formatting, transposed):
-        best = min(results, key=lambda r:r[0])
+        best = min(range(len(results)), key=lambda r:results[r][0])
+        print('min idx', best, results[best])
         if fmt == 'd':
             results = [tuple(map(int, r)) for r in results]
         stringified = [statprint(x, fmt, '$') for x in results]
-        for i, res in enumerate(results):
-            if res == best:
-                mean, space, var, end = stringified[i].split('$')
-                stringified[i] = r'$\textbf{' + mean + r'}$'
-                stringified[i] += space
-                stringified[i] += r'$\textbf{' + var + r'}$'
-                stringified[i] += end
-                break
+        mean, space, var, end = (x for x in stringified[best].split('$') if x)
+        stringified[best] = r'$\textbf{' + mean + r'}$'
+        stringified[best] += space
+        stringified[best] += r'$\textbf{' + var + r'}$'
+        stringified[best] += end
         row = ' & '.join([metric] + stringified) + r'\\' + '\n'
         latex += row
 

### 2017-04-04 benchmarks/representation-cmp/run.sh
Fixed bug in representation-cmp latex table printing

@@ -107,8 +107,14 @@ for i in 3 10 17; do
     # we only want d, r, q
     numbersonly=$(echo $numbersonly | cut -d' ' -f2-)    
     
+    avgs=""
+    for i in $(cat /tmp/avgs); do
+        i=$(printf "%0.2f" $i)
+        avgs="${avgs} $i"
+    done
+    
     row=""
-    for number in $numbersonly $(cat /tmp/avgs); do
+    for number in $numbersonly $avgs ; do
         if [ "$number" = "$minnum" ]; then
             left="\\\\textbf{"
             right="}"
@@ -116,7 +122,6 @@ for i in 3 10 17; do
             left=""
             right=""
         fi
-        number=$(printf "%0.2f" $number)
         row="${row} & \$ ${left}${number}${right} \$"
     done
     row=$(echo $row | cut -c3-)

### 2017-04-06 benchmarks/representation-cmp/run.sh
Fixed representation-cmp latex table printing and reran

@@ -108,13 +108,8 @@ for i in 3 10 17; do
     numbersonly=$(echo $numbersonly | cut -d' ' -f2-)    
     
     avgs=""
-    for i in $(cat /tmp/avgs); do
-        i=$(printf "%0.2f" $i)
-        avgs="${avgs} $i"
-    done
-    
-    row=""
-    for number in $numbersonly $avgs ; do
+    for number in $(cat /tmp/avgs); do
+        i=$(printf "%0.2f" $number)
         if [ "$number" = "$minnum" ]; then
             left="\\\\textbf{"
             right="}"
@@ -122,7 +117,12 @@ for i in 3 10 17; do
             left=""
             right=""
         fi
-        row="${row} & \$ ${left}${number}${right} \$"
+        avgs="${avgs} ${left}${i}${right}"
+    done
+    
+    row=""
+    for number in $numbersonly $avgs ; do
+        row="${row} & \$ $number \$"
     done
     row=$(echo $row | cut -c3-)
     newline=$'\n'

### 2017-04-11 paper/paper.tex
updated paper with re-run benchmarks
Minor paper touch-up / removed TODOs

@@
 
-\section{Matrix-free LMC Learning}
-\label{sec:matrix-free}
+\section{Matrix-free LMC Learning} \label{sec:matrix-free}
 
@@
 
-Although this is a method of optimizing the marginal log likelihood, it can easily be augmented with a hyperprior $\log p(\bsth)$ term to Eq.~\ref{datall}.% what does "augmented with a hyperprior $\log p(\bsth)$ term to Eq.~\ref{datall}" mean?
-make this explicit here.
-a log probabliity is not a hyper prior (generally)
+Although Eq.~\ref{datall} does not have a hyperprior $\log p(\bsth)$ term, it can easily be augmented with one.
 
@@
 
-\section{Results} % TODO read these over
+\section{Results}
 \label{sec:results}
@@
 
-The data used in this example are from 2007, and include $n=3,054$ training points and $150$ test points.
+The data used in this example are from 2007, and include $n=3054$ training points and $150$ test points.
 The test points include 50 contiguous points extracted from each of the CAD, JPY, and AUD exchanges.
@@
 \bibliography{paper}
-\bibliographystyle{icml2017} % TODO - follow blei notes for this
+\bibliographystyle{icml2017}
 

### 2017-05-07 paper/paper.tex
icml -> nips paper conversion

@@
 \documentclass{article}
-\usepackage{times}
+
+\PassOptionsToPackage{numbers, compress}{natbib}
+% Camera ready: \usepackage[final]{nips_2017}
+\usepackage{nips_2017}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
 \usepackage{graphicx}
+\usepackage{algorithm}% http://ctan.org/pkg/algorithms
+\usepackage{algpseudocode}% http://ctan.org/pkg/algorithmicx
+\usepackage{booktabs}
 \usepackage{subfigure}
-\usepackage{natbib}
-\usepackage{algorithm}
-\usepackage[nice]{nicefrac}
-\usepackage{algorithmic}
-\usepackage{hyperref}
-\newcommand{\theHalgorithm}{\arabic{algorithm}}
-\usepackage{icml2017}
-%\usepackage[accepted]{icml2017}
 
-\def\bsa{{\boldsymbol\alpha}}
-\def\bse{{\boldsymbol\epsilon}}
-\def\bsk{{\boldsymbol\kappa}}
-\def\bsm{{\boldsymbol\mu}}
-\def\bsl{{\boldsymbol\lambda}}
-\def\bsph{{\boldsymbol\phi}}
-\def\bsth{{\boldsymbol\theta}}
-\usepackage{amsmath}
-\usepackage{amssymb}
-\usepackage{amsfonts}
-\usepackage{enumerate}
-\usepackage{fullpage}
-\usepackage{bbm}
-\usepackage{caption}
-\usepackage{mathrsfs}
+\input{defs}
 
-\newcommand{\pa}[1]{ \left({#1}\right) }
+\title{Large Linear Multi-output Gaussian Process Learning for Time Series}
 
-\def \N {\mathbb{N}}
-\def \Nbr {\mathcal{N}}
-\def \mcL {\mathcal{L}}
-\def \Q {\mathbb{Q}}
-\def \F {\mathbb{F}}
-\def \then {\implies &}
-\def \oif {\Longleftrightarrow &\,}
-\def \given {\text{Given }&}
-\def \assume {\text{Assume }&}
-\def \thfr {\therefore &\enskip}
-\def \bij {\leftrightarrow}
-\def \inj {\rightarrowtail}
-\def \sur {\twoheadedrightarrow}
-\def \Z {\mathbb{Z}}
-\def \R {\mathbb{R}}
-\def \C {\mathbb{C}}
-\def \T {\mathbb{T}}
-\def \iff {\Longleftrightarrow}
-\def \kron {\boldsymbol\delta}
-\def \indicator {\mathbbm{1}}
-
-\def\Tx{\textbf{x}}
-\def\TX{\textbf{X}}
-\def\TU{\textbf{U}}
-\def\Ty{\textbf{y}}
-\def\tbr{\textbf{r}}
-\def\Tz{\textbf{z}}
-\def\quotient{\mathclose{}/\mathopen{}}
-\def\Tf{\textbf{f}}
-\def\Th{\textbf{h}}
-\def\Tg{\textbf{g}}
-\def\sumn{\sum_{n=0}^\infty}
-\def\limn{\lim_{n\rightarrow\infty}}
-\def\prodn{\prod_{n=0}^\infty}
-
-\DeclareMathOperator\adj{adj}
-\DeclareMathOperator\cov{cov}
-\DeclareMathOperator\Unif{Unif}
-\DeclareMathOperator\blockdiag{blockdiag}
-\DeclareMathOperator\Pois{Pois}
-\DeclareMathOperator\mathProb{\mathbb{P}}
-\def\P{\mathProb} % need to overwrite stupid paragraph symbol
-\DeclareMathOperator\mathExp{\mathbb{E}}
-\def \E {\mathExp}
-
-\newcommand{\stc}[1]{\widetilde{#1}}   
-\newcommand{\set}[2]{ \left\{ #1 \,\middle|\, #2 \right\} }
-\newcommand{\shift}[1]{&\quad & \text{#1}\\}
-\newcommand{\lem}[1]{\text{\textbf{L.\ref{#1}}}}
-\newcommand{\card}[1]{\left\vert{#1}\right\vert}
-\newcommand{\Ps}[1]{\mathcal{P}\left({ #1 }\right)}
-\newcommand{\colv}[1]{\begin{pmatrix} #1 \end{pmatrix}}
-\newcommand{\mat}[1]{\begin{pmatrix} #1 \end{pmatrix}}
-\newcommand{\detmat}[1]{\begin{vmatrix} #1 \end{vmatrix}}
-\newcommand{\spanb}[1]{\text{span}\{ #1 \}}
-\newcommand{\abs}[1]{\left|#1\right|}
-\newcommand{\Inner}[1]{\langle #1 \rangle}
-\newcommand{\Innercpy}[1]{\langle #1, #1 \rangle}
-\newcommand{\norm}[1]{\mathopen\| #1 \mathclose\|}% use instead of $\|x\|$
-
-\def\eqd{\mathrel{\overset{\Delta}{=}}}
-
-\DeclareMathOperator{\diag}{diag}
-\DeclareMathOperator{\vcdim}{VC-dim}
-\DeclareMathOperator*{\Err}{\text{err}}
-\DeclareMathOperator*{\ErrE}{\mathbb{E}}
-\DeclareMathOperator{\Tr}{tr}
-\DeclareMathOperator{\Dim}{dim}
-\DeclareMathOperator{\Rank}{rank}
-\DeclareMathOperator*{\argmin}{argmin}
-\DeclareMathOperator*{\proj}{proj}
-\DeclareMathOperator{\Ker}{ker}
-\DeclareMathOperator{\Diam}{diam}
-\DeclareMathOperator{\Int}{int}
-\DeclareMathOperator{\Clo}{clo}
-\DeclareMathOperator{\sgn}{sgn}
-\DeclareMathOperator{\MyRe}{Re}
-\DeclareMathOperator{\MyIm}{Im}
-\DeclareMathOperator{\Supp}{supp}
-\DeclareMathOperator{\Var}{Var}
-
-% probability stuff
-\def \sa {{$\sigma$-algebra}}
-\def\OR{{\overline{\R}}}
-\def\OX{{\overline{X}}}
-
-\def \mcX {\mathcal{X}}
-\def \mcY {\mathcal{Y}}
-\def \mcH {\mathcal{H}}
-\def \mcD {\mathcal{D}}
-\def \mcA {\mathcal{A}}
-\def \mcK {\mathcal{K}}
-\def\mcG{{\mathcal{G}}}
-\def\mcH{{\mathcal{H}}}
-\def\mcF{{\mathcal{F}}}
-\def\mcB{{\mathcal{B}}}
-\def\mcE{{\mathcal{E}}}
-\def\mcP{{\mathcal{P}}}
-\def\mcM{{\mathcal{M}}}
-\def\mcC{{\mathcal{C}}}
-\def\mcT{{\mathcal{T}}}
-
-% \icmltitlerunning{Submission and Formatting Instructions for ICML 2017}
+\author{
+  Vladimir Feinberg, Li-Fang Cheng, {Barbara Engelhardt}, {Kai Li}\\
+  Department of Computer Science\\
+  Princeton University\\
+  Princeton, NJ 08544 \\
+  \texttt{\{vyf,lifangc,bee,li\}@princeton.edu} \\
+}
 
@@
 
-\twocolumn[
-\icmltitle{Large Linear Multi-output Gaussian Process Learning}
-%\icmlsetsymbol{equal}{*}
-\begin{icmlauthorlist}
-  \icmlauthor{Vladimir Feinberg}{pu}
-  \icmlauthor{Li-Fang Cheng}{pu}
-  \icmlauthor{Barbara E. Engelhardt}{pu}
-  \icmlauthor{Kai Li}{pu}
-\end{icmlauthorlist}
-\icmlaffiliation{pu}{Princeton University, Princeton, USA}
-\icmlcorrespondingauthor{Vladimir Feinberg}{vyf@princeton.edu}
-
-\icmlkeywords{multi-task, multi-output, Gaussian process, large scale}
-
-\vskip 0.3in
-]
-\printAffiliationsAndNotice{}
-%\printAffiliationsAndNotice{\icmlEqualContribution}
+\maketitle
 
@@
 Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}.
-Such models can learn correlations across the multiple outputs, such as temperature data from disparate regions, which can then be exploited to share knowledge across the multiple outputs.
-While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large $n$, the total number of observations across all outputs.
-In this work, we describe a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs.
-We show results of our approach applied to real time series data, and compare these results against related methods.
+Such models can learn correlations across the multiple outputs, which can then be exploited to share knowledge across the multiple outputs.
+For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than the same model applied to a single region.
+While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large numbers of observations across all outputs.
+In this work, we propose a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs.
+Our approach learns GP kernel hyperparameters at an asymptotically faster rate than the current state of the art.
+When applied to real time series data, we find this theoretical improvement is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
 Finally, we discuss extensions of our approach to multidimensional inputs.
@@
 
-\section{Introduction}
-\label{introduction}
+% NOTES
+% 8 pages of content
+% \citet for inline citations
+% footnotes after punctuation
+% no vertical rules in tables
+% \subsubsection*{Acknowledgments} only in final paper
+
+\section{Introduction}\label{introduction}
 
@@
 
-Such an extension is especially important in time series data, where multi-output GPs have been applied to the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
+Multiple-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
+Efficient model learning would enable researchers to quickly explore large spaces of model configurations to find an appropriate one for their task.
 
-Exact GP inference comes at a cost: to compute the gradients for kernel hyperparameters, an exact computation requires an $O(n^3)$ decomposition of an $O(n^2)$ matrix in memory for $n$ samples \cite{gpml}.
-Each hyperparameter's scalar partial derivative then requires another $O(n^2)$ computation.
-While these operations may be performed in parallel, the memory use and runtime is prohibitive for problems with a large number of outputs, since the number of hyperparameters grows accordingly.
+Exact GP inference is infeasible in this scenario since it requires maintaining pairwise covariance between all inputs in a large matrix and performing inversions with that matrix \cite{williams1996gaussian}.
+As the number of hyperparameters grows, so do the number of matrix operations (Tab.~\ref{asymp}).
 For this reason, an accurate approximate approach has been an important goal of machine learning research.
-
-For a single output, variational approaches that work on $m\ll n$ inducing points, or locations at which the true kernel is sampled, letting users trade off accuracy for efficiency \cite{hensman2013gaussian}.
-
-Nguyen and Bonilla observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once \yrcite{nguyen2014collaborative}.
-We analogously exploit sharing in the covariance structure.
 
@@
 
-Our approach makes the following contributions to approximate inference in multi-output GPs.
+Our approach makes the following contributions to approximate inference in multi-output GPs, scoped to time series input.
 First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid.
@@
 
-When inputs do not lie on a uniform grid (i.e., for time series, the length of time between observations differs), we apply a technique proposed in Massively Scalable Gaussian Processes (MSGP) \cite{msgp}.
-For single-output kernels, MSGP leverages the structured kernel interpolation (SKI) framework to make inputs act as if they are on a grid \cite{kiss-gp}.
-In a single-output case, this opens up the possibility for a matrix-free GP, avoiding construction of the explicit Gram covariance matrix.
+When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}.
 Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.
-While direct application of SKI is not possible, we can make an extension that makes matrix-free LMC kernel learning viable even on non-grid inputs.
+
+For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
 
 Our paper is organized as follows.
-In Section~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs.
-Section~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs.
-Section~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion.
-Then, in Section~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Section~\ref{conclusion}.
+In Sec.~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs.
+Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs.
+Sec.~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion.
+Then, in Sec.~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Sec.~\ref{conclusion}.
 
@@
 \label{sec:background}
+
 \subsection{General Gaussian processes}
@@
 
-Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by marginalization \cite{gpml}:
-\begin{equation*}
-y_*|\Ty\sim N\pa{K_{*, {X}}K^{-1}_{{X}, {X}}\Ty, K_{*, *}-K_{*, {X}}K^{-1}_{{X}, {X}}K_{{X}, *}}.
-\end{equation*}
-Accuracy is often sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
+Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty$ \cite{williams1996gaussian}.
+Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
 \begin{align}
-  \mcL(\bsth)&=\log p(\Tf_{X}|X,\bsth)\label{datall}\\
-             &=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.
+  \mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.\label{datall}
 \end{align}
 As such, the heart of GP learning lies in optimization of $\mcL$.
-Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
+Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
 \begin{align}
@@
 
+Thus, the LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs.
+The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
+By choosing $B_q$ to have rank $R_q=D$, so the contribution between those two outputs is learned independently of all other contributions, the entries of $B_q$.
+By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction.
+
 \subsection{Structured covariance matrices}
@@
 We can represent it by keeping representations of $A$ and $B$ separately, rather than the product.
-Then, the corresponding MVMs can be computed in time $O(a\text{MVM}(B)+b\text{MVM}(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
+Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
 For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.
@@
 \]
-Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs  $C\mat{\Tx\\\textbf{0}}=\mat{T\Tx\\\textbf{0}}$ in $O(n\log n)$ time.
+Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs $C\mat{\Tx & \textbf{0}}^\top=\mat{T\Tx &\textbf{0}}^\top$ in $O(n\log n)$ time.
 This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.
 
-\section{Related Work}
+\section{Related work}
 \label{sec:related-work}
@@
 Setting the inducing points $\TU=\TX$ recovers the original exact GP.
+Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
 
-In the context of LMC, a framework for variational DTC, DTCVAR, provided a matrix-free method where each iteration has complexity $O(n(Dm)^2)$, where $m$ is the average number of inducing points per output, $\card{\TU}/D$ \cite{alvarez2010efficient}.
-
-By sharing the $m$ inducing points across all outputs, Collaborative Multi-output Gaussian Processes (COGP) improves runtime per iteration to $O(m^3)$.
-Moreover, COGP only requires $O(Dnm)$ memory \cite{nguyen2014collaborative}.
-To ensure that $O(m^3)$ is computationally tractable for COGP, $m$ must be sufficiently small.
-COGP makes large multi-output GP estimation feasible through a variational approximation, the evidence lower bound, to the log likelihood.
+\citet{nguyen2014collaborative} observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once.
+COGP further uses a variational approximation, the evidence lower bound, to the log likelihood.
+COGP supports a subset of LMC kernels, namely those that match the SLFM model \cite{seeger2005semiparametric}.
+In an LMC representation (Eq.~\ref{lmcpointwise}), these models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
+Moreover, SLFM and COGP models add in an independent GP to each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.
 
@@
 
-\section{Matrix-free LMC Learning} \label{sec:matrix-free}
+\section{Matrix-free LMC learning} \label{sec:matrix-free}
 
 We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches.
-Critically, the accuracy of the method need not be reduced by keeping $m$ low because its reliance on structure allows better asymptotic performance.
+Critically, the accuracy of the method need not be reduced by keeping the number of interpolation points $m$ low because its reliance on structure allows better asymptotic performance.
 For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.
 
-Although Eq.~\ref{datall} does not have a hyperprior $\log p(\bsth)$ term, it can easily be augmented with one.
+\subsection{Matrix-free GP learning}
 
-\subsection{Matrix-free GP Learning}
+For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$.
+Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
+Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
+We considered several stochastic approximations for finding the pathological term $\log\abs{K_{|\bsth}}$ in Eq.~\ref{datall} \cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.
 
-Gibbs and MacKay describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix \yrcite{gibbs1996cient}.
+\subsection{Gradient construction}
+
+\citet{gibbs1996cient} describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix.
 In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG).
@@
 For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
-As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$.
+As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number of $N$ samples from $\tbr$.
 
-We depart from Gibbs and MacKay in selecting MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
+We depart from Gibbs and MacKay in two ways, yielding Algorithm~\ref{alg:grad}.
+First, we do not construct $K_{|\bsth}$, but a low-memory representation $\tilde{K}_{|\bsth}$, described in Sec.~\ref{fast-mvm}.
+Second, we select MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
 MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}.
-This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations (Fig.~\ref{fx2007-iterations}).
+This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations because of increases in $\kappa_2$, the spectral condition number of $K$ (Fig.~\ref{fx2007-iterations}).
 
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{iterations.eps}}
-\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Section~\ref{fx2007-results}.
+\begin{figure}[!h]
+\centering
+\includegraphics[width=0.5\columnwidth]{iterations.eps}
+\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
 The iteration cutoff is the number of training points $n$ in the dataset.
@@
 \label{fx2007-iterations}
-\end{center}
-\vskip -0.1in
 \end{figure}
 
-If we sample $N_t$ instances of $\tbr$, every iteration needs to perform $N_t+1$ inversions including $\Ty$, if we reuse the same samples $K_{|\bsth}^{-1}\tbr$ between derivatives $\partial_{\theta_j}$.
-These inversions are done in parallel.
-Then, since $\partial_{\theta_j}K_{|\bsth}$ is both structured and smaller than $K_{|\bsth}$, each of the terms in the approximation Eq.~\ref{eq:trace} require $\tilde{O}(n)$ time to compute.
+Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K}_{|\bsth})\sqrt{\kappa_2})$ \cite{raykar2007fast}.
+This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K_{|\bsth}$ for each $j$ at least as fast as $\mvm({\tilde{K}_{|\bsth}})$.
+Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K_{|\bsth}$.
 
-Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine \cite{zeiler2012adadelta}.
-We considered several stochastic approximations for finding $\log\abs{K_{|\bsth}}$~\cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.
-
-In other words, all the work is done in the inversions $K_{|\bsth}^{-1}\tbr,K_{|\bsth}^{-1}\Ty$, which in turn rely on the MVM operation, discussed in Section~\ref{fast-mvm}.
-Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation, which need not be dense.
+\begin{algorithm}[!ht]
+  \caption{Compute an approximation of $\nabla L$.
+Assume \textsc{minres} is the inversion routine.
+We also assume we have access to linear operators $D_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}_{|\bsth}$.} \label{alg:grad}
+\begin{algorithmic}[1]
+  \Procedure{LLGP}{$\tilde{K}_{|\bsth}$, $\Ty$, $N$, $\{D_{\theta_j}\}$}
+  \State $R\gets\idx{\tbr_i}{i=1}{N}$, sampling $\tbr\sim\Unif\{\pm 1\}$.
+\For{$\Tz$ in $\{\Ty\}\cup R$, in parallel}
+\State $K^{-1}\Tz\gets\textsc{minres}(\tilde{K}_{|\bsth},\Tz)$.
+\EndFor
+\State $g\gets \bsz$
+\For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
+\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot D_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
+\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}_{|\bsth}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
+\EndFor
+\State \Return $g$ 
+\EndProcedure
+\end{algorithmic}
+\end{algorithm}
 
@@
 
-When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$.
+The bottleneck of Algorithm~\ref{alg:grad} is the iterative MVM operations in $\textsc{minres}$.
+Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense, so long as it can reconstruct multiplication with any vector to arbitrary, fixed precision.
+
+When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$, the shared interpolating points for all outputs:
 \begin{align}
-  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}
+  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}.
 \end{align}
-The SLFM and COGP models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
-Moreover, we include $D$ additional kernels for the independent GP components, which can be incorporated using kernels $K_d$ where $A_d=0$ and $\bsk_d$ as the $d$th standard basis vector of $\R^D$ for $d\in[D]$.
-This shows a reduction from SLFM to LMC.
 
 Importantly, the Kronecker structure of Eq.~\ref{lmcgrid} lets us re-use the same grid $K_q$ in a computational sense across the different outputs.
-Recalling our SKI extension to multiple outputs (Section~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
+Recalling our SKI extension to multiple outputs (Sec.~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
 \begin{align}
@@
 But this fact does not prevent efficient representation or multiplication since the matrix is diagonal.
-In particular, the MVM operation $K_{\TX,\TX}\Tx$ can be approximated by $WK_{\TU,\TU}W^\top\Tx+\bse\Tx$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.
+Then, the MVM operation $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.
 
-We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$ (where we have computed $\Tz=W^\top\Tx$).
+We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$.
 
@@
 In \textsc{sum}, we represent $K_{\TU,\TU}$ with a $Q$-length list.
-At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, where only its top row needs to be represented to perform MVMs.
-
-In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results, corresponding to .
+At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, with only the top row maintained.
+In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
 As described before, the Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
-In turn, the runtime for each of the $Q$ terms is $O(Dm\log m)$.
+In turn, assuming $D\ll m$, the runtime for each of the $Q$ terms is $O(Dm\log m)$.
 
@@
 
+On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$.
+In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence of Krylov-based inversion routines using fewer MVMs \cite{chan1994circulant}.
+
 \subsubsection{\textsc{slfm}: SLFM representation}
 
-\textsc{slfm} uses a rank-based representation. Let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average added rank, $R\le D$.
-
-We first rewrite the grid kernel:
+For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average added rank, $R\le D$, and re-write the kernel:
 \begin{align*}
-  K_{U,U}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
+  K_{\TU,\TU}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
 \end{align*}
@@
 \end{align*}
-where the second simplification has $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}.
+where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}.
 Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-
 Thus, the \textsc{slfm} representation writes $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix, so MVMs take $O((QR + D)m\log m)$ time.
 
-\subsection{Asymptotic performance}
-
-Because the runtimes of \textsc{bt} and \textsc{slfm} are complimentary in the sense that one performs better than the other when $D^2>QR$ and vice-versa, an algorithm that uses the aforementioned condition between to decide between which representation to use can minimize runtime.
+Because the runtimes of \textsc{bt} and \textsc{slfm} are complimentary in the sense that one performs better than the other when $D^2>QR$ and vice-versa, an algorithm that uses the aforementioned condition between to decide between which representation to use can minimize runtime (Tab.~\ref{asymp}).
 We also found that \textsc{sum} is efficient in practice for $Q=1$.
 
-Using the switching condition, MVMs with the original matrix $K_{\TX,\TX}$ altogether have a space and time upper bound of $\tilde{O}(\min(QR,D^2)m+n)$, where the $\min$ is earned thanks to the choice between different representations.
-Every AdaDelta iteration then takes $\tilde{O}(\sqrt{\kappa_2})$ such matrix-vector products for machine-precision answers, with $\kappa_2$ the spectral condition number \cite{raykar2007fast}.
-
-On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence using fewer MVMs \cite{chan1994circulant}.
+\begin{table}[!h]
+  \caption{
+   For both LLGP and COGP, $m$ is a configurable parameter which increases up to $n$ to improve accuracy.
+$Q,R,D,\kappa_2$ are coefficients dependent on the setting of the LMC kernel, which has about $QRD$ hyperparameters (Eq.~\ref{lmcpointwise}).
+The resulting asymptotic performance is given in the table.
+COGP is only independent of $R$ because it cannot represent models for $R\neq 1$.
+  }
+\label{asymp}
+\begin{sc}
+\begin{center}
+\begin{small}
+\begin{tabular}{ccc}
+  \toprule
+  Method & Up-front cost for $\nabla \mcL$ & \begin{tabular}{c}Additional cost for $\partial_{\theta_j}\mcL$\\ per hyperparameter\end{tabular}\\
+  \midrule
+  Exact & $n^3$ & $n^2 $\\
+  COGP & $Qm^3$ & $nm$ \\
+  LLGP & $\min(QR, D^2)\sqrt{\kappa_2} (n + m\log m)$ &  $n + m\log m$ \\
+  \bottomrule
+\end{tabular}
+\end{small}
+\end{center}
+\end{sc}
+\end{table}
 
@@
 If gradient norms drop below a preset proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate.
-For example, when applied to the foreign exchange rate prediction (FX2007) dataset in Section~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).
+For example, when applied to the foreign exchange rate prediction dataset in Sec.~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).
 
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{running_cutoff.eps}}
+\begin{figure}[!h]
+\centering
+{\includegraphics[width=0.5\columnwidth]{running_cutoff.eps}}
 \caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
 \label{fx2007-stop}
-\end{center}
-\vskip -0.1in
 \end{figure}
@@
 
-\subsection{Hyperparameter Settings}
+\subsection{Hyperparameter settings}
 
@@
 \texttt{mix} refers to a mix of all three kernels.
-
 Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
@@
 For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
-\begin{table}[!ht]
-  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Section~\ref{empirical-rep} using MINRES for each of the kernel representations.
+\begin{table}[!h]
+  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using MINRES for each of the kernel representations.
 For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system.
@@
 \label{complement}
-\vskip 0.15in
 \begin{center}
@@
 \end{center}
-\vskip -0.1in
 \end{table}
 
-We next evaluated the accuracy of the gradients for $N_t=10$ trace samples.
+We next evaluated the accuracy of the gradients for $N=10$ trace samples.
 Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$.
-Since, for each partial derivative, LLGP requires only $N_t$ $n$-sized vector dot products (Eq.~\ref{eq:trace}), it generally runs faster than the exact approach (Fig.~\ref{fig:gradtime}), which must compute a matrix-matrix dot product (Eq.~\ref{llgradient}).
-The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}).
-Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor MINRES convergence in the inversions (Fig.~\ref{fig:inverr}).
+Since, for each partial derivative, LLGP requires only $N$ $n$-sized vector dot products (Eq.~\ref{eq:trace}), it generally runs faster than the exact approach (Fig.~\ref{fig:rellog}), which must compute a matrix-matrix dot product (Eq.~\ref{llgradient}).
+The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:rellog}).
+Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor MINRES convergence in the inversions.
 We will show that the stochastic gradients suffice for optimization in practical examples.
 
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{relgrad_l2.eps}}
-\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and LLGP gradients.
-Higher is better, and extremal values are noted.
+\begin{figure}[!h]
+  \centering
+
+ \subfigure[time]{
+    \resizebox{0.48\textwidth}{!}{\includegraphics{time_ratio.eps}}}
+  \subfigure[gradient accuracy]{
+    \resizebox{0.48\textwidth}{!}{\includegraphics{relgrad_l2.eps}}}
+%  \subfigure[inversion accuracy]{
+%    \resizebox{0.40\textwidth}{!}{\includegraphics{relalpha_l2.eps}}}
+ \caption{Negative logarithm of the ratio of the exact algorithm to LLGP's analogue for (a) time and (b) accuracy metrics for evaluating gradient computation performance.
+In each case, higher is better, and extremal values are noted.
 For each data point, the average was taken over five runs.}
-\label{fig:relgrad}
-\end{center}
-\vskip -0.1in
-\end{figure}
-
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{time_ratio.eps}}
-\caption{Logarithm of the ratio of time required by the exact approach to that of LLGP for computing the gradient over all parameters from Eq.~\ref{llgradient}.
-Higher is better, and extremal values are noted.
-For each data point, the average was taken over five runs.}
-\label{fig:gradtime}
-\end{center}
-\vskip -0.1in
-\end{figure}
-
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{relalpha_l2.eps}}
-\caption{Negative logarithm of the relative error in $\ell_2$ norm between exact and MINRES solutions to $K\Tx=\Ty$ for $\Tx$. For each data point, the average was taken over five runs.}
-\label{fig:inverr}
-\end{center}
-\vskip -0.1in
+\label{fig:rellog}
 \end{figure}
@@
 
-\begin{table}[!ht]
+\begin{table}[!h]
   \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset.
@@
 \label{fx2007-tbl}
-\vskip 0.15in
 \begin{center}
@@
 \end{center}
-\vskip -0.1in
 \end{table}
 
-\begin{figure*}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
+\begin{figure}[!h]
+\centering
+{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
 \caption{Test outputs for the FX2007 dataset.
@@
 \label{fx2007-graph}
-\end{center}
-\vskip -0.1in
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{table}[!ht]
+\begin{table}[!h]
   \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the weather dataset.
@@
 \label{weather-tbl}
-\vskip 0.15in
 \begin{center}
@@
 \end{center}
-\vskip -0.1in
 \end{table}
@@
 
-\begin{figure}[!ht]
-\vskip 0.1in
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{m_time_nlpd.eps}}
+\begin{figure}[!h]
+\centering
+{\includegraphics[width=0.5\columnwidth]{m_time_nlpd.eps}}
 \caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
 \label{fig:llgpweather}
-\end{center}
-\vskip -0.1in
 \end{figure}
@@
 
+% \begin{table}[t]
+%   \caption{Sample table title}
+%   \label{sample-table}
+%   \centering
+%   \begin{tabular}{lll}
+%     \toprule
+%     \multicolumn{2}{c}{Part}                   \\
+%     \cmidrule{1-2}
+%     Name     & Description     & Size ($\mu$m) \\
+%     \midrule
+%     Dendrite & Input terminal  & $\sim$100     \\
+%     Axon     & Output terminal & $\sim$10      \\
+%     Soma     & Cell body       & up to $106$  \\
+%     \bottomrule
+%   \end{tabular}
+% \end{table}
+
+\bibliographystyle{unsrtnat}
 \bibliography{paper}
-\bibliographystyle{icml2017}
 
 \end{document}
-
-%  LocalWords:  Feinberg pu Cheng Engelhardt Kai LLGP Bonilla COGP qA
-%  LocalWords:  hyperparameter's nguyen Semiparametric SLFM Scalable
-%  LocalWords:  MSGP vectorization matricization parameterization PSD

### 2017-05-08 paper/paper.tex
nips paper future work corrections, rm comment

@@
 
-LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tbl.~\ref{weather-tbl}).
+LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tab.~\ref{weather-tbl}).
 Varying the number of interpolating points $m$ from $500$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}).
@@
 We have shown on several datasets that this can be done in a way that is faster and leads to more accurate results than variational approximations.
+Because LLGP's representation can be efficient without being sparse (i.e., $m$ may be large), it can can capture complex interactions in the covariance.
+As pictured in Fig.~\ref{fx2007-graph}, and demonstrated by NLPD performance on both the FX2007 and weather datasets (Tab.~\ref{fx2007-tbl},\ref{weather-tbl}), such an efficient, non-sparse representation is integral to making a model whose confidence at a test site reflects the amount of related training data for that site.
 
@@
 
-% \begin{table}[t]
-%   \caption{Sample table title}
-%   \label{sample-table}
-%   \centering
-%   \begin{tabular}{lll}
-%     \toprule
-%     \multicolumn{2}{c}{Part}                   \\
-%     \cmidrule{1-2}
-%     Name     & Description     & Size ($\mu$m) \\
-%     \midrule
-%     Dendrite & Input terminal  & $\sim$100     \\
-%     Axon     & Output terminal & $\sim$10      \\
-%     Soma     & Cell body       & up to $106$  \\
-%     \bottomrule
-%   \end{tabular}
-% \end{table}
-
 \bibliographystyle{unsrtnat}

### 2017-05-13 benchmarks/benchlib/bench.py
Truncated normal coreg init

@@ -10,6 +10,7 @@ import numpy as np
 import scipy.linalg as la
 import scipy.spatial.distance
 import scipy.sparse.linalg as sla
+import scipy.stats
 
 from runlmc.approx.interpolation import multi_interpolant
 from runlmc.approx.iterative import Iterative
@@ -102,7 +103,8 @@ def _main():
     print('n_o {} d {} r_q {} q {} eps {} kern {} seed {} test-type {}'.format(
         n_o, d, r_q, q, eps, kern, seed, testtype))
 
-    coreg_vecs = np.random.randn(q, r_q, d)
+    distrib = scipy.stats.truncnorm(-1, 1)
+    coreg_vecs = distrib.rvs(size=(q, r_q, d))
     coreg_diags = np.reciprocal(np.random.gamma(shape=1, scale=1, size=(q, d)))
     noise = np.reciprocal(np.random.gamma(
         shape=(1 + (1 / eps)), scale=1, size=d))

### 2017-05-13 paper/paper.tex
Added paper corrections from Li-Fang

@@
 \begin{abstract}
-Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}.
-Such models can learn correlations across the multiple outputs, which can then be exploited to share knowledge across the multiple outputs.
-For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than the same model applied to a single region.
-While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large numbers of observations across all outputs.
-In this work, we propose a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs.
-Our approach learns GP kernel hyperparameters at an asymptotically faster rate than the current state of the art.
+Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization (LMC) \cite{alvarez2012kernels}.
+Such models can learn and exploit correlations across the multiple outputs.
+For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than a single, localized model.
+While model learning can be performed efficiently for single-output GPs \cite{msgp}, such methods assume stationarity, a luxury unavailable in the multi-output case.
+We propose Large Linear GP (LLGP), which recovers stationarity thanks to LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and one-dimensional inputs.
+Our approach learns hyperparameters at an asymptotically faster rate than the current state of the art.
 When applied to real time series data, we find this theoretical improvement is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
@@
 Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}.
-Both of these structures coordinate for fast matrix-vector multiplication $K\Ty$ with the covariance matrix $K$.
+Both of these structures coordinate for fast matrix-vector multiplication $K\Tz$ with the covariance matrix $K$ for any vector $\Tz$.
 
@@
 First, we give an overview of the theoretical foundation for GPs.
-A GP is a set of random variables (RVs) $\{f_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that for any finite collection $X\subset\mcX$, the RVs are jointly Gaussian with a prescribed mean function $\bsm:\mcX\rightarrow\R$ and covariance $K:\mcX^2\rightarrow\R$, i.e., $f_X\sim N\pa{\bsm_X, K_{X, X}}$ \cite{williams1996gaussian}.
-As usual, we drop the mean with $\Ty\triangleq f_X-\bsm_X$ and denote vectorization $f_X=\colv{f_{\Tx}}_{\Tx\in X}$ or matricization $K_{X,Z}=\mat{K(\Tx,\Tz)}_{\Tx\in X,\Tz\in Z}$ with subscripts.
-
-Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty$ \cite{williams1996gaussian}.
-Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
+A GP is a set of random variables (RVs) $\{f_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}.
+With zero mean wlog and a prescribed covariance $K:\mcX^2\rightarrow\R$, $\Tf_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Tf_X)_i=f_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
+Given observations $\Ty\in\R^n$ of $\Tf_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Tf_X$ \cite{williams1996gaussian}.
+Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)$.
+As such, the heart of GP learning lies in optimization of $\mcL$.
+Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
 \begin{align}
-  \mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.\label{datall}
-\end{align}
-As such, the heart of GP learning lies in optimization of $\mcL$.
-Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
-\begin{align}
-\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}}.\label{llgradient}
+\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}};\;\; \bsa=K_{|\bsth}^{-1}\Ty.\label{llgradient}
 \end{align}
@@
 We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2012kernels}.
-
 Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$: an input is in a point from a shared domain coupled with an output tag.
@@
 \begin{align}
-K([i,\Tx_i],[j,\Tx_j])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx_i-\Tx_j})+\epsilon_i\delta_{ij},\label{lmcpointwise}
+K([i,\Tx_i],[j,\Tx_j])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx_i-\Tx_j})+\epsilon_i1_{i=j},\label{lmcpointwise}
 \end{align} 
@@
 
-Thus, the LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs.
+The LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs.
 The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
 By choosing $B_q$ to have rank $R_q=D$, so the contribution between those two outputs is learned independently of all other contributions, the entries of $B_q$.
-By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction.
+By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$ then we have an independent GP for each output).
 
@@
 The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$.
-Importantly, $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
+$W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
 
@@
 Massively Scalable Gaussian Processes (MSGP) observes that the kernel $K_{U,U}$ on a grid exhibits Kronecker and Toeplitz matrix structure \cite{msgp}.
-Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{datall}.
-In addition, \cite{wilson2014fast} mentions an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{datall}.
+Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
+In addition, \cite{wilson2014fast} mentions an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{llgradient}.
 
@@
 
-\section{Matrix-free LMC learning} \label{sec:matrix-free}
+\section{Large Linear GP} \label{sec:matrix-free}
 
@@
 
-\subsection{Matrix-free GP learning}
-
 For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$.
@@
 Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
-We considered several stochastic approximations for finding the pathological term $\log\abs{K_{|\bsth}}$ in Eq.~\ref{datall} \cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.
 
@@
 First, we do not construct $K_{|\bsth}$, but a low-memory representation $\tilde{K}_{|\bsth}$, described in Sec.~\ref{fast-mvm}.
-Second, we select MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
-MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}.
+Second, we select \textsc{minres} instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
+\textsc{minres} handles numerically semidefinite matrices with more grace \cite{fong2012cg}.
 This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations because of increases in $\kappa_2$, the spectral condition number of $K$ (Fig.~\ref{fx2007-iterations}).
@@
 \includegraphics[width=0.5\columnwidth]{iterations.eps}
-\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
+\caption{Number of MVMs that \textsc{minres} must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
 The iteration cutoff is the number of training points $n$ in the dataset.
@@
 
-We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$.
+We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$ in the following sections.
 
@@
 COGP is only independent of $R$ because it cannot represent models for $R\neq 1$.
+We distinguish in the up-front cost and per-hyperparameter cost for computing $\nabla \mcL$ to capture the varying asymptotic behaviors of the gradient algorithms as the number of hyperparemters increases: the total cost would be the up-front cost plus $QRD$ multiplied by the per-hyperparameter cost.
   }
@@
 The full predictive covariance estimate requires finding a new term $K_{*,X}K_{X,X}^{-1}K_{X,*}$.
-This is done by solving the linear system in a matrix-free manner on-the-fly; in particular, $K_{X,X}^{-1}K_{X,*}$ is computed via MINRES for every new test point $K_{X,*}$.
+This is done by solving the linear system in a matrix-free manner on-the-fly; in particular, $K_{X,X}^{-1}K_{X,*}$ is computed via \textsc{minres} for every new test point $K_{X,*}$.
 Over several test points, this is embarrassingly parallelizable.
@@
 One example is the sampling-based approach proposed in \cite{papandreou2011efficient}, which extends to linear combinations of kernels that allow fast eigendecompositions.
-
-\subsection{Hyperparameter settings}
-
-AdaDelta parameters were set to the following on all tests: rate $=1$, decay $=0.9$, momentum $=0.5$, offset $=0.0001$.
-The stopping criterion parameters permit the gradient $\infty$-norm to drop below a threshold of its maximum value so far a small, fixed number of times, 5.
-The maximum number of iterations was 100.
-
-For learning, we initialize entries $A_q$ according to a unit normal and all $\bsk_q$ to $1$.
-Note that COGP initializes the coregionalization matrix to $1$ uniformly.
-Like COGP, we initialize noise to $0.1$ for all outputs.
 
@@
 \texttt{mix} refers to a mix of all three kernels.
-Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
+
+For each setting, we randomly sample entries for each $A_q,\bsk_q$, and $\bse$ and the inverse length scale $\gamma$ for each kernel.\footnote{Hyperparameters for the sampling distributions, as well as all code and benchmarking scripts, are available in <anonymous repository>} Then, we investigate the average gradient construction accuracy and speed of LLGP for different settings of $Q,\epsilon$, and \texttt{ktype}.
+
+%Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
 Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
@@
 \begin{table}[!h]
-  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using MINRES for each of the kernel representations.
+  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using \textsc{minres} for each of the kernel representations.
 For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system.
@@
 The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:rellog}).
-Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor MINRES convergence in the inversions.
+Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions.
 We will show that the stochastic gradients suffice for optimization in practical examples.
@@
 In each case, higher is better, and extremal values are noted.
-For each data point, the average was taken over five runs.}
+For each data point, the average was taken over five runs.
+Reducing average noise, corresponding to increases in $-\log_{10}\varepsilon$, and increases in $Q$, generally make gradient reconstruction more difficult by making $K$ more ill-conditioned, reducing the accuracy of \textsc{minres}.}
 \label{fig:rellog}
@@
 All kernels are RBF.
-
 The data used in this example are from 2007, and include $n=3054$ training points and $150$ test points.
 The test points include 50 contiguous points extracted from each of the CAD, JPY, and AUD exchanges.
-
 For this application, LLGP uses $m=n/D=238$ interpolating points.

### 2017-05-13 paper/supplement.tex
Added paper corrections from Li-Fang

--- 2017-05-07 paper/paper.tex
@@
 
-\title{Large Linear Multi-output Gaussian Process Learning for Time Series}
+\title{Supplementary Material for Large Linear Multi-output Gaussian Process Learning for Time Series}
 
@@
 
-\begin{abstract}
-Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization \cite{alvarez2012kernels}.
-Such models can learn correlations across the multiple outputs, which can then be exploited to share knowledge across the multiple outputs.
-For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than the same model applied to a single region.
-While model learning can be performed efficiently for single-output GPs, the multi-output case still requires approximations for large numbers of observations across all outputs.
-In this work, we propose a new method, Large Linear GP (LLGP), which estimates covariance hyperparameters for multi-dimensional outputs and one-dimensional inputs.
-Our approach learns GP kernel hyperparameters at an asymptotically faster rate than the current state of the art.
-When applied to real time series data, we find this theoretical improvement is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
-Finally, we discuss extensions of our approach to multidimensional inputs.
-\end{abstract}
+\section{Implementation Details}
 
-% NOTES
-% 8 pages of content
-% \citet for inline citations
-% footnotes after punctuation
-% no vertical rules in tables
-% \subsubsection*{Acknowledgments} only in final paper
+LLGP was implemented in Python 3 from the Anaconda, which offered an Intel MKL-linked scipy \cite{scipy}.
+The code made heavy use of other packages, namely climin \cite{climin}, GPy \cite{gpy}, and paramz \cite{paramz}.
+Code and benchmarks are available at \texttt{<anonymous repository>}.
 
-\section{Introduction}\label{introduction}
-
-Gaussian processes (GPs) are a popular nonlinear regression method that innately capture function smoothness across inputs as defined by their covariance function \cite{williams1996gaussian}.
-GPs seamlessly extend to multi-output learning, picking up on latent cross-output processes, where in multi-output learning the objective is to build a probabilistic regression model over vector-valued observations.
-
-Multiple-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
-Efficient model learning would enable researchers to quickly explore large spaces of model configurations to find an appropriate one for their task.
-
-Exact GP inference is infeasible in this scenario since it requires maintaining pairwise covariance between all inputs in a large matrix and performing inversions with that matrix \cite{williams1996gaussian}.
-As the number of hyperparameters grows, so do the number of matrix operations (Tab.~\ref{asymp}).
-For this reason, an accurate approximate approach has been an important goal of machine learning research.
-
-\subsection{Contributions}
-
-Our approach makes the following contributions to approximate inference in multi-output GPs, scoped to time series input.
-First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid.
-Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}.
-Both of these structures coordinate for fast matrix-vector multiplication $K\Ty$ with the covariance matrix $K$.
-
-When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}.
-Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.
-
-For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
-
-Our paper is organized as follows.
-In Sec.~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs.
-Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs.
-Sec.~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion.
-Then, in Sec.~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Sec.~\ref{conclusion}.
-
-\section{Background}
-\label{sec:background}
-
-\subsection{General Gaussian processes}
-
-First, we give an overview of the theoretical foundation for GPs.
-A GP is a set of random variables (RVs) $\{f_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that for any finite collection $X\subset\mcX$, the RVs are jointly Gaussian with a prescribed mean function $\bsm:\mcX\rightarrow\R$ and covariance $K:\mcX^2\rightarrow\R$, i.e., $f_X\sim N\pa{\bsm_X, K_{X, X}}$ \cite{williams1996gaussian}.
-As usual, we drop the mean with $\Ty\triangleq f_X-\bsm_X$ and denote vectorization $f_X=\colv{f_{\Tx}}_{\Tx\in X}$ or matricization $K_{X,Z}=\mat{K(\Tx,\Tz)}_{\Tx\in X,\Tz\in Z}$ with subscripts.
-
-Given a set of $n$ observations at each $X$ of the $\R^n$-valued RV $\Ty$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty$ \cite{williams1996gaussian}.
-Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$:
-\begin{align}
-  \mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)=-\frac{1}{2}\Ty^\top K_{|\bsth}^{-1}\Ty-\frac{1}{2}\log \abs{K_{|\bsth}}-\frac{n}{2}\log 2\pi.\label{datall}
-\end{align}
-As such, the heart of GP learning lies in optimization of $\mcL$.
-Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$ using $\bsa=K_{|\bsth}^{-1}\Ty$:
-\begin{align}
-\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}}.\label{llgradient}
-\end{align}
-
-\subsection{Multi-output linear GPs}
-
-We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2012kernels}.
-
-Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$: an input is in a point from a shared domain coupled with an output tag.
-Then, if we make observations at $X_d\subset\mcX$ for output $d\in[D]$, we can set:
-\begin{align*}
-\TX=\set{(d, x)}{d\in[D],x\in X_d}\subset{\mcX'};\;\; n=\card{\TX}.
-\end{align*}
-
-An LMC kernel $K$ is of the form 
-\begin{align}
-K([i,\Tx_i],[j,\Tx_j])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx_i-\Tx_j})+\epsilon_i\delta_{ij},\label{lmcpointwise}
-\end{align} 
-where $k_q:\R\rightarrow\R$ is a stationary kernel on $\mcX$.
-Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\bsk_q I_D$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a preset rank.
-Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz.
-Settings of $B_q$ that arise in practice, where we wish to capture covariance across outputs as an arbitrary $D^2$-dimensional latent process, prevent stationarity in $K$ and therefore prohibit direct application of SKI.
-
-Thus, the LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs.
-The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
-By choosing $B_q$ to have rank $R_q=D$, so the contribution between those two outputs is learned independently of all other contributions, the entries of $B_q$.
-By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction.
-
-\subsection{Structured covariance matrices}
-
-If we can identify structure in the covariance matrices $K$, then we can recover fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$.
-
-The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$-th block $A_{ij}B$.
-We can represent it by keeping representations of $A$ and $B$ separately, rather than the product.
-Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
-For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.
-
-Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_j$, yielding an $O(n)$ representation.
-Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs.
-Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that:
-\[
-T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}
-\]
-Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs $C\mat{\Tx & \textbf{0}}^\top=\mat{T\Tx &\textbf{0}}^\top$ in $O(n\log n)$ time.
-This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.
-
-\section{Related work}
-\label{sec:related-work}
-\subsection{Approximate inference methods}
-
-To cope with the lack of tractable GP inference methods, inducing point approaches create a tractable model to use instead of the classic GP.
-Such approaches fix or estimate inducing points $\TU$ and claim that the data $f_X$ is conditionally deterministic (deterministic training conditional, or DTC), independent (fully independent training conditional, or FITC), or partially independent (partially independent training conditional, or PITC) given RVs $f_{\TU}$  \cite{quinonero2005unifying}.
-These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime.
-Setting the inducing points $\TU=\TX$ recovers the original exact GP.
-Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
-
-\citet{nguyen2014collaborative} observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once.
-COGP further uses a variational approximation, the evidence lower bound, to the log likelihood.
-COGP supports a subset of LMC kernels, namely those that match the SLFM model \cite{seeger2005semiparametric}.
-In an LMC representation (Eq.~\ref{lmcpointwise}), these models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
-Moreover, SLFM and COGP models add in an independent GP to each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.
-
-\subsection{Approaches for stationary kernels}
-
-\subsubsection{Structured Kernel Interpolation (SKI)}\label{ski-section}
-
-SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \cite{kiss-gp}.
-To do this efficiently, SKI relies on the differentiability of $K$.
-For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$.
-The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$.
-Importantly, $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
-
-In order to adapt SKI to our context of multiple outputs, we build grid $\TU\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with $\TU = [D]\times U$.
-Since the LMC kernel evaluated between two sets of outputs $K_{X_i,X_j}$ is differentiable, as long as $U$ contains each $\{X_d\}_{d\in[D]}$, the corresponding SKI approximation $K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top$ holds with the same asymptotic convergence cubic in $1/m$.
-
-Massively Scalable Gaussian Processes (MSGP) observes that the kernel $K_{U,U}$ on a grid exhibits Kronecker and Toeplitz matrix structure \cite{msgp}.
-Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{datall}.
-In addition, \cite{wilson2014fast} mentions an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{datall}.
-
-While evaluating $\log\abs{K_{|\bsth}}$ is not feasible in the LMC setting (because the LMC sum breaks Kronecker and Toeplitz structure), the general notion of creating structure with SKI carries over to LLGP.
-
-\section{Matrix-free LMC learning} \label{sec:matrix-free}
-
-We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches.
-Critically, the accuracy of the method need not be reduced by keeping the number of interpolation points $m$ low because its reliance on structure allows better asymptotic performance.
-For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.
-
-\subsection{Matrix-free GP learning}
-
-For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$.
-Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
-Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
-We considered several stochastic approximations for finding the pathological term $\log\abs{K_{|\bsth}}$ in Eq.~\ref{datall} \cite{dorn2015stochastic, han2015large}, but found these too slow and inaccurate for use in optimization.
-
-\subsection{Gradient construction}
-
-\citet{gibbs1996cient} describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix.
-In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG).
-Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
-\begin{align}
-  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
-\end{align}
-For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
-As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number of $N$ samples from $\tbr$.
-
-We depart from Gibbs and MacKay in two ways, yielding Algorithm~\ref{alg:grad}.
-First, we do not construct $K_{|\bsth}$, but a low-memory representation $\tilde{K}_{|\bsth}$, described in Sec.~\ref{fast-mvm}.
-Second, we select MINRES instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
-MINRES handles numerically semidefinite matrices with more grace \cite{fong2012cg}.
-This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations because of increases in $\kappa_2$, the spectral condition number of $K$ (Fig.~\ref{fx2007-iterations}).
-
-\begin{figure}[!h]
-\centering
-\includegraphics[width=0.5\columnwidth]{iterations.eps}
-\caption{Number of MVMs that MINRES must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
-The iteration cutoff is the number of training points $n$ in the dataset.
-}
-\label{fx2007-iterations}
-\end{figure}
-
-Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K}_{|\bsth})\sqrt{\kappa_2})$ \cite{raykar2007fast}.
-This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K_{|\bsth}$ for each $j$ at least as fast as $\mvm({\tilde{K}_{|\bsth}})$.
-Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K_{|\bsth}$.
-
-\begin{algorithm}[!ht]
-  \caption{Compute an approximation of $\nabla L$.
-Assume \textsc{minres} is the inversion routine.
-We also assume we have access to linear operators $D_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}_{|\bsth}$.} \label{alg:grad}
-\begin{algorithmic}[1]
-  \Procedure{LLGP}{$\tilde{K}_{|\bsth}$, $\Ty$, $N$, $\{D_{\theta_j}\}$}
-  \State $R\gets\idx{\tbr_i}{i=1}{N}$, sampling $\tbr\sim\Unif\{\pm 1\}$.
-\For{$\Tz$ in $\{\Ty\}\cup R$, in parallel}
-\State $K^{-1}\Tz\gets\textsc{minres}(\tilde{K}_{|\bsth},\Tz)$.
-\EndFor
-\State $g\gets \bsz$
-\For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
-\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot D_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
-\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}_{|\bsth}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
-\EndFor
-\State \Return $g$ 
-\EndProcedure
-\end{algorithmic}
-\end{algorithm}
-
-\subsection{Fast MVMs and parsimonious kernels}\label{fast-mvm}
-
-The bottleneck of Algorithm~\ref{alg:grad} is the iterative MVM operations in $\textsc{minres}$.
-Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense, so long as it can reconstruct multiplication with any vector to arbitrary, fixed precision.
-
-When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$, the shared interpolating points for all outputs:
-\begin{align}
-  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}.
-\end{align}
-
-Importantly, the Kronecker structure of Eq.~\ref{lmcgrid} lets us re-use the same grid $K_q$ in a computational sense across the different outputs.
-Recalling our SKI extension to multiple outputs (Sec.~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
-\begin{align}
-  K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top+\bse.\label{lmcski}
-\end{align}
-We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel, so the SKI approximation fails.
-But this fact does not prevent efficient representation or multiplication since the matrix is diagonal.
-Then, the MVM operation $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.
-
-We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$.
-
-\subsubsection{\textsc{sum}: sum representation}
-
-In \textsc{sum}, we represent $K_{\TU,\TU}$ with a $Q$-length list.
-At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, with only the top row maintained.
-In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
-As described before, the Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
-In turn, assuming $D\ll m$, the runtime for each of the $Q$ terms is $O(Dm\log m)$.
-
-\subsubsection{\textsc{bt}: block-Toeplitz representation}
-
-In \textsc{bt}, we notice that $K_{\TU,\TU}$ is a block matrix with blocks $T_{ij}$:
-\begin{align*}
-\sum_qB_q\otimes K_q =\mat{T_{ij}}_{i,j\in[D]^2},\;\; T_{ij}=\sum_qb_{ij}^{(q)}K_q.
-\end{align*}
-On a one-dimensional grid $U$, these matrices are Toeplitz since they are linear combinations of Toeplitz matrices.
-\textsc{bt} requires $D^2$ $m$-sized rows to represent each $T_{ij}$.
-Then, using usual block matrix multiplication, an MVM $K_{\TU,\TU}\Tz$ takes $O(D^2m\log m)$ time.
-
-On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence of Krylov-based inversion routines using fewer MVMs \cite{chan1994circulant}.
-
-\subsubsection{\textsc{slfm}: SLFM representation}
-
-For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average added rank, $R\le D$, and re-write the kernel:
-\begin{align*}
-  K_{\TU,\TU}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
-\end{align*}
-Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank-1.
-Under some re-indexing $q'\in[RQ]$ which flattens the double sum such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
-\begin{align*}
-  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top;
-\end{align*}
-where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}.
-Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-Thus, the \textsc{slfm} representation writes $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix, so MVMs take $O((QR + D)m\log m)$ time.
-
-Because the runtimes of \textsc{bt} and \textsc{slfm} are complimentary in the sense that one performs better than the other when $D^2>QR$ and vice-versa, an algorithm that uses the aforementioned condition between to decide between which representation to use can minimize runtime (Tab.~\ref{asymp}).
-We also found that \textsc{sum} is efficient in practice for $Q=1$.
-
-\begin{table}[!h]
-  \caption{
-   For both LLGP and COGP, $m$ is a configurable parameter which increases up to $n$ to improve accuracy.
-$Q,R,D,\kappa_2$ are coefficients dependent on the setting of the LMC kernel, which has about $QRD$ hyperparameters (Eq.~\ref{lmcpointwise}).
-The resulting asymptotic performance is given in the table.
-COGP is only independent of $R$ because it cannot represent models for $R\neq 1$.
-  }
-\label{asymp}
-\begin{sc}
-\begin{center}
-\begin{small}
-\begin{tabular}{ccc}
-  \toprule
-  Method & Up-front cost for $\nabla \mcL$ & \begin{tabular}{c}Additional cost for $\partial_{\theta_j}\mcL$\\ per hyperparameter\end{tabular}\\
-  \midrule
-  Exact & $n^3$ & $n^2 $\\
-  COGP & $Qm^3$ & $nm$ \\
-  LLGP & $\min(QR, D^2)\sqrt{\kappa_2} (n + m\log m)$ &  $n + m\log m$ \\
-  \bottomrule
-\end{tabular}
-\end{small}
-\end{center}
-\end{sc}
-\end{table}
-
-\subsection{Stopping conditions}\label{stopping}
-
-For a gradient-only stopping heuristic, we maintain the running maximum gradient $\infty$-norm.
-If gradient norms drop below a preset proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate.
-For example, when applied to the foreign exchange rate prediction dataset in Sec.~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).
-
-\begin{figure}[!h]
-\centering
-{\includegraphics[width=0.5\columnwidth]{running_cutoff.eps}}
-\caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
-\label{fx2007-stop}
-\end{figure}
-
-\subsection{Prediction}
-
-The predictive mean can be computed in $O(1)$ time as observed in \cite{msgp} by $K_{*,X}\bsa\approx W_{*,\TU}K_{\TU,\TU}\bsa$.
-
-The full predictive covariance estimate requires finding a new term $K_{*,X}K_{X,X}^{-1}K_{X,*}$.
-This is done by solving the linear system in a matrix-free manner on-the-fly; in particular, $K_{X,X}^{-1}K_{X,*}$ is computed via MINRES for every new test point $K_{X,*}$.
-Over several test points, this is embarrassingly parallelizable.
-
-A more efficient predictive variance algorithm is outside the scope of this paper: for a research setting, we expect training time to be the bottleneck.
-Moreover, one can couple LLGP learning with other prediction mechanisms.
-One example is the sampling-based approach proposed in \cite{papandreou2011efficient}, which extends to linear combinations of kernels that allow fast eigendecompositions.
-
-\subsection{Hyperparameter settings}
-
-AdaDelta parameters were set to the following on all tests: rate $=1$, decay $=0.9$, momentum $=0.5$, offset $=0.0001$.
-The stopping criterion parameters permit the gradient $\infty$-norm to drop below a threshold of its maximum value so far a small, fixed number of times, 5.
-The maximum number of iterations was 100.
-
-For learning, we initialize entries $A_q$ according to a unit normal and all $\bsk_q$ to $1$.
-Note that COGP initializes the coregionalization matrix to $1$ uniformly.
-Like COGP, we initialize noise to $0.1$ for all outputs.
-
-\section{Results}
-\label{sec:results}
-We evaluate the methods on held out data by using standardized mean square error (SMSE) of the test points with the predicted mean, and the negative log predictive density (NLPD) of the Gaussian likelihood of the inferred model.
-Notably, NLPD takes confidence into account, while SMSE only evaluates the mean prediction.
-In both cases, lower values represent better performance.
-
-\subsection{Representation evaluation}\label{empirical-rep}
-
-We evaluated the performance of the different kernel representations over various rank and parameterization settings.
-In particular, we have the following parameters: $n$ total sample size across all outputs, $D$ number of outputs, $Q$ number of kernels $k_q$, $R$ average added rank, $\epsilon$ mean noise, and \texttt{ktype} kernel type.
-Kernel type is one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions.
-\texttt{mix} refers to a mix of all three kernels.
-Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
-Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
-Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
-
-As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$---this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
-For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
-\begin{table}[!h]
-  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using MINRES for each of the kernel representations.
-For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system.
-We averaged over five runs.
-In every run, we use $n=5000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
-}
-\label{complement}
-\begin{center}
-  \begin{small}
-    \input{representation.tex}
-\end{small}
-\end{center}
-\end{table}
-
-We next evaluated the accuracy of the gradients for $N=10$ trace samples.
-Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$.
-Since, for each partial derivative, LLGP requires only $N$ $n$-sized vector dot products (Eq.~\ref{eq:trace}), it generally runs faster than the exact approach (Fig.~\ref{fig:rellog}), which must compute a matrix-matrix dot product (Eq.~\ref{llgradient}).
-The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:rellog}).
-Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor MINRES convergence in the inversions.
-We will show that the stochastic gradients suffice for optimization in practical examples.
-
-\begin{figure}[!h]
-  \centering
-
- \subfigure[time]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{time_ratio.eps}}}
-  \subfigure[gradient accuracy]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{relgrad_l2.eps}}}
-%  \subfigure[inversion accuracy]{
-%    \resizebox{0.40\textwidth}{!}{\includegraphics{relalpha_l2.eps}}}
- \caption{Negative logarithm of the ratio of the exact algorithm to LLGP's analogue for (a) time and (b) accuracy metrics for evaluating gradient computation performance.
-In each case, higher is better, and extremal values are noted.
-For each data point, the average was taken over five runs.}
-\label{fig:rellog}
-\end{figure}
-
-\subsection{Foreign exchange rate prediction (FX2007)}\label{fx2007-results}
-
-We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance.
-The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below.
-The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP.
-COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP.
-All kernels are RBF.
-
-The data used in this example are from 2007, and include $n=3054$ training points and $150$ test points.
-The test points include 50 contiguous points extracted from each of the CAD, JPY, and AUD exchanges.
-
-For this application, LLGP uses $m=n/D=238$ interpolating points.
-We use the COGP settings from the paper.\footnote{COGP hyperparameters for FX2007 were 100 inducing points, 500 iterations, 200 mini-batch size.} LLGP, for both LMC, outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).
-
-\begin{table}[!h]
-  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset.
-Parenthesized values are standard error.
-LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points.
-COGP used a $Q=2$ kernel with $100$ inducing points.}
-\label{fx2007-tbl}
-\begin{center}
-  \begin{small}
-    \begin{sc}
-      \input{results_fx2007.tex}
-\end{sc}
-\end{small}
-\end{center}
-\end{table}
-
-\begin{figure}[!h]
-\centering
-{\includegraphics[width=\textwidth]{fx2007graph.pdf}}
-\caption{Test outputs for the FX2007 dataset.
-COGP mean is black, with 95\% confidence intervals shaded in grey.
-LLGP mean is a solid red curve, with light green 95\% confidence intervals.
-Magenta points are in the training set, while blue ones are in the test set.
-Notice LLGP variance corresponds to an appropriate level of uncertainty on the test set and certainty on the training set, as opposed to the uniform variance from COGP.}
-\label{fx2007-graph}
-\end{figure}
-
-\subsection{Weather dataset}\label{large-bench}
-
-Next, we replicate results from a weather dataset, a large time series used to validate COGP.
-In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure.
-Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15789$ training measurements and $374$ test measurements.
-
-We use the COGP parameters that were set by default in the code provided by the authors.\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}.
-The weather dataset was run on 1500 iterations, mini-batch size 1000.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
-We tested LLGP models on different numbers of interpolating points $m$.
-
-\begin{table}[!h]
-  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the weather dataset.
-Parenthesized values are standard error.
-Both LLGP and COGP trained the SLFM model.
-We show LLGP with $500$ and $1000$ interpolating points and COGP with $200$ inducing points.}
-\label{weather-tbl}
-\begin{center}
-  \begin{small}
-    \begin{sc}
-      \input{results_weather.tex}
-\end{sc}
-\end{small}
-\end{center}
-\end{table}
-
-LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tbl.~\ref{weather-tbl}).
-Varying the number of interpolating points $m$ from $500$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}).
-While NLPD improvement diminishes as $m$ increases, LLGP is still an improvement compared to COGP for a range of $m$ by an order of magnitude in runtime and almost two orders of magnitude for NLPD.
-
-\begin{figure}[!h]
-\centering
-{\includegraphics[width=0.5\columnwidth]{m_time_nlpd.eps}}
-\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
-\label{fig:llgpweather}
-\end{figure}
-\section{Conclusion}\label{conclusion}
-
-LLGP recovers speedups from SKI \cite{kiss-gp} for the problem of multi-output GP regression by recognizing structure unique to LMC kernels, and otherwise not necessarily recoverable in general multi-output kernels.
-This structure further enables a parsimonious representation that allows even large GPs to be learned without explicit construction of the covariance matrix.
-
-LLGP provides a means to approximate the log-likelihood function gradients through interpolation.
-We have shown on several datasets that this can be done in a way that is faster and leads to more accurate results than variational approximations.
-
-Future work would extend the inputs to accept multiple dimensions.
-This can be done without losing internal structure in the kernel \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
-The cubic interpolation requires and exponential number of terms, so projection into lower dimensions learned in a supervised manner would be essential.
-Another useful line for investigation would be more informed stopping heuristics.
-Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \cite{cutajar2016preconditioning}.
-
-% \begin{table}[t]
-%   \caption{Sample table title}
-%   \label{sample-table}
-%   \centering
-%   \begin{tabular}{lll}
-%     \toprule
-%     \multicolumn{2}{c}{Part}                   \\
-%     \cmidrule{1-2}
-%     Name     & Description     & Size ($\mu$m) \\
-%     \midrule
-%     Dendrite & Input terminal  & $\sim$100     \\
-%     Axon     & Output terminal & $\sim$10      \\
-%     Soma     & Cell body       & up to $106$  \\
-%     \bottomrule
-%   \end{tabular}
-% \end{table}
+Application of our approach to all replication studies was carried out on a large server in a multiprogramming environment: CentOS 6.5 with 80 Intel(R) Xeon(R) CPU E7-4870 @ 2.40GHz.
+The representation evaluation benchmarks were done at once on a cluster of machines running CentOS 5.2-5.9 with Intel(R) Xeon(R) CPU E5430 @ 2.66GHz, where these jobs ran on a single thread per CPU.
 

### 2017-05-13 paper/paper.tex
Remove representation eval section, move it to supplement

@@
 In both cases, lower values represent better performance.
-
-\subsection{Representation evaluation}\label{empirical-rep}
-
-We evaluated the performance of the different kernel representations over various rank and parameterization settings.
-In particular, we have the following parameters: $n$ total sample size across all outputs, $D$ number of outputs, $Q$ number of kernels $k_q$, $R$ average added rank, $\epsilon$ mean noise, and \texttt{ktype} kernel type.
-Kernel type is one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions.
-\texttt{mix} refers to a mix of all three kernels.
-
-For each setting, we randomly sample entries for each $A_q,\bsk_q$, and $\bse$ and the inverse length scale $\gamma$ for each kernel.\footnote{Hyperparameters for the sampling distributions, as well as all code and benchmarking scripts, are available in <anonymous repository>} Then, we investigate the average gradient construction accuracy and speed of LLGP for different settings of $Q,\epsilon$, and \texttt{ktype}.
-
-%Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
-Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
-Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
-
-As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$---this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
-For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
-\begin{table}[!h]
-  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using \textsc{minres} for each of the kernel representations.
-For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system.
-We averaged over five runs.
-In every run, we use $n=5000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
-}
-\label{complement}
-\begin{center}
-  \begin{small}
-    \input{representation.tex}
-\end{small}
-\end{center}
-\end{table}
-
-We next evaluated the accuracy of the gradients for $N=10$ trace samples.
-Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$.
-Since, for each partial derivative, LLGP requires only $N$ $n$-sized vector dot products (Eq.~\ref{eq:trace}), it generally runs faster than the exact approach (Fig.~\ref{fig:rellog}), which must compute a matrix-matrix dot product (Eq.~\ref{llgradient}).
-The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:rellog}).
-Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions.
-We will show that the stochastic gradients suffice for optimization in practical examples.
-
-\begin{figure}[!h]
-  \centering
-
- \subfigure[time]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{time_ratio.eps}}}
-  \subfigure[gradient accuracy]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{relgrad_l2.eps}}}
-%  \subfigure[inversion accuracy]{
-%    \resizebox{0.40\textwidth}{!}{\includegraphics{relalpha_l2.eps}}}
- \caption{Negative logarithm of the ratio of the exact algorithm to LLGP's analogue for (a) time and (b) accuracy metrics for evaluating gradient computation performance.
-In each case, higher is better, and extremal values are noted.
-For each data point, the average was taken over five runs.
-Reducing average noise, corresponding to increases in $-\log_{10}\varepsilon$, and increases in $Q$, generally make gradient reconstruction more difficult by making $K$ more ill-conditioned, reducing the accuracy of \textsc{minres}.}
-\label{fig:rellog}
-\end{figure}
+We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact log-likelihood gradients using the standard Cholesky algorithm for a grid of different kernels; this evaluation is available in the paper supplement.\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in <anonymous repository>.}
 

### 2017-05-13 paper/supplement.tex
Remove representation eval section, move it to supplement

@@
 
+\subsection{Representation evaluation}\label{empirical-rep}
+
+We evaluated the performance of the different kernel representations over various rank and parameterization settings.
+In particular, we have the following parameters: $n$ total sample size across all outputs, $D$ number of outputs, $Q$ number of kernels $k_q$, $R$ average added rank, $\epsilon$ mean noise, and \texttt{ktype} kernel type.
+Kernel type is one of \texttt{mat, periodic, rbf, mix} corresponding to Mat\'{e}rn-3/2, standard periodic\footnote{We define the periodic kernel as $k(r) = \exp \left(\frac{-\gamma}{2}\sin^2\frac{\pi r}{T}\right)$.}, and radial basis functions.
+\texttt{mix} refers to a mix of all three kernels.
+
+For each setting, we randomly sample entries for each $A_q,\bsk_q$, and $\bse$ and the inverse length scale $\gamma$ for each kernel.
+Then, we investigate the average gradient construction accuracy and speed of LLGP for different settings of $Q,\epsilon$, and \texttt{ktype}.Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
+Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
+Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
+
+As expected from their asymptotic runtime, \textsc{sum}, \textsc{bt}, and \textsc{slfm} representations are complimentary in MVM speed for different configurations of $D,R,Q$---this results in sparse inversion computation that consistently outperforms Cholesky decomposition in runtime (Tab.~\ref{complement}).
+For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
+\begin{table}[!h]
+  \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using \textsc{minres} for each of the kernel representations.
+For comparison, the \textsc{chol} representation is wallclock time to compute the Cholesky decomposition of the matrix, which must be constructed, and use this decomposition to invert the system.
+We averaged over five runs.
+In every run, we use $n=5000$ simulated data points, $\texttt{mix}$ kernels, and $\epsilon=0.1$.
+}
+\label{complement}
+\begin{center}
+  \begin{small}
+    \input{representation.tex}
+\end{small}
+\end{center}
+\end{table}
+
+We next evaluated the accuracy of the gradients for $N=10$ trace samples.
+Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$.
+Since, for each partial derivative, LLGP requires only $N$ $n$-sized vector dot products, it generally runs faster than the exact approach (Fig.~\ref{fig:rellog}), which must compute a matrix-matrix dot product.
+The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}).
+Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions (Fig.~\ref{fig:rellog}.
+We will show that the stochastic gradients suffice for optimization in practical examples.
+
+\begin{figure}[!h]
+  \centering
+
+ \subfigure[time]{
+    \resizebox{0.48\textwidth}{!}{\includegraphics{time_ratio.eps}}}
+  \subfigure[inversion accuracy]{
+    \resizebox{0.48\textwidth}{!}{\includegraphics{relalpha_l2.eps}}}
+ \caption{Negative logarithm of the ratio of the exact algorithm to LLGP's analogue for (a) time and (b) accuracy metrics for evaluating gradient computation performance.
+In each case, higher is better, and extremal values are noted.
+For each data point, the average was taken over five runs.
+Reducing average noise, corresponding to increases in $-\log_{10}\varepsilon$, and increases in $Q$, generally make gradient reconstruction more difficult by making $K$ more ill-conditioned, reducing the accuracy of \textsc{minres}.}
+\label{fig:rellog}
+\end{figure}
+
+\begin{figure}[!h]
+\centering
+{\includegraphics[width=\textwidth]{relgrad_l2.eps}}
+\caption{Negative logarithm of the relative error in $\nabla\mcL$ between LLGP and the exact computation, over the same grid of kernels as Fig.~\ref{fig:rellog}}\label{fig:relgrad}
+\end{figure}
+
 \bibliographystyle{unsrtnat}

### 2017-05-14 paper/supplement.tex
Updated supplement to reflect truncated normal initialization

@@
 \usepackage{booktabs}
-\usepackage{subfigure}
+\usepackage[section]{placeins}
 
@@
 
-\subsection{Representation evaluation}\label{empirical-rep}
+\section{Representation evaluation}\label{empirical-rep}
 
@@
 Then, we investigate the average gradient construction accuracy and speed of LLGP for different settings of $Q,\epsilon$, and \texttt{ktype}.Each kernel's inverse length scales and periods were selected by sampling uniformly in log space from 1 to 10 with $Q$ samples.
-Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
+Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution truncated to the unit interval, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
 Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
@@
 We next evaluated the accuracy of the gradients for $N=10$ trace samples.
-Fixing $R=3,n=5000,D=10$, we quantified the accuracy and speed of computing $\nabla\mcL$.
-Since, for each partial derivative, LLGP requires only $N$ $n$-sized vector dot products, it generally runs faster than the exact approach (Fig.~\ref{fig:rellog}), which must compute a matrix-matrix dot product.
-The gradients between the two, however, are virtually indistinguishable for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}).
+Fixing $R=3,n=5000,D=10$, we quantified the accuracy LLGP's $\nabla\mcL$ by comparing against the exact Cholesky approach.
+The relative error in the gradients is generally low for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}).
 Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions (Fig.~\ref{fig:rellog}.
-We will show that the stochastic gradients suffice for optimization in practical examples.
+We showed in the main paper that the stochastic gradients suffice for optimization in practical examples.
 
@@
   \centering
-
- \subfigure[time]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{time_ratio.eps}}}
-  \subfigure[inversion accuracy]{
-    \resizebox{0.48\textwidth}{!}{\includegraphics{relalpha_l2.eps}}}
- \caption{Negative logarithm of the ratio of the exact algorithm to LLGP's analogue for (a) time and (b) accuracy metrics for evaluating gradient computation performance.
+{\includegraphics[width=\textwidth]{relgrad_l2.eps}}
+\caption{Negative logarithm of the relative error from the LLGP gradient construction to the exact log likelihood gradient.
 In each case, higher is better, and extremal values are noted.
@@
 Reducing average noise, corresponding to increases in $-\log_{10}\varepsilon$, and increases in $Q$, generally make gradient reconstruction more difficult by making $K$ more ill-conditioned, reducing the accuracy of \textsc{minres}.}
-\label{fig:rellog}
+\label{fig:relgrad}
 \end{figure}
@@
 \centering
-{\includegraphics[width=\textwidth]{relgrad_l2.eps}}
-\caption{Negative logarithm of the relative error in $\nabla\mcL$ between LLGP and the exact computation, over the same grid of kernels as Fig.~\ref{fig:rellog}}\label{fig:relgrad}
+{\includegraphics[width=\textwidth]{relalpha_l2.eps}}
+\caption{Negative logarithm of the relative error in $K^{-1}\Ty$, using \textsc{minres} compared to the Cholesky solution.
+Higher is better, and extremal values are noted.
+An average was taken over five runs.}
+\label{fig:rellog}
 \end{figure}

### 2017-05-16 paper/paper.tex
Updated supplement to reflect truncated normal initialization
Incorporated some of Prof. E's changes

@@
 \usepackage{booktabs}
-\usepackage{subfigure}
 
@@
 \author{
-  Vladimir Feinberg, Li-Fang Cheng, {Barbara Engelhardt}, {Kai Li}\\
+  Vladimir Feinberg, Li-Fang Cheng,  {Kai Li}, {Barbara E Engelhardt}\\
   Department of Computer Science\\
   Princeton University\\
-  Princeton, NJ 08544 \\
-  \texttt{\{vyf,lifangc,bee,li\}@princeton.edu} \\
+  Princeton, NJ 08540 \\
+  \texttt{\{vyf,lifangc,li,bee\}@princeton.edu} \\
 }
@@
 \begin{abstract}
-Gaussian process (GP) models, which put a distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case; a common way of doing this is to use a linear model of coregionalization (LMC) \cite{alvarez2012kernels}.
-Such models can learn and exploit correlations across the multiple outputs.
-For instance, temperature data from disparate regions over time can contribute to a predictive weather model that is more accurate than a single, localized model.
-While model learning can be performed efficiently for single-output GPs \cite{msgp}, such methods assume stationarity, a luxury unavailable in the multi-output case.
-We propose Large Linear GP (LLGP), which recovers stationarity thanks to LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and one-dimensional inputs.
-Our approach learns hyperparameters at an asymptotically faster rate than the current state of the art.
-When applied to real time series data, we find this theoretical improvement is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
-Finally, we discuss extensions of our approach to multidimensional inputs.
+Gaussian processes, or distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \cite{alvarez2012kernels}.
+LMCs estimate and exploit correlations across the multiple outputs.
+While model estimation can be performed efficiently for single-output GPs \cite{msgp}, these assume stationarity, but in the multi-output case the cross-covariance interaction is not stationary.
+We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and one-dimensional inputs.
+When applied to real time series data, we find our theoretical improvement relative to the current state of the art is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
 \end{abstract}
@@
 
-Gaussian processes (GPs) are a popular nonlinear regression method that innately capture function smoothness across inputs as defined by their covariance function \cite{williams1996gaussian}.
-GPs seamlessly extend to multi-output learning, picking up on latent cross-output processes, where in multi-output learning the objective is to build a probabilistic regression model over vector-valued observations.
+Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \cite{williams1996gaussian}.
+GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.
 
-Multiple-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations and reconstructing missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
-Efficient model learning would enable researchers to quickly explore large spaces of model configurations to find an appropriate one for their task.
+Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
+Efficient model estimation would enable researchers to quickly explore large spaces of parameterizations to find an appropriate one for their task.
 
-Exact GP inference is infeasible in this scenario since it requires maintaining pairwise covariance between all inputs in a large matrix and performing inversions with that matrix \cite{williams1996gaussian}.
-As the number of hyperparameters grows, so do the number of matrix operations (Tab.~\ref{asymp}).
+For $n$ input points, exact GP inference requires maintaining the $n^2$ pairwise covariances between inputs in a matrix explicitly and performing inversions with that matrix \cite{williams1996gaussian}.
+In the multi-output setting, the number of hyperparameters grows (since we must consider cross-output covariance), in turn requiring more matrix operations (Tab.~\ref{asymp}).
 For this reason, an accurate approximate approach has been an important goal of machine learning research.
-
-\subsection{Contributions}
-
-Our approach makes the following contributions to approximate inference in multi-output GPs, scoped to time series input.
-First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid.
-Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}.
-Both of these structures coordinate for fast matrix-vector multiplication $K\Tz$ with the covariance matrix $K$ for any vector $\Tz$.
-
-When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}.
-Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.
-
-For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
+The approximations developed for multi-output methods effectively reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}.
+The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure.
+We use this structure to avoid matrix inversion.
 
@@
 In Sec.~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs.
-Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs.
+Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs, followed by Sec.~\ref{sec:contributions} describing our contributions.
 Sec.~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion.
@@
 
-\subsection{General Gaussian processes}
+\subsection{Gaussian processes (GPs)}
 
-First, we give an overview of the theoretical foundation for GPs.
-A GP is a set of random variables (RVs) $\{f_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}.
-With zero mean wlog and a prescribed covariance $K:\mcX^2\rightarrow\R$, $\Tf_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Tf_X)_i=f_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
-Given observations $\Ty\in\R^n$ of $\Tf_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Tf_X$ \cite{williams1996gaussian}.
-Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Tf_{X}|X,\bsth)$.
-As such, the heart of GP learning lies in optimization of $\mcL$.
+A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}.
+With zero mean wlog and a prescribed covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
+Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
+Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$.
 Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
@@
 We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2012kernels}.
-Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$: an input is in a point from a shared domain coupled with an output tag.
+Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$; a point from a shared domain coupled with an output index.
 Then, if we make observations at $X_d\subset\mcX$ for output $d\in[D]$, we can set:
@@
 Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz.
-Settings of $B_q$ that arise in practice, where we wish to capture covariance across outputs as an arbitrary $D^2$-dimensional latent process, prevent stationarity in $K$ and therefore prohibit direct application of SKI.
+In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx_i],[j,\Tx_j])\neq K([i+1,\Tx_i+1],[j+1,\Tx_j+1])$.
 
-The LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between to inputs for two different outputs.
+The LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between two inputs for two different outputs.
 The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
-By choosing $B_q$ to have rank $R_q=D$, so the contribution between those two outputs is learned independently of all other contributions, the entries of $B_q$.
-By lowering the rank $R_q$, researchers applying the LMC model can specify that the contributions of the kernels for each output pair are determined by lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$ then we have an independent GP for each output).
+By choosing $B_q$ to have rank $R_q=D$, the contribution of the two outputs is estimated independently from all other contributions.
+By lowering the rank $R_q$, we specify that the interactions of the outputs have lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$ then we have an independent GP for each output).
 
@@
 
-If we can identify structure in the covariance matrices $K$, then we can recover fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$.
+If we can identify structure in the covariance $K$, then we can recover fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$.
 
@@
 Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
-For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs by a $(1/P)$-th power \cite{gilboa2015scaling}.
+For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs of the exact GP algorithm by a $(1/P)$-th power \cite{gilboa2015scaling}.
 
-Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_j$, yielding an $O(n)$ representation.
+Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_{j=1}^n$, yielding an $O(n)$ representation.
 Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs.
@@
 
-To cope with the lack of tractable GP inference methods, inducing point approaches create a tractable model to use instead of the classic GP.
-Such approaches fix or estimate inducing points $\TU$ and claim that the data $f_X$ is conditionally deterministic (deterministic training conditional, or DTC), independent (fully independent training conditional, or FITC), or partially independent (partially independent training conditional, or PITC) given RVs $f_{\TU}$  \cite{quinonero2005unifying}.
+Inducing point approaches create a tractable model to use instead of the exact GP.
+Such approaches fix or estimate inducing points $\TU$ and claim that the data $\Ty_X$ is Gaussian when conditioned on the RVs $\Ty_{\TU}$  \cite{quinonero2005unifying}.
 These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime.
-Setting the inducing points $\TU=\TX$ recovers the original exact GP.
+Setting the inducing points $\TU=\TX$ recovers the exact GP.
 Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
-
+% TODO expand; describe COGP more
 \citet{nguyen2014collaborative} observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once.
@@
 
+\section{Contributions}\label{sec:contributions}
+
+% TODO: contributions don't have edits yet
+
+First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid.
+Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}.
+Both of these structures coordinate for fast matrix-vector multiplication $K\Tz$ with the covariance matrix $K$ for any vector $\Tz$.
+
+When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}.
+Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.
+
+For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
+
 \section{Large Linear GP} \label{sec:matrix-free}
@@
 
-\citet{gibbs1996cient} describe the algorithm for GP model learning in terms of only MVMs with the covariance matrix.
+\citet{gibbs1996cient} describe the algorithm for GP model estimation in terms of only MVMs with the covariance matrix.
 In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG).

### 2017-05-19 paper/paper.tex
Incorporated v2 edits from Prof. E

@@
 \usepackage{booktabs}
+\usepackage{subfigure}
+\usepackage{wrapfig}
 
@@
 \begin{abstract}
-Gaussian processes, or distribution over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \cite{alvarez2012kernels}.
+Gaussian processes, or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \cite{alvarez2012kernels}.
 LMCs estimate and exploit correlations across the multiple outputs.
@@
 
-For $n$ input points, exact GP inference requires maintaining the $n^2$ pairwise covariances between inputs in a matrix explicitly and performing inversions with that matrix \cite{williams1996gaussian}.
-In the multi-output setting, the number of hyperparameters grows (since we must consider cross-output covariance), in turn requiring more matrix operations (Tab.~\ref{asymp}).
-For this reason, an accurate approximate approach has been an important goal of machine learning research.
-The approximations developed for multi-output methods effectively reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}.
+% TODO: is this paragraph OK now?
+% yes, except you need to cite more.
+For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \cite{williams1996gaussian}.
+Some single-output GP methods exploit structure in this matrix to reduce runtime \cite{msgp}.
+In the multi-output setting, the same structure does not exist.
+Approximations developed for multi-output methods instead reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}.
 The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure.
-We use this structure to avoid matrix inversion.
+We use this structure to avoid direct matrix inversion.
 
 Our paper is organized as follows.
-In Sec.~\ref{sec:background} we give a background on single-output and multi-output GPs, as well as some history in structured linear algebra in GPs.
+In Sec.~\ref{sec:background} we give background on single-output and multi-output GPs, as well as some history in exploiting structure for matrix inversions in GPs.
 Sec.~\ref{sec:related-work} details both related work that was built upon in LLGP and existing methods for multi-output GPs, followed by Sec.~\ref{sec:contributions} describing our contributions.
-Sec.~\ref{sec:matrix-free} describes our method, including a matrix-free heuristic stopping criterion.
+Sec.~\ref{sec:matrix-free} describes our method.
 Then, in Sec.~\ref{sec:results} we compare the performance of LLGP to existing methods and offer concluding remarks in Sec.~\ref{conclusion}.
@@
 Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
-Predictive accuracy is sensitive to a particular parameterization of our kernel, so model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$.
+Predictive accuracy is sensitive to a particular parameterization of our kernel, and model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$.
 Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
@@
 
-We build multi-output GP models as instances of general GPs, where a multi-output GP model identifies correlations between outputs through a shared input space \cite{alvarez2012kernels}.
-Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$; a point from a shared domain coupled with an output index.
+We build multi-output GP models as instances of general GPs, where a multi-output model explicitly represents correlations between outputs through a shared input space \cite{alvarez2012kernels}.
+Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$, a point from a shared domain coupled with an output index.
 Then, if we make observations at $X_d\subset\mcX$ for output $d\in[D]$, we can set:
@@
 The LMC kernel provides a flexible way of specifying multiple additive contributions to the covariance between two inputs for two different outputs.
-The contribution of the $q$-th kernel $k_q$ to the covariance between the $i$-th and $j$-th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
-By choosing $B_q$ to have rank $R_q=D$, the contribution of the two outputs is estimated independently from all other contributions.
-By lowering the rank $R_q$, we specify that the interactions of the outputs have lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$ then we have an independent GP for each output).
+The contribution of the $q$th kernel $k_q$ to the covariance between the $i$th and $j$th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
+By choosing $B_q$ to have rank $R_q=D$, the corresponding LMC model can have any positive contribution between two outputs that best fits the data, so long as $B_q$ remains PSD.
+By reducing the rank $R_q$, the interactions of the outputs have lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$, then we have an independent GP for each output).
 
@@
 
-If we can identify structure in the covariance $K$, then we can recover fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$.
-
-The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$-th block $A_{ij}B$.
-We can represent it by keeping representations of $A$ and $B$ separately, rather than the product.
+If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation.
+The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$th block $A_{ij}B$.
+We can represent the product by keeping representations of $A$ and $B$ separately, rather than the product.
 Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
-For GPs on uniform dimension-$P$ grids, this approximately reduces the runtime and representation costs of the exact GP algorithm by a $(1/P)$-th power \cite{gilboa2015scaling}.
+For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O(n^{3/P})$ \cite{gilboa2015scaling}.
 
@@
 \[
-T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}
+T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}.
 \]
@@
 
-Inducing point approaches create a tractable model to use instead of the exact GP.
-Such approaches fix or estimate inducing points $\TU$ and claim that the data $\Ty_X$ is Gaussian when conditioned on the RVs $\Ty_{\TU}$  \cite{quinonero2005unifying}.
-These approaches are agnostic to kernel stationarity, and their quality can be improved by increasing the number of inducing points at the cost of a longer runtime.
-Setting the inducing points $\TU=\TX$ recovers the exact GP.
+Inducing point approaches create a tractable model to approximate the exact GP.
+For example, the deterministic training conditional (DTC) for a single-output GP fixes inducing points $T\subset\mcX$ and estimates kernel hyperparameters for $\Ty_{X}|\Ty_{T}\sim N(K_{X,T}K_{T,T}^{-1}\Ty_T,\sigma^2)$ \cite{quinonero2005unifying}.
+This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $\textbf{T}\subset \mcX'$, with the model equal to the original GP model when $\textbf{T}=\TX$ \cite{alvarez2010efficient}.
 Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
-% TODO expand; describe COGP more
-\citet{nguyen2014collaborative} observed in Collaborative Multi-output Gaussian Processes (COGP) that multi-output GP algorithms can further share an internal representation of the covariance structure among all outputs at once.
-COGP further uses a variational approximation, the evidence lower bound, to the log likelihood.
-COGP supports a subset of LMC kernels, namely those that match the SLFM model \cite{seeger2005semiparametric}.
-In an LMC representation (Eq.~\ref{lmcpointwise}), these models correspond to all $\bsk_q$ set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
-Moreover, SLFM and COGP models add in an independent GP to each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.
+
+In Collaborative Multi-output Gaussian Processes (COGP), multi-output GP algorithms further share an internal representation of the covariance structure among all outputs at once~\cite{nguyen2014collaborative}.
+COGP fixes inducing points $\textbf{T}=[D]\times T$ for some $m$-sized $T\subset \mcX$ and puts a GP prior on $\Ty_{\textbf{T}}$ with a restricted LMC kernel that matches the Semiparametric Latent Factor Model (SLFM) \cite{seeger2005semiparametric}.
+Applying the COGP prior to $\Ty_X$ corresponds to an LMC kernel (Eq.~\ref{lmcpointwise}) where $\bsk_q$ is set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
+Moreover, SLFM and COGP models include an independent GP for each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.
+COGP uses its shared structure to derive efficient expressions for variational inference for parameter estimation.
 
@@
 $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
+SKI thus creates an $m$-rank approximation to $K_{X,X}$, reducing representation size even without relying on structure.
 
-In order to adapt SKI to our context of multiple outputs, we build grid $\TU\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with $\TU = [D]\times U$.
-Since the LMC kernel evaluated between two sets of outputs $K_{X_i,X_j}$ is differentiable, as long as $U$ contains each $\{X_d\}_{d\in[D]}$, the corresponding SKI approximation $K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top$ holds with the same asymptotic convergence cubic in $1/m$.
+Massively Scalable Gaussian Processes (MSGP) exploits structure as well: the kernel $K_{U,U}$ on a grid has Kronecker and Toeplitz matrix structure \cite{msgp}.
+Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
+In addition, an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{llgradient} has been noted previously \cite{wilson2014fast}.
 
-Massively Scalable Gaussian Processes (MSGP) observes that the kernel $K_{U,U}$ on a grid exhibits Kronecker and Toeplitz matrix structure \cite{msgp}.
-Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
-In addition, \cite{wilson2014fast} mentions an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{llgradient}.
+Although evaluating $\log\abs{K_{|\bsth}}$ is not feasible in the LMC setting because the LMC sum breaks Kronecker and Toeplitz structure, the approach of creating structure with SKI carries over to LLGP.
 
-While evaluating $\log\abs{K_{|\bsth}}$ is not feasible in the LMC setting (because the LMC sum breaks Kronecker and Toeplitz structure), the general notion of creating structure with SKI carries over to LLGP.
+\section{Contributions of LLGP}\label{sec:contributions}
 
-\section{Contributions}\label{sec:contributions}
-
-% TODO: contributions don't have edits yet
-
-First, we identify a block-Toeplitz structure induced by the linear model of coregionalization (LMC) kernel on a grid.
-Next, we adapt a previously identified kernel structure based on Semiparametric Latent Factor Models (SLFM) \cite{seeger2005semiparametric}.
+First, we identify a block-Toeplitz structure induced by the LMC kernel evaluated on a uniformly spaced grid of inputs.
+Next, we show how an LMC kernel can be decomposed into two block diagonal components, one of which has structure similar to that of SLFM \cite{seeger2005semiparametric}.
 Both of these structures coordinate for fast matrix-vector multiplication $K\Tz$ with the covariance matrix $K$ for any vector $\Tz$.
 
-When inputs do not lie on a uniform grid (for time series, the length of time between observations differs), we demonstrate how to leverage structured kernel interpolation (SKI) in the multi-output setting, where SKI, which requires stationary kernels, does not naturally apply \cite{kiss-gp}.
-Because LMC kernels exhibit the previously-identified block-Toeplitz structure, they harmonize with SKI.
+With multiple outputs, non-differentiable cross-covariance interactions violate the assumptions of SKI's cubic interpolation.
+We show  a modification to SKI is compatible with the piecewise-differentiable LMC kernel.
+This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel even for non-uniform inputs, which we use to accelerate GP estimation in the general case.
 
-For low-dimensional inputs, the above contributions offers an asymptotic and practical runtime improvement in hyperparameter learning while also expanding feasible kernel typs to arbitrary differntiable LMC kernels, with improvement taken to be relative to existing multi-GP methods (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
+For low-dimensional inputs, the above contributions offer an asymptotic and practical runtime improvement in hyperparameter estimation while also expanding the feasible kernel families to any differentiable LMC kernels, relative to COGP (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
 
@@
 We propose a linear model of coregionalization (LMC) method based on recent structure-based optimizations for GP estimation instead of variational approaches.
-Critically, the accuracy of the method need not be reduced by keeping the number of interpolation points $m$ low because its reliance on structure allows better asymptotic performance.
+Critically, the accuracy of the method need not be reduced by keeping the number of interpolation points $m$ low, because its reliance on structure allows better asymptotic performance.
 For simplicity, our work focuses on multi-dimensional outputs, one-dimensional inputs, and Gaussian likelihoods.
@@
 \end{align}
-For this approximation, the number of samples need not be large, and the estimate improves as the size of $K_{|\bsth}$ increases.
-As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number of $N$ samples from $\tbr$.
+For this approximation, the number of samples may be small, and estimates improve as the size of $K_{|\bsth}$ increases.
+As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
 
-We depart from Gibbs and MacKay in two ways, yielding Algorithm~\ref{alg:grad}.
-First, we do not construct $K_{|\bsth}$, but a low-memory representation $\tilde{K}_{|\bsth}$, described in Sec.~\ref{fast-mvm}.
-Second, we select \textsc{minres} instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
-\textsc{minres} handles numerically semidefinite matrices with more grace \cite{fong2012cg}.
-This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks over the course of learning, making inversion-based methods require additional iterations because of increases in $\kappa_2$, the spectral condition number of $K$ (Fig.~\ref{fx2007-iterations}).
+We depart from Gibbs and MacKay in two important ways (Algorithm~\ref{alg:grad}).
+First, we do not construct $K_{|\bsth}$, but instead keep a low-memory representation $\tilde{K}_{|\bsth}$ (Sec.~\ref{fast-mvm}).
+Second, we use \textsc{minres} instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
+Iterative \textsc{minres} solutions to numerically semidefinite matrices monotonically improve in practice, as opposed to LCG \cite{fong2012cg}.
+This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks throughout learning.
+Inversion-based methods then require additional iterations because $\kappa_2$, the spectral condition number of $K_{|\bsth}$, increases as $K_{|\bsth}$ becomes less diagonally dominant (Fig.~\ref{fx2007-iterations}).
 
 \begin{figure}[!h]
-\centering
-\includegraphics[width=0.5\columnwidth]{iterations.eps}
-\caption{Number of MVMs that \textsc{minres} must perform at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
-The iteration cutoff is the number of training points $n$ in the dataset.
+\begin{center}
+  \subfigure[]{\includegraphics[width=0.4\columnwidth]{iterations.eps}}
+  \subfigure[]{\includegraphics[width=0.4\columnwidth]{running_cutoff.eps}}
+\end{center}
+\caption{Trace of (a) the number of MVMs that \textsc{minres} must perform to invert $K_{|\bsth}^{-1}\Ty$ and (b) $\mcL, \norm{\nabla\mcL}$ given $\bsth$ at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
+In (b), in green, we have 20\% of the rolling maximum $\infty$-norm of previous gradients.
 }
@@
 Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K_{|\bsth}$.
+
+For a gradient-only stopping heuristic, we maintain the running maximum gradient $\infty$-norm.
+If gradient norms drop below a proportion of the running max norm more than a pre-set number of times, we terminate (Fig.~\ref{fx2007-iterations}).
 
@@
 The bottleneck of Algorithm~\ref{alg:grad} is the iterative MVM operations in $\textsc{minres}$.
-Since $K_{|\bsth}$ only enters computation as an MVM operator, the amount of memory consumed is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense, so long as it can reconstruct multiplication with any vector to arbitrary, fixed precision.
+Since $K_{|\bsth}$ only enters computation as an MVM operator, the required memory is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense as long as we can perform MVM with any vector to arbitrary, fixed precision.
 
-When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise Eq.~\ref{lmcgrid} over $\TU$ holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated at pairs of $U\times U$, the shared interpolating points for all outputs:
+When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise over $\TU$  (Eq.~\ref{lmcgrid}) holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated the shared interpolating points $U$ for all outputs:
 \begin{align}
@@
 \end{align}
-
-Importantly, the Kronecker structure of Eq.~\ref{lmcgrid} lets us re-use the same grid $K_q$ in a computational sense across the different outputs.
-Recalling our SKI extension to multiple outputs (Sec.~\ref{ski-section}), we build a corresponding approximation for the differentiable part of our kernel:
+To adapt SKI to our context of multi-output GPs, we build a grid $\TU\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with $\TU = [D]\times U$.
+Since the LMC kernel evaluated between two sets of outputs $K_{X_i,X_j}$ is differentiable, as long as $U$ spans a range larger than each $\{X_d\}_{d\in[D]}$, the corresponding SKI approximation (Eq.~\ref{lmcski}) holds with the same asymptotic convergence cubic in $1/m$.
 \begin{align}
@@
 \end{align}
-We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel, so the SKI approximation fails.
-But this fact does not prevent efficient representation or multiplication since the matrix is diagonal.
-Then, the MVM operation $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ require $O(n)$ space and time.
+We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel.
+However, since $\bse$ is diagonal, it is efficient to represent and multiply with.
+Then, the MVM $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ requires $O(n)$ space and time.
 
-We consider different representations of $K_{\TU,\TU}$ from (Eq.~\ref{lmcski}) to reduce the memory and runtime overhead for performing the multiplication $K_{\TU,\TU}\Tz$ in the following sections.
+We consider different representations of $K_{\TU,\TU}$ (Eq.~\ref{lmcski}) to reduce the memory and runtime requirements for performing the multiplication $K_{\TU,\TU}\Tz$ in the following sections.
 
@@
 In \textsc{sum}, we represent $K_{\TU,\TU}$ with a $Q$-length list.
-At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, with only the top row maintained.
+At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, represented using only the top row.
 In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
-As described before, the Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
+The Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
 In turn, assuming $D\ll m$, the runtime for each of the $Q$ terms is $O(Dm\log m)$.
@@
 
-In \textsc{bt}, we notice that $K_{\TU,\TU}$ is a block matrix with blocks $T_{ij}$:
+In \textsc{bt}, we note that $K_{\TU,\TU}$ is a block matrix with blocks $T_{ij}$:
 \begin{align*}
@@
 \end{align*}
-On a one-dimensional grid $U$, these matrices are Toeplitz since they are linear combinations of Toeplitz matrices.
+On a one-dimensional grid $U$, these matrices are Toeplitz because they are linear combinations of Toeplitz matrices.
 \textsc{bt} requires $D^2$ $m$-sized rows to represent each $T_{ij}$.
-Then, using usual block matrix multiplication, an MVM $K_{\TU,\TU}\Tz$ takes $O(D^2m\log m)$ time.
+Then, using the usual block matrix multiplication, an MVM $K_{\TU,\TU}\Tz$ takes $O(D^2m\log m)$ time.
 
-On a grid of inputs with $\TX=\TU$, the SKI interpolation vanishes with $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in a Krylov-subspace based routine has experimentally shown convergence of Krylov-based inversion routines using fewer MVMs \cite{chan1994circulant}.
+On a grid of inputs with $\TX=\TU$, the SKI interpolation becomes $W=I$.
+In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in reduces the number of MVMs necessary to iteratively find an inverse~\cite{chan1994circulant}.
 
@@
 
-For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average added rank, $R\le D$, and re-write the kernel:
+For the rank-based \textsc{slfm} representation, let $R\triangleq \nicefrac{\sum_qR_q}{Q}$ be the average rank, $R\le D$, and re-write the kernel:
 \begin{align*}
@@
 \end{align*}
-Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank-1.
-Under some re-indexing $q'\in[RQ]$ which flattens the double sum such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
+Note $\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top$ is rank $1$.
+Under some re-indexing $q'\in[RQ]$, which flattens the double summation such that each $q'$ corresponds to a unique $(r, q)$, the term $\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q $ may be rewritten as
 \begin{align*}
-  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top;
+  \sum_{q'}\textbf{a}_{q'}\textbf{a}_{q'}^\top\otimes K_{q'} = \textbf{A}\blockdiag_{q'}\colv{K_{q'}}\textbf{A}^\top,
 \end{align*}
@@
 Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-Thus, the \textsc{slfm} representation writes $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix, so MVMs take $O((QR + D)m\log m)$ time.
+Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs take $O((QR + D)m\log m)$ time.
 
-Because the runtimes of \textsc{bt} and \textsc{slfm} are complimentary in the sense that one performs better than the other when $D^2>QR$ and vice-versa, an algorithm that uses the aforementioned condition between to decide between which representation to use can minimize runtime (Tab.~\ref{asymp}).
-We also found that \textsc{sum} is efficient in practice for $Q=1$.
+Note that \textsc{bt} and \textsc{slfm} each have a faster run time than the other depending on whether $D^2>QR$.
+An algorithm that uses this condition to decide between representations can minimize runtime (Tab.~\ref{asymp}).
+We found that \textsc{sum} is efficient in practice for $Q=1$.
 
@@
   \caption{
-   For both LLGP and COGP, $m$ is a configurable parameter which increases up to $n$ to improve accuracy.
-$Q,R,D,\kappa_2$ are coefficients dependent on the setting of the LMC kernel, which has about $QRD$ hyperparameters (Eq.~\ref{lmcpointwise}).
-The resulting asymptotic performance is given in the table.
+   For both LLGP and COGP, $m$ is a configurable parameter that increases up to $n$ to improve accuracy.
+$Q,R,D,\kappa_2$ depend on the LMC kernel, which has $O(QRD)$ hyperparameters (Eq.~\ref{lmcpointwise}).
+The asymptotic performance is given in the table.
 COGP is only independent of $R$ because it cannot represent models for $R\neq 1$.
-We distinguish in the up-front cost and per-hyperparameter cost for computing $\nabla \mcL$ to capture the varying asymptotic behaviors of the gradient algorithms as the number of hyperparemters increases: the total cost would be the up-front cost plus $QRD$ multiplied by the per-hyperparameter cost.
+Computing $\nabla\mcL$ at $\bsth$ requires an up-front cost in addition to the per-hyperparameter cost for each $\theta_j\in\bsth$.
+Multiplicative log terms in $\kappa_2, m$ are hidden.
   }
@@
   \toprule
-  Method & Up-front cost for $\nabla \mcL$ & \begin{tabular}{c}Additional cost for $\partial_{\theta_j}\mcL$\\ per hyperparameter\end{tabular}\\
+  Method & Up-front cost for $\nabla \mcL$ & Additional cost per hyperparameter\\
   \midrule
@@
   COGP & $Qm^3$ & $nm$ \\
-  LLGP & $\min(QR, D^2)\sqrt{\kappa_2} (n + m\log m)$ &  $n + m\log m$ \\
+  LLGP & $\sqrt{\kappa_2} \pa{n +\min(QR+D, D^2) m}$ &  $n + D m$ \\
   \bottomrule
@@
 
-\subsection{Stopping conditions}\label{stopping}
+\subsection{GP mean and variance prediction}
 
-For a gradient-only stopping heuristic, we maintain the running maximum gradient $\infty$-norm.
-If gradient norms drop below a preset proportion of the running maximum norm more than a pre-set tolerance number of times, we terminate.
-For example, when applied to the foreign exchange rate prediction dataset in Sec.~\ref{fx2007-results}, the heuristic eventually notices that we have slowed down increases in $\mcL$ because the gradients occasionally drop below the threshold at that point, while not displacing the solution $\bsth$ significantly since we must be near a local minimum (Fig.~\ref{fx2007-stop}).
-
-\begin{figure}[!h]
-\centering
-{\includegraphics[width=0.5\columnwidth]{running_cutoff.eps}}
-\caption{In green, we have 20\% of the rolling maximum norm. In red and blue are $\mcL$ (computed exactly and therefore unavailable during benchmarks) and $ \norm{\nabla\mcL}_\infty $, respectively. }
-\label{fx2007-stop}
-\end{figure}
-
-\subsection{Prediction}
-
-The predictive mean can be computed in $O(1)$ time as observed in \cite{msgp} by $K_{*,X}\bsa\approx W_{*,\TU}K_{\TU,\TU}\bsa$.
+The predictive mean can be computed in $O(1)$ time by $K_{*,X}\bsa\approx W_{*,\TU}K_{\TU,\TU}\bsa$ \cite{msgp}.
 
@@
 This is done by solving the linear system in a matrix-free manner on-the-fly; in particular, $K_{X,X}^{-1}K_{X,*}$ is computed via \textsc{minres} for every new test point $K_{X,*}$.
-Over several test points, this is embarrassingly parallelizable.
-
-A more efficient predictive variance algorithm is outside the scope of this paper: for a research setting, we expect training time to be the bottleneck.
-Moreover, one can couple LLGP learning with other prediction mechanisms.
-One example is the sampling-based approach proposed in \cite{papandreou2011efficient}, which extends to linear combinations of kernels that allow fast eigendecompositions.
+Over several test points, this process is parallelizable.
 
@@
 We evaluate the methods on held out data by using standardized mean square error (SMSE) of the test points with the predicted mean, and the negative log predictive density (NLPD) of the Gaussian likelihood of the inferred model.
-Notably, NLPD takes confidence into account, while SMSE only evaluates the mean prediction.
+NLPD takes confidence into account, while SMSE only evaluates the mean prediction.
 In both cases, lower values represent better performance.
-We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact log-likelihood gradients using the standard Cholesky algorithm for a grid of different kernels; this evaluation is available in the paper supplement.\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in <anonymous repository>.}
+We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in <anonymous repository>.}
 
@@
 We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance.
-The dataset consists of ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} As in COGP, we retrieved the asset to USD rate, then used its reciprocal in all the results discussed below.
-The LLGP setting has $Q=1,R=2$, as recommended in \cite{alvarez2010efficient} for LMC models on this dataset; let this be the LMC model on LLGP.
+The dataset includes ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} In LLGP, we set $Q=1,R=2$, as recommended for LMC models on this dataset~\cite{alvarez2010efficient}.
 COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP.
-All kernels are RBF.
+All kernels are squared exponential.
 The data used in this example are from 2007, and include $n=3054$ training points and $150$ test points.
@@
 For this application, LLGP uses $m=n/D=238$ interpolating points.
-We use the COGP settings from the paper.\footnote{COGP hyperparameters for FX2007 were 100 inducing points, 500 iterations, 200 mini-batch size.} LLGP, for both LMC, outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).
+We used the COGP settings from the paper.\footnote{COGP hyperparameters for FX2007 were 100 inducing points, 500 iterations, 200 mini-batch size.} LLGP outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).
 
@@
 Next, we replicate results from a weather dataset, a large time series used to validate COGP.
-In this dataset, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure.
+Here, $D=4$ weather sensors Bramblemet, Sotonmet, Cambermet, and Chimet record air temperature over five days in five minute intervals, with some dropped records due to equipment failure.
 Parts of Cambernet and Chimet are dropped for imputation, yielding $n=15789$ training measurements and $374$ test measurements.
 
-We use the COGP parameters that were set by default in the code provided by the authors.\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}.
-The weather dataset was run on 1500 iterations, mini-batch size 1000.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
-We tested LLGP models on different numbers of interpolating points $m$.
+We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}.
+The weather dataset was run on 1500 iterations, mini-batch size $1000$, $200$ interpolating points.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
+We tested LLGP models on different numbers of interpolating points $m = \{500, 1000\}$.
 
 \begin{table}[!h]
-  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the weather dataset.
+  \caption{Average predictive performance and training time over $10$ runs of LLGP and COGP on the weather dataset.
 Parenthesized values are standard error.
@@
 LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tab.~\ref{weather-tbl}).
-Varying the number of interpolating points $m$ from $500$ to $1000$ constructs a tradeoff frontier between increases in $m$ and NLPD decrease at the cost of additional runtime (Fig.~\ref{fig:llgpweather}).
-While NLPD improvement diminishes as $m$ increases, LLGP is still an improvement compared to COGP for a range of $m$ by an order of magnitude in runtime and almost two orders of magnitude for NLPD.
+Varying the number of interpolating points $m$ from $500$ to $1000$ demonstrates the runtime versus NLPD tradeoff (Fig.~\ref{fig:llgpweather}).
+While NLPD improvement diminishes as $m$ increases, LLGP still improves upon COGP for a wide range of $m$ by an order of magnitude in runtime and almost two orders of magnitude in NLPD.
 
-\begin{figure}[!h]
+\begin{wrapfigure}{r}{0.5\textwidth}
+%\begin{figure}[!h]
 \centering
-{\includegraphics[width=0.5\columnwidth]{m_time_nlpd.eps}}
-\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run 10 times.}
+{\includegraphics[width=0.4\columnwidth]{m_time_nlpd.eps}}
+\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run $10$ times.}
 \label{fig:llgpweather}
-\end{figure}
+%\end{figure}
+\end{wrapfigure}
+
 \section{Conclusion}\label{conclusion}
 
-LLGP recovers speedups from SKI \cite{kiss-gp} for the problem of multi-output GP regression by recognizing structure unique to LMC kernels, and otherwise not necessarily recoverable in general multi-output kernels.
-This structure further enables a parsimonious representation that allows even large GPs to be learned without explicit construction of the covariance matrix.
+In this paper, we present LLGP, which we show adapts and accelerates SKI \cite{kiss-gp} for the problem of multi-output GP regression.
+LLGP exploits structure unique to LMC kernels,  enabling a parsimonious representation of the covariance matrix, and gradient computations in $\tilde{O}\pa{\sqrt{\kappa_2}(m+n)}$.
 
-LLGP provides a means to approximate the log-likelihood function gradients through interpolation.
+LLGP provides an efficient means to approximate the log-likelihood gradients using interpolation.
 We have shown on several datasets that this can be done in a way that is faster and leads to more accurate results than variational approximations.
-Because LLGP's representation can be efficient without being sparse (i.e., $m$ may be large), it can can capture complex interactions in the covariance.
-As pictured in Fig.~\ref{fx2007-graph}, and demonstrated by NLPD performance on both the FX2007 and weather datasets (Tab.~\ref{fx2007-tbl},\ref{weather-tbl}), such an efficient, non-sparse representation is integral to making a model whose confidence at a test site reflects the amount of related training data for that site.
+Because LLGP scales well with increases in $m$, capturing complex interactions in the covariance with an accurate interpolation is cheap, as demonstrated by performance on both the FX2007 and weather datasets (Fig.~\ref{fx2007-graph}, Tab.~\ref{fx2007-tbl}, Tab.~\ref{weather-tbl}).
 
@@
 This can be done without losing internal structure in the kernel \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
-The cubic interpolation requires and exponential number of terms, so projection into lower dimensions learned in a supervised manner would be essential.
-Another useful line for investigation would be more informed stopping heuristics.
+The cubic interpolation $W$ requires a number of terms exponential in the dimension, so projection into lower dimensions estimated in a supervised manner would be essential.
+Another useful line for investigation would be more informed stopping heuristics. % cut if space needed
 Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \cite{cutajar2016preconditioning}.

### 2017-05-19 paper/paper.tex
targetted editing

@@
 GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.
-
 Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
@@
 
-% TODO: is this paragraph OK now?
-% yes, except you need to cite more.
 For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \cite{williams1996gaussian}.
@@
 The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure.
-We use this structure to avoid direct matrix inversion.
+Here, we exploit this structure to avoid direct matrix inversion.
 
@@
 A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}.
-With zero mean wlog and a prescribed covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
+With zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
 Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
@@
 
-If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation.
+If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation \cite{gilboa2015scaling, cunningham2008fast}.
 The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$th block $A_{ij}B$.
@@
 $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
-SKI thus creates an $m$-rank approximation to $K_{X,X}$, reducing representation size even without relying on structure.
+Even without relying on structure, SKI reduces the representation of $K_{X,X}$ to an $m$-rank matrix.
 
@@
 We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel.
-This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel even for non-uniform inputs, which we use to accelerate GP estimation in the general case.
+This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel, enabling acceleration of GP estimation for non-uniform inputs.
 
@@
 Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
-Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as the high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
+Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as a gradient-only high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
 
@@
 \begin{align}
-  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}
+  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}.
 \end{align}
-For this approximation, the number of samples may be small, and estimates improve as the size of $K_{|\bsth}$ increases.
-As in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
+This approximation improves as the size of $K_{|\bsth}$ increases, so, as in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
 
@@
 Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs take $O((QR + D)m\log m)$ time.
+Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
 
@@
 The weather dataset was run on 1500 iterations, mini-batch size $1000$, $200$ interpolating points.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
-We tested LLGP models on different numbers of interpolating points $m = \{500, 1000\}$.
+We tested LLGP models on $m=\{500, 1000\}$ interpolating points.
 

### 2017-05-30 paper/copyresults.sh
got rid of files unrelated to paper directly

@@ -1,31 +1,0 @@
-#!/bin/bash
-
-echo 'iterations.eps'
-cp ../benchmarks/picture-fx2007/out/iterations.eps .
-
-echo 'running_cutoff.eps'
-cp ../benchmarks/picture-fx2007/out/running_cutoff.eps .
-
-echo 'relgrad_l2.eps'
-cp ../benchmarks/grad-grid/out/relgrad_l2.eps .
-
-echo 'time_ratio.eps'
-cp ../benchmarks/grad-grid/out/time_ratio.eps .
-
-echo 'relalpha_l2.eps'
-cp ../benchmarks/grad-grid/out/relalpha_l2.eps .
-
-echo 'fx2007graph.pdf'
-cp ../benchmarks/picture-fx2007/out/fx2007graph.pdf .
-
-echo 'm_time_nlpd.eps'
-cp ../benchmarks/weather/out/m_time_nlpd.eps .
-
-echo 'representation-cmp/out/results.tex -> representation.tex'
-cp ../benchmarks/representation-cmp/out/results.tex representation.tex
-
-echo 'results_fx2007.tex'
-cp ../benchmarks/fx2007/out/results_fx2007.tex .
-
-echo 'results_weather.tex'
-cp ../benchmarks/weather/out/results_weather.tex .

### 2017-05-30 paper/paper.tex
Remove NIPS references from arxiv paper
got rid of eps files
missing url

@@
 % Camera ready: \usepackage[final]{nips_2017}
-\usepackage{nips_2017}
+\usepackage[final]{nips_2017}
 
@@
 \begin{center}
-  \subfigure[]{\includegraphics[width=0.4\columnwidth]{iterations.eps}}
-  \subfigure[]{\includegraphics[width=0.4\columnwidth]{running_cutoff.eps}}
+  \subfigure[]{\includegraphics[width=0.4\columnwidth]{iterations.pdf}}
+  \subfigure[]{\includegraphics[width=0.4\columnwidth]{running_cutoff.pdf}}
 \end{center}
@@
 In both cases, lower values represent better performance.
-We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in <anonymous repository>.}
+We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in \url{https://github.com/vlad17/runlmc}.}
 
@@
 \centering
-{\includegraphics[width=0.4\columnwidth]{m_time_nlpd.eps}}
+{\includegraphics[width=0.4\columnwidth]{m_time_nlpd.pdf}}
 \caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run $10$ times.}

### 2017-07-31 paper/paper.tex
Fixed corrections from Prof. Cunningham
1. Section 2.2, O(n^{3/P}) -> O(Pn^{(1+P)/P})
2. Section 4, emphasized how stationarity still plays a role in our work

--- 2017-05-19 paper/paper.tex
@@
 Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
-For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O(n^{3/P})$ \cite{gilboa2015scaling}.
+For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O\pa{Pn^{\nicefrac{1+P}{P}}}$ \cite{gilboa2015scaling}.
 
@@
 
-With multiple outputs, non-differentiable cross-covariance interactions violate the assumptions of SKI's cubic interpolation.
-We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel.
-This approximation induces the previously-identified block-Toeplitz structure of the LMC kernel, enabling acceleration of GP estimation for non-uniform inputs.
+With multiple outputs, LMC cross-covariance interactions violate the assumptions of SKI's cubic interpolation, which require full stationarity (invariance to translation in the indexing space $\mcX'$).
+We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel, which is only invariant to translation along the indexing subspace $\mcX$ (the subkernels $k_q$ are stationary).
+This partial stationarity induces the previously-identified block-Toeplitz structure of the LMC kernel and enables acceleration of GP estimation for non-uniform inputs.
 

### 2017-07-31 paper/paper.tex
Made corrections from NIPS reviewers
1. A couple of mathematical typos fixed
2. consistent derivative notation as \partial
3. No more K_{|theta} to represent the kernel. Only K_{X,X} since reviewer got confused
4. No more bold X to represent inputs across all outputs. Consistent with indexing set mathcal{X} being the continuous space and mathcal{X}'=[D] x mathcal{X}, X' and U' refer to the non-uniform and uniform inputs for the multi-GP across all outputs

@@
 \begin{align}
-\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{|\bsth}\bsa -\frac{1}{2}\Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}};\;\; \bsa=K_{|\bsth}^{-1}\Ty.\label{llgradient}
+\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{X,X}\bsa -\frac{1}{2}\Tr\pa{K_{X,X}^{-1}\partial_{\theta_j}K_{X,X}};\;\; \bsa=K_{X,X}^{-1}\Ty.\label{llgradient}
 \end{align}
@@
 \begin{align*}
-\TX=\set{(d, x)}{d\in[D],x\in X_d}\subset{\mcX'};\;\; n=\card{\TX}.
+X'=\set{(d, x)}{d\in[D],x\in X_d}\subset{\mcX'};\;\; n=\card{X'}.
 \end{align*}
@@
 where $k_q:\R\rightarrow\R$ is a stationary kernel on $\mcX$.
-Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\bsk_q I_D$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a preset rank.
+Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\diag\bsk_q$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a preset rank.
 Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz.
-In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx_i],[j,\Tx_j])\neq K([i+1,\Tx_i+1],[j+1,\Tx_j+1])$.
+In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx_i],[j,\Tx_j])\neq K([i+1,\Tx_i],[j+1,\Tx_j])$.
 
@@
 For example, the deterministic training conditional (DTC) for a single-output GP fixes inducing points $T\subset\mcX$ and estimates kernel hyperparameters for $\Ty_{X}|\Ty_{T}\sim N(K_{X,T}K_{T,T}^{-1}\Ty_T,\sigma^2)$ \cite{quinonero2005unifying}.
-This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $\textbf{T}\subset \mcX'$, with the model equal to the original GP model when $\textbf{T}=\TX$ \cite{alvarez2010efficient}.
+This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $T'\subset \mcX'$, with the model equal to the exact GP model when $T'={X'}$ \cite{alvarez2010efficient}.
 Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
@@
 In Collaborative Multi-output Gaussian Processes (COGP), multi-output GP algorithms further share an internal representation of the covariance structure among all outputs at once~\cite{nguyen2014collaborative}.
-COGP fixes inducing points $\textbf{T}=[D]\times T$ for some $m$-sized $T\subset \mcX$ and puts a GP prior on $\Ty_{\textbf{T}}$ with a restricted LMC kernel that matches the Semiparametric Latent Factor Model (SLFM) \cite{seeger2005semiparametric}.
+COGP fixes inducing points ${T'}=[D]\times T$ for some $m$-sized $T\subset \mcX$ and puts a GP prior on $\Ty_{{T'}}$ with a restricted LMC kernel that matches the Semiparametric Latent Factor Model (SLFM) \cite{seeger2005semiparametric}.
 Applying the COGP prior to $\Ty_X$ corresponds to an LMC kernel (Eq.~\ref{lmcpointwise}) where $\bsk_q$ is set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
@@
 
-\subsection{Approaches for stationary kernels}
-
-\subsubsection{Structured Kernel Interpolation (SKI)}\label{ski-section}
+\subsection{Structured Kernel Interpolation (SKI)}\label{ski-section}
 
@@
 For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$.
-The simultaneous interpolation $W\triangleq W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx WK_{U,U}W^\top$.
+The simultaneous interpolation $W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx W_{X,U}K_{U,U}W_{U,X}^\top$.
 $W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
@@
 Massively Scalable Gaussian Processes (MSGP) exploits structure as well: the kernel $K_{U,U}$ on a grid has Kronecker and Toeplitz matrix structure \cite{msgp}.
-Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{|\bsth}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
-In addition, an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{|\bsth}}$ term in Eq.~\ref{llgradient} has been noted previously \cite{wilson2014fast}.
+Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{X,X}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
+In addition, an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{X,X}}$ term in Eq.~\ref{llgradient} has been noted previously \cite{wilson2014fast}.
 
-Although evaluating $\log\abs{K_{|\bsth}}$ is not feasible in the LMC setting because the LMC sum breaks Kronecker and Toeplitz structure, the approach of creating structure with SKI carries over to LLGP.
+Although evaluating $\log\abs{K_{X',X'}}$ is not feasible in the LMC setting because the LMC sum breaks Kronecker and Toeplitz structure, the approach of creating structure with SKI carries over to LLGP.
 
@@
 
-With multiple outputs, LMC cross-covariance interactions violate the assumptions of SKI's cubic interpolation, which require full stationarity (invariance to translation in the indexing space $\mcX'$).
+With multiple outputs, LMC cross-covariance interactions violate the assumptions of SKI's cubic interpolation, which require full stationarity (invariance to translation in the indexing space $\mcX'$) and differentiability.
 We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel, which is only invariant to translation along the indexing subspace $\mcX$ (the subkernels $k_q$ are stationary).
@@
 
-For a given $\bsth$, we construct an operator $\tilde{K}_{|\bsth}$ which approximates MVMs with the covariance matrix, $K_{|\bsth}\Tz\approx \tilde{K}_{|\bsth}\Tz$.
+For a given $\bsth$, we construct an operator $\tilde{K}$ which approximates MVMs with the exact covariance matrix evaluated, which for brevity we overload as $K\triangleq K_{X',X'}$, so $K\Tz\approx \tilde{K}\Tz$.
 Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
@@
 \citet{gibbs1996cient} describe the algorithm for GP model estimation in terms of only MVMs with the covariance matrix.
-In particular, they note that we can solve for $\bsa$ satisfying $K_{|\bsth}\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG).
-Moreover, they develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
+In particular, we can solve for $\bsa$ satisfying $K\bsa=\Ty$ in Eq.~\ref{llgradient} using linear conjugate gradient descent (LCG).
+Moreover, Gibbs and MacKay develop a stochastic approximation by introducing RV $\tbr$ with $\cov \tbr=I$:
 \begin{align}
-  \Tr\pa{K_{|\bsth}^{-1}\partial_{\theta_j}K_{|\bsth}} = \E\left[(K_{|\bsth}^{-1}\tbr)^\top\partial_{\theta_j}K_{|\bsth}\tbr\right]\label{eq:trace}.
+  \Tr\pa{K^{-1}\partial_{\theta_j}K} = \E\left[(K^{-1}\tbr)^\top\partial_{\theta_j}K\tbr\right]\label{eq:trace}.
 \end{align}
-This approximation improves as the size of $K_{|\bsth}$ increases, so, as in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
+This approximation improves as the size of $K$ increases, so, as in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
 
 We depart from Gibbs and MacKay in two important ways (Algorithm~\ref{alg:grad}).
-First, we do not construct $K_{|\bsth}$, but instead keep a low-memory representation $\tilde{K}_{|\bsth}$ (Sec.~\ref{fast-mvm}).
+First, we do not construct $K$, but instead keep a low-memory representation $\tilde{K}$ (Sec.~\ref{fast-mvm}).
 Second, we use \textsc{minres} instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
@@
 This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks throughout learning.
-Inversion-based methods then require additional iterations because $\kappa_2$, the spectral condition number of $K_{|\bsth}$, increases as $K_{|\bsth}$ becomes less diagonally dominant (Fig.~\ref{fx2007-iterations}).
+Inversion-based methods then require additional iterations because $\kappa_2$, the spectral condition number of $K$, increases as $K$ becomes less diagonally dominant (Fig.~\ref{fx2007-iterations}).
 
@@
 \end{center}
-\caption{Trace of (a) the number of MVMs that \textsc{minres} must perform to invert $K_{|\bsth}^{-1}\Ty$ and (b) $\mcL, \norm{\nabla\mcL}$ given $\bsth$ at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
+\caption{Trace of (a) the number of MVMs that \textsc{minres} must perform to invert $K^{-1}\Ty$ and (b) $\mcL, \norm{\nabla\mcL}$ given $\bsth$ at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
 In (b), in green, we have 20\% of the rolling maximum $\infty$-norm of previous gradients.
@@
 
-Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K}_{|\bsth})\sqrt{\kappa_2})$ \cite{raykar2007fast}.
-This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K_{|\bsth}$ for each $j$ at least as fast as $\mvm({\tilde{K}_{|\bsth}})$.
-Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K_{|\bsth}$.
+Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K})\sqrt{\kappa_2})$ \cite{raykar2007fast}.
+This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K$ for each $j$ at least as fast as $\mvm({\tilde{K}})$.
+Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K$.
 
@@
 Assume \textsc{minres} is the inversion routine.
-We also assume we have access to linear operators $D_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}_{|\bsth}$.} \label{alg:grad}
+We also assume we have access to linear operators $\partial_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}$.} \label{alg:grad}
 \begin{algorithmic}[1]
-  \Procedure{LLGP}{$\tilde{K}_{|\bsth}$, $\Ty$, $N$, $\{D_{\theta_j}\}$}
+  \Procedure{LLGP}{$\tilde{K}$, $\Ty$, $N$, $\{\partial_{\theta_j}\}$}
   \State $R\gets\idx{\tbr_i}{i=1}{N}$, sampling $\tbr\sim\Unif\{\pm 1\}$.
 \For{$\Tz$ in $\{\Ty\}\cup R$, in parallel}
-\State $K^{-1}\Tz\gets\textsc{minres}(\tilde{K}_{|\bsth},\Tz)$.
+\State $K^{-1}\Tz\gets\textsc{minres}(\tilde{K},\Tz)$.
 \EndFor
@@
 \For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
-\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot D_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
-\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}_{|\bsth}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
+\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot \partial_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
+\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
 \EndFor
@@
 The bottleneck of Algorithm~\ref{alg:grad} is the iterative MVM operations in $\textsc{minres}$.
-Since $K_{|\bsth}$ only enters computation as an MVM operator, the required memory is dictated by its representation $\tilde{K}_{|\bsth}$, which need not be dense as long as we can perform MVM with any vector to arbitrary, fixed precision.
+Since $K$ only enters computation as an MVM operator, the required memory is dictated by its representation $\tilde{K}$, which need not be dense as long as we can perform MVM with any vector to arbitrary, fixed precision.
 
-When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise over $\TU$  (Eq.~\ref{lmcgrid}) holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated the shared interpolating points $U$ for all outputs:
+When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise over ${U'}$  (Eq.~\ref{lmcgrid}) holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated the shared interpolating points $U$ for all outputs:
 \begin{align}
-  K_{\TU,\TU}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}.
+  K_{{U'},{U'}}=\sum_q(A_qA_q^\top+\diag\bsk_q)\otimes K_q\label{lmcgrid}.
 \end{align}
-To adapt SKI to our context of multi-output GPs, we build a grid $\TU\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with $\TU = [D]\times U$.
+To adapt SKI to our context of multi-output GPs, we build a grid ${U'}\subset\mcX'$ out of a common subgrid $U\subset\mcX$ that extends to all outputs with ${U'} = [D]\times U$.
 Since the LMC kernel evaluated between two sets of outputs $K_{X_i,X_j}$ is differentiable, as long as $U$ spans a range larger than each $\{X_d\}_{d\in[D]}$, the corresponding SKI approximation (Eq.~\ref{lmcski}) holds with the same asymptotic convergence cubic in $1/m$.
 \begin{align}
-  K_{\TX,\TX}\approx WK_{\TU,\TU}W^\top+\bse.\label{lmcski}
+  K\approx \tilde{K}=WK_{{U'},{U'}}W^\top+\bse.\label{lmcski}
 \end{align}
-We cannot fold $\bse$ into the interpolated term $K_{\TU,\TU}$ since it does not correspond to a differentiable kernel.
+We cannot fold $\bse$ into the interpolated term $K_{{U'},{U'}}$ since it does not correspond to a differentiable kernel.
 However, since $\bse$ is diagonal, it is efficient to represent and multiply with.
-Then, the MVM $K_{\TX,\TX}\Tz$ can be approximated by $WK_{\TU,\TU}W^\top\Tz+\bse\Tz$, where matrix multiplication by the sparse matrices $\bse, W,W^\top$ requires $O(n)$ space and time.
+MVM with $\tilde{K}$ requires MVM by the sparse matrices $\bse, W,W^\top$, which all take $O(n)$ space and time.
 
-We consider different representations of $K_{\TU,\TU}$ (Eq.~\ref{lmcski}) to reduce the memory and runtime requirements for performing the multiplication $K_{\TU,\TU}\Tz$ in the following sections.
+We consider different representations of $K_{{U'},{U'}}$ (Eq.~\ref{lmcski}) to reduce the memory and runtime requirements for performing the multiplication $K_{{U'},{U'}}\Tz$ in the following sections.
 
@@
 
-In \textsc{sum}, we represent $K_{\TU,\TU}$ with a $Q$-length list.
+In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
 At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, represented using only the top row.
-In turn, multiplication $K_{\TU,\TU}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
+In turn, multiplication $K_{{U'},{U'}}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
 The Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
@@
 
-In \textsc{bt}, we note that $K_{\TU,\TU}$ is a block matrix with blocks $T_{ij}$:
+In \textsc{bt}, we note that $K_{{U'},{U'}}$ is a block matrix with blocks $T_{ij}$:
 \begin{align*}
@@
 \textsc{bt} requires $D^2$ $m$-sized rows to represent each $T_{ij}$.
-Then, using the usual block matrix multiplication, an MVM $K_{\TU,\TU}\Tz$ takes $O(D^2m\log m)$ time.
+Then, using the usual block matrix multiplication, an MVM $K_{{U'},{U'}}\Tz$ takes $O(D^2m\log m)$ time.
 
-On a grid of inputs with $\TX=\TU$, the SKI interpolation becomes $W=I$.
+On a grid of inputs with ${X'}={U'}$, the SKI interpolation becomes $W=I$.
 In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in reduces the number of MVMs necessary to iteratively find an inverse~\cite{chan1994circulant}.
@@
 \begin{align*}
-  K_{\TU,\TU}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
+  K_{{U'},{U'}}=\sum_q\sum_{r=1}^{R_q}\textbf{a}_q^{(r)}{\textbf{a}_q^{(r)}}^\top\otimes K_q + \sum_q\diag\bsk_q \otimes K_q.
 \end{align*}
@@
 Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-Thus, \textsc{slfm} represents $K_{\TU,\TU}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
+Thus, \textsc{slfm} represents $K_{{U'},{U'}}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
 
@@
 
-The predictive mean can be computed in $O(1)$ time by $K_{*,X}\bsa\approx W_{*,\TU}K_{\TU,\TU}\bsa$ \cite{msgp}.
+The predictive mean can be computed in $O(1)$ time by $K_{*,X}\bsa\approx W_{*,{U'}}K_{{U'},{U'}}\bsa$ \cite{msgp}.
 
@@
 Future work would extend the inputs to accept multiple dimensions.
-This can be done without losing internal structure in the kernel \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
+This can be done without losing internal structure in the kernel, again so long as only $k_q$ are stationary \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
 The cubic interpolation $W$ requires a number of terms exponential in the dimension, so projection into lower dimensions estimated in a supervised manner would be essential.

### 2017-08-01 paper/paper.tex
Made paper arxiv-compatible

@@
 % Camera ready: \usepackage[final]{nips_2017}
-\usepackage{nips_2017}
+% currently, doesn't print NIPS
+\usepackage[final]{nips_2017}
 
@@
 The contribution of the $q$th kernel $k_q$ to the covariance between the $i$th and $j$th outputs is then specified by the multiplicative factor $b_{ij}^{(q)}$.
-By choosing $B_q$ to have rank $R_q=D$, the corresponding LMC model can have any positive contribution between two outputs that best fits the data, so long as $B_q$ remains PSD.
+By choosing $B_q$ to have rank $R_q=D$, the corresponding LMC model can have any contribution between two outputs that best fits the data, so long as $B_q$ remains PSD.
 By reducing the rank $R_q$, the interactions of the outputs have lower-rank latent processes, with rank 0 indicating no interaction (i.e., if $A=0$, then we have an independent GP for each output).
@@
 
-Symmetric Toeplitz matrices $T$ are constant along their diagonal and fully determined by their top row $\{T_{1j}\}_{j=1}^n$, yielding an $O(n)$ representation.
+Symmetric Toeplitz matrices $T$ are constant along their diagonals and fully determined by their top row $\{T_{1j}\}_{j=1}^n$, yielding an $O(n)$ representation.
 Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs.
@@
 This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K$ for each $j$ at least as fast as $\mvm({\tilde{K}})$.
-Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $K$.
+Indeed, we assume a differentiable kernel and then recall that applying the linear operator $\partial_{\theta_j}$ will maintain the structure of $\tilde{K}$.
 
@@
 In both cases, lower values represent better performance.
-We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in <anonymous repository>.}
-
+We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in
+\url{https://github.com/vlad17/runlmc}.} % camera ready
+%\texttt{<anonymous repository>}.}
 \subsection{Foreign exchange rate prediction (FX2007)}\label{fx2007-results}
@@
 For this application, LLGP uses $m=n/D=238$ interpolating points.
-We used the COGP settings from the paper.\footnote{COGP hyperparameters for FX2007 were 100 inducing points, 500 iterations, 200 mini-batch size.} LLGP outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).
+We used the COGP settings from the paper.
+LLGP outperforms COGP in terms of predictive mean and variance estimation as well as runtime (Tab.~\ref{fx2007-tbl}).
 
@@
 
-We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11838e89700cfb58d26ca39b119a35}.
-The weather dataset was run on 1500 iterations, mini-batch size $1000$, $200$ interpolating points.} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
+We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11}} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
 We tested LLGP models on $m=\{500, 1000\}$ interpolating points.

### 2017-08-01 arxiv-tar.sh

@@ -1,0 +1,3 @@
+#!/bin/bash
+
+tar --exclude="paper/*.eps" --exclude="paper/.gitignore" --exclude="*.sh" --exclude="paper/*.out" --exclude="paper/supplement*" --exclude="paper/paper.pdf" -czvf paper.tgz paper/

### 2017-08-01 paper/copyresults.sh

@@ -1,25 +1,38 @@
 #!/bin/bash
 
-echo 'iterations.eps'
+function epsconvert {
+    mv $1.eps $1-unbounded.eps
+    epstool --quiet --copy --bbox $1-unbounded.eps $1.eps
+    epstopdf --quiet --hires $1.eps
+    rm $1-unbounded.eps
+}
+
+echo 'iterations.eps -> iterations.pdf'
 cp ../benchmarks/picture-fx2007/out/iterations.eps .
+epsconvert iterations
 
-echo 'running_cutoff.eps'
+echo 'running_cutoff.eps -> running_cutoff.pdf'
 cp ../benchmarks/picture-fx2007/out/running_cutoff.eps .
+epsconvert running_cutoff
 
-echo 'relgrad_l2.eps'
+echo 'relgrad_l2.eps -> relgrad_l2.pdf'
 cp ../benchmarks/grad-grid/out/relgrad_l2.eps .
+epsconvert relgrad_l2
 
-echo 'time_ratio.eps'
+echo 'time_ratio.eps -> time_ratio.pdf'
 cp ../benchmarks/grad-grid/out/time_ratio.eps .
+epsconvert time_ratio
 
-echo 'relalpha_l2.eps'
+echo 'relalpha_l2.eps -> relalpha_l2.pdf'
 cp ../benchmarks/grad-grid/out/relalpha_l2.eps .
+epsconvert relalpha_l2
 
 echo 'fx2007graph.pdf'
 cp ../benchmarks/picture-fx2007/out/fx2007graph.pdf .
 
-echo 'm_time_nlpd.eps'
+echo 'm_time_nlpd.eps -> m_time_nlpd.pdf'
 cp ../benchmarks/weather/out/m_time_nlpd.eps .
+epsconvert m_time_nlpd
 
 echo 'representation-cmp/out/results.tex -> representation.tex'
 cp ../benchmarks/representation-cmp/out/results.tex representation.tex

### 2017-09-13 benchmarks/benchlib/bench.py
fix benchmark bug with recent API change

@@ -19,7 +19,7 @@ from runlmc.kern.matern32 import Matern32
 from runlmc.kern.std_periodic import StdPeriodic
 from runlmc.models.lmc import LMC
 from runlmc.lmc.stochastic_deriv import StochasticDeriv
-from runlmc.lm.functional_kernel import FunctionalKernel
+from runlmc.lmc.functional_kernel import FunctionalKernel
 from runlmc.lmc.grid_kernel import *
 from runlmc.lmc.kernel import ExactLMCKernel, ApproxLMCKernel
 

### 2017-09-13 benchmarks/benchlib/standard_tester.py
updated standard tester to fix new API bug

@@ -227,18 +227,8 @@ def bench_runlmc(num_runs, m, xss, yss, test_xss, test_yss,
             lmc_ranks=rgen(),
             slfm_kernels=slfmgen(),
             indep_gp=indepgen())
-        lmc = LMC(xss, yss, kernels=ks, ranks=rs,
-                  slfm_kernels=slfm, indep_gp=indep,
+        lmc = LMC(xss, yss, functional_kernel=fk,
                   normalize=True, m=m, **kwargs)
-        for i in range(lmc.nkernels['lmc']):
-            print('LMC kernel', i, 'A matrix')
-            print(eval('lmc.a{}'.format(i)).values)
-            print('LMC kernel', i, 'kappa diag')
-            print(eval('lmc.kappa{}'.format(i)).values)
-        for i in range(lmc.nkernels['slfm']):
-            i += lmc.nkernels['lmc']
-            print('SLFM kernel', i, 'A matrix')
-            print(eval('lmc.a{}'.format(i)).values)
         opt = AdaDelta(**optimizer_opts)
         with contexttimer.Timer() as t:
             lmc.optimize(optimizer=opt)

### 2017-09-13 runlmc/models/lmc.py

@@ -355,7 +355,7 @@ class LMC(MultiGP):
             return self._var_predict_precompute(W, Xs)
 
         tot_pred_size = sum(map(len, Xs))
-        tot_grid_size = len(self.inducing_grid) * len(self.noise)
+        tot_grid_size = len(self.inducing_grid) * len(self._functional_kernel.noise)
         if tot_pred_size > tot_grid_size:
             return self._var_predict_precompute(W, Xs)
         return self._var_predict_on_the_fly(W, Xs)
@@ -453,7 +453,7 @@ class LMC(MultiGP):
         K_UX = Composition([K_UU, Matrix.wrap(WT.shape, WT.dot)])
 
         m = len(self.inducing_grid)
-        D = len(self.noise)
+        D = len(self._functional_kernel.noise)
         Dm = D * m
         assert Dm == K_XU.shape[1] and Dm == K_UX.shape[0]
 

### 2017-09-13 benchmarks/benchlib/standard_tester.py
updated standard tester to fix new API bug

@@ -227,8 +227,7 @@ def bench_runlmc(num_runs, m, xss, yss, test_xss, test_yss,
             lmc_ranks=rgen(),
             slfm_kernels=slfmgen(),
             indep_gp=indepgen())
-        lmc = LMC(xss, yss, kernels=ks, ranks=rs,
-                  slfm_kernels=slfm, indep_gp=indep,
+        lmc = LMC(xss, yss, functional_kernel=fk,
                   normalize=True, m=m, **kwargs)
         for i in range(lmc.nkernels['lmc']):
             print('LMC kernel', i, 'A matrix')

### 2017-09-13 runlmc/models/lmc.py
fixed line being too long

@@ -355,7 +355,8 @@ class LMC(MultiGP):
             return self._var_predict_precompute(W, Xs)
 
         tot_pred_size = sum(map(len, Xs))
-        tot_grid_size = len(self.inducing_grid) * len(self._functional_kernel.noise)
+        tot_grid_size = len(self.inducing_grid) * len(
+            self._functional_kernel.noise)
         if tot_pred_size > tot_grid_size:
             return self._var_predict_precompute(W, Xs)
         return self._var_predict_on_the_fly(W, Xs)

### 2017-10-04 runlmc/lmc/functional_kernel.py
Added testing (and bugfixes) for split kernels

@@ -235,8 +235,8 @@ class FunctionalKernel(Parameterized):
         """Computes the array of k_q applied to each distance in `dists`, where
         `dists` should be a dict of `active_dim`-keyed distances."""
         assert self.P
-        return np.array([k.from_dist(dists[k.active_dims])
-                         for k in self._kernels])
+        return [k.from_dist(dists[k.active_dims])
+                for k in self._kernels]
 
     def eval_kernels_fixed_dim(self, dists, active_dim):
         """Computes the array of k_q applied to each distance in `dists`,
@@ -291,4 +291,12 @@ class FunctionalKernel(Parameterized):
         return len(self._kernels)
 
     def get_active_dims(self, q):
+        """Returns the active dimensions for the kernel with index `q`"""
         return self._kernels[q].active_dims
+
+    def filter_non_indep_idxs(self, idxs):
+        """Return only the kernel indices associated with coregionalized
+        (non-independent) kernels"""
+        non_indep_max = self._num_lmc + self._num_slfm
+        non_indep = [kidx for kidx in idxs if kidx < non_indep_max]
+        return non_indep

### 2017-10-04 runlmc/lmc/grid_kernel.py

@@ -32,8 +32,9 @@ class GridKernel(Matrix):
         elif ktype == 'bt':
             self.grid_K = _gen_bt_grid(functional_kernel, grid_k, active_dim)
         elif ktype == 'slfm':
+            m = interpolant.shape[1]
             self.grid_K = _gen_slfm_grid(
-                functional_kernel, grid_k, n, active_dim)
+                functional_kernel, grid_k, m, active_dim)
         else:
             assert False, ktype
 
@@ -73,17 +74,18 @@ def gen_grid_kernel(fk, grid_dists, interpolants, lens_per_output):
     return SumMatrix(ls), grid_kerns
 
 
-def _gen_slfm_grid(functional_kernel, grid_k, n, active_dim):
-    coreg_Ks = _gen_coreg_Ks(functional_kernel, grid_k, active_dim)
-    diag_Ks = _gen_diag_Ks(functional_kernel, grid_k, n, active_dim)
+def _gen_slfm_grid(functional_kernel, grid_k, m, active_dim):
+    coreg_Ks = _gen_coreg_Ks(functional_kernel, grid_k, m, active_dim)
+    diag_Ks = _gen_diag_Ks(functional_kernel, grid_k, m, active_dim)
     return SumMatrix([coreg_Ks, diag_Ks])
 
 
-def _gen_coreg_Ks(fk, grid_k, active_dim):
+def _gen_coreg_Ks(fk, grid_k, m, active_dim):
     kidxs = fk.active_dims[active_dim]
-    non_indep_max = fk.num_lmc[active_dim] + fk.num_slfm[active_dim]
-    non_indep = [kidx for kidx in kidxs if kidx < non_indep_max]
+    non_indep = fk.filter_non_indep_idxs(kidxs)
     all_coreg = [fk.coreg_vecs[idx] for idx in non_indep]
+    if not all_coreg:
+        return Identity(m)
     ranks = np.array([len(coreg) for coreg in all_coreg])
     A_star = np.vstack(all_coreg).T
     I_m = Identity(np.prod(grid_k.shape[1:]))
@@ -98,9 +100,9 @@ def _gen_coreg_Ks(fk, grid_k, active_dim):
     return coreg_Ks
 
 
-def _gen_diag_Ks(fk, grid_k, n, active_dim):
+def _gen_diag_Ks(fk, grid_k, m, active_dim):
     if fk.num_lmc[active_dim] == 0 and fk.num_indep[active_dim] == 0:
[…]

### 2017-10-04 runlmc/lmc/likelihood.py

@@ -146,7 +146,6 @@ class ExactLMCLikelihood(LMCLikelihood):
         self.K = sum(self._personalized_coreg_scale(A, Kq) for A, Kq in
                      zip(self.functional_kernel.coreg_mats(),
                          self.materialized_kernels))
-
         self.K += np.diag(np.repeat(functional_kernel.noise, self.lens))
         self.L = la.cho_factor(self.K)
         self.deriv = ExactDeriv(self.L, self.y)

### 2017-10-04 runlmc/models/interpolated_llgp.py

@@ -332,7 +332,8 @@ class InterpolatedLLGP(MultiGP):
         mean = np.zeros(sum(lens))
         prediction_W = {}
         for active_dim, grid_alpha_ad in grid_alpha.items():
-            W = multi_interpolant(Xs, *self.grid_axes[active_dim])
+            Xs_active = [X[:, active_dim] for X in Xs]
+            W = multi_interpolant(Xs_active, *self.grid_axes[active_dim])
             prediction_W[active_dim] = W
             mean += W.dot(grid_alpha_ad)
 

### 2017-10-04 runlmc/models/test_interpolated_llgp.py

@@ -2,6 +2,7 @@
 # Licensed under the BSD 3-clause license (see LICENSE)
 
 import itertools
+from functools import partial
 
 import numpy as np
 from parameterized import parameterized
@@ -101,6 +102,15 @@ class LMCTestUtils:
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
 
     @staticmethod
+    def _case_2d_splitkern(split1, split2, input_dim):
+        assert input_dim >= 2
+        kernels = [RBF(inv_lengthscale=3, active_dims=split1),
+                   RBF(inv_lengthscale=2, active_dims=split2)]
+        szs = [30, 40]
+        coregs = [np.array(x).reshape(1, -1) for x in [[1, 2], [3, 4]]]
+        return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
+
+    @staticmethod
     def _case_multirank(input_dim):
         kernels = [RBF(inv_lengthscale=3),
                    RBF(inv_lengthscale=2)]
@@ -135,7 +145,17 @@ class LMCTestUtils:
         return {
             'input_1dimplicit': None,
             'input_1d': 1,
-            'input_2d': 2}
+            'input_2d': 2,
+            'input_3d': 3}
+
+    @classmethod
+    def _input_splits(cls, in_dim):
+        if in_dim is None or in_dim < 2:
+            return []
+        elif in_dim == 2:
+            return [((0, 1), (0,)), ((0,), (1,))]
+        assert in_dim == 3, in_dim
+        return [((0, 1), (2,)), ((0, 1), (1, 2)), ((0,), (1,)), ((0, 1), (0,))]
 
     @classmethod
     def input_output_cases_grid(cls):
@@ -143,7 +163,17 @@ class LMCTestUtils:
         ins = cls._input_cases()
         for (out_name, out), (in_name, in_dim) in itertools.product(
                 outs.items(), ins.items()):
-            yield out_name + '_' + in_name, out, in_dim
+            # Non-split kernels: no more than two input dims
+            if in_dim is None or in_dim <= 2:
+                yield out_name + '_' + in_name, out, in_dim
+        for in_name, in_dim in ins.items():
+            for splits in cls._input_splits(in_dim):
+                name = 'output_2d_input_2d_splitkern_'
+                name += '_'.join(map(str, splits[0]))
+                name += '___'
+                name += '_'.join(map(str, splits[1]))
[…]

### 2017-10-05 paper/paper.tex
move to AISTATS

@@
-\documentclass{article}
+\documentclass[twoside]{article}
+\usepackage{aistats2018}
+% Camera ready: \usepackage[accepted]{aistats2018}
 
-\PassOptionsToPackage{numbers, compress}{natbib}
+\usepackage[round]{natbib}
 % Camera ready: \usepackage[final]{nips_2017}
 % currently, doesn't print NIPS
-\usepackage[final]{nips_2017}
 
@@
 \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}
 \usepackage{hyperref}       % hyperlinks
@@
 
-\title{Large Linear Multi-output Gaussian Process Learning for Time Series}
+\begin{document}
+\twocolumn[
 
-\author{
-  Vladimir Feinberg, Li-Fang Cheng,  {Kai Li}, {Barbara E Engelhardt}\\
-  Department of Computer Science\\
-  Princeton University\\
-  Princeton, NJ 08540 \\
-  \texttt{\{vyf,lifangc,li,bee\}@princeton.edu} \\
-}
+\aistatstitle{Large Linear Multi-output Gaussian Process Learning}
 
-\begin{document}
+\aistatsauthor{   Vladimir Feinberg \And Li-Fang Cheng \And  {Kai Li} \And {Barbara E Engelhardt} }
 
-\maketitle
+\aistatsaddress{ UC Berkeley \And Princeton University \And Princeton University \And Princeton University }
+]
 
 \begin{abstract}
-Gaussian processes, or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \cite{alvarez2012kernels}.
+Gaussian processes, or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \citep{alvarez2012kernels}.
 LMCs estimate and exploit correlations across the multiple outputs.
-While model estimation can be performed efficiently for single-output GPs \cite{msgp}, these assume stationarity, but in the multi-output case the cross-covariance interaction is not stationary.
+While model estimation can be performed efficiently for single-output GPs \citep{msgp}, these assume stationarity, but in the multi-output case the cross-covariance interaction is not stationary.
 We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and one-dimensional inputs.
@@
 % 8 pages of content
-% \citet for inline citations
 % footnotes after punctuation
@@
 
-Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \cite{williams1996gaussian}.
+Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \citep{williams1996gaussian}.
 GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.
-Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \cite{osborne2008towards, alvarez2010efficient}.
+Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \citep{osborne2008towards, alvarez2010efficient}.
 Efficient model estimation would enable researchers to quickly explore large spaces of parameterizations to find an appropriate one for their task.
 
-For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \cite{williams1996gaussian}.
-Some single-output GP methods exploit structure in this matrix to reduce runtime \cite{msgp}.
+For $n$ input points, exact GP inference requires maintaining an $n^2$  matrix of covariances between response variables at each input and performing $O(n^3)$ inversions with that matrix \citep{williams1996gaussian}.
+Some single-output GP methods exploit structure in this matrix to reduce runtime \citep{msgp}.
 In the multi-output setting, the same structure does not exist.
-Approximations developed for multi-output methods instead reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \cite{nguyen2014collaborative}.
+Approximations developed for multi-output methods instead reduce the dimensionality of the GP estimation problem from $n$ to $m<n$, but still require $m$ to scale with $n$ to retain accuracy \citep{nguyen2014collaborative}.
 The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure.
@@
 
-A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \cite{williams1996gaussian}.
+A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \citep{williams1996gaussian}.
 With zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
-Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \cite{williams1996gaussian}.
+Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \citep{williams1996gaussian}.
 Predictive accuracy is sensitive to a particular parameterization of our kernel, and model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$.
 Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
+
 \begin{align}
@@
 
-We build multi-output GP models as instances of general GPs, where a multi-output model explicitly represents correlations between outputs through a shared input space \cite{alvarez2012kernels}.
+We build multi-output GP models as instances of general GPs, where a multi-output model explicitly represents correlations between outputs through a shared input space \citep{alvarez2012kernels}.
 Here, for $D$ outputs, we write our indexing set as $\mcX'=[D]\times \mcX$, a point from a shared domain coupled with an output index.
@@
 
-If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation \cite{gilboa2015scaling, cunningham2008fast}.
+If we can identify structure in the covariance $K$, then we can develop fast in-memory representations and efficient matrix-vector multiplications (MVMs) for $K$---this has been used in the past to accelerate GP model estimation \citep{gilboa2015scaling, cunningham2008fast}.
 The Kronecker product $A\otimes B$ of matrices of order $a,b$ is a block matrix of order $ab$ with $ij$th block $A_{ij}B$.
@@
 Then, the corresponding MVMs can be computed in time $O(a\mvm(B)+b\mvm(A))$, where $\text{MVM}(\cdot)$ is the runtime of a MVM.
-For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O\pa{Pn^{\nicefrac{1+P}{P}}}$ \cite{gilboa2015scaling}.
+For GPs on uniform dimension-$P$ grids, this reduces the runtime of finding $\mcL$ from $O(n^3)$ to $O\pa{Pn^{\nicefrac{1+P}{P}}}$ \citep{gilboa2015scaling}.
 
@@
 Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs $C\mat{\Tx & \textbf{0}}^\top=\mat{T\Tx &\textbf{0}}^\top$ in $O(n\log n)$ time.
-This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \cite{cunningham2008fast}.
+This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \citep{cunningham2008fast}.
 
@@
 Inducing point approaches create a tractable model to approximate the exact GP.
-For example, the deterministic training conditional (DTC) for a single-output GP fixes inducing points $T\subset\mcX$ and estimates kernel hyperparameters for $\Ty_{X}|\Ty_{T}\sim N(K_{X,T}K_{T,T}^{-1}\Ty_T,\sigma^2)$ \cite{quinonero2005unifying}.
-This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $T'\subset \mcX'$, with the model equal to the exact GP model when $T'={X'}$ \cite{alvarez2010efficient}.
+For example, the deterministic training conditional (DTC) for a single-output GP fixes inducing points $T\subset\mcX$ and estimates kernel hyperparameters for $\Ty_{X}|\Ty_{T}\sim N(K_{X,T}K_{T,T}^{-1}\Ty_T,\sigma^2)$ \citep{quinonero2005unifying}.
+This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $T'\subset \mcX'$, with the model equal to the exact GP model when $T'={X'}$ \citep{alvarez2010efficient}.
 Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
 
-In Collaborative Multi-output Gaussian Processes (COGP), multi-output GP algorithms further share an internal representation of the covariance structure among all outputs at once~\cite{nguyen2014collaborative}.
-COGP fixes inducing points ${T'}=[D]\times T$ for some $m$-sized $T\subset \mcX$ and puts a GP prior on $\Ty_{{T'}}$ with a restricted LMC kernel that matches the Semiparametric Latent Factor Model (SLFM) \cite{seeger2005semiparametric}.
+In Collaborative Multi-output Gaussian Processes (COGP), multi-output GP algorithms further share an internal representation of the covariance structure among all outputs at once~\citep{nguyen2014collaborative}.
+COGP fixes inducing points ${T'}=[D]\times T$ for some $m$-sized $T\subset \mcX$ and puts a GP prior on $\Ty_{{T'}}$ with a restricted LMC kernel that matches the Semiparametric Latent Factor Model (SLFM) \citep{seeger2005semiparametric}.
 Applying the COGP prior to $\Ty_X$ corresponds to an LMC kernel (Eq.~\ref{lmcpointwise}) where $\bsk_q$ is set to 0 and $A_q=\textbf{a}_q\in\R^{D\times 1}$.
@@
 
-SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \cite{kiss-gp}.
+SKI abandons the inducing-point approach: instead of using an intrinsically sparse model, SKI approximates the original $K_{X,X}$ directly \citep{kiss-gp}.
 To do this efficiently, SKI relies on the differentiability of $K$.
-For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\cite{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$.
+For $\Tx,\Tz$ within a grid $U$, $\card{U}=m$, and $W_{\Tx,U}\in\R^{1\times m}$ as the cubic interpolation weights~\citep{keys1981cubic}, $\abs{K_{\Tx,\Tz}-W_{\Tx,U}K_{U,\Tz}}=O(m^{-3})$.
 The simultaneous interpolation $W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx W_{X,U}K_{U,U}W_{U,X}^\top$.
@@
 
-Massively Scalable Gaussian Processes (MSGP) exploits structure as well: the kernel $K_{U,U}$ on a grid has Kronecker and Toeplitz matrix structure \cite{msgp}.
-Drawing on previous work on structured GPs \cite{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{X,X}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
-In addition, an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{X,X}}$ term in Eq.~\ref{llgradient} has been noted previously \cite{wilson2014fast}.
+Massively Scalable Gaussian Processes (MSGP) exploits structure as well: the kernel $K_{U,U}$ on a grid has Kronecker and Toeplitz matrix structure \citep{msgp}.
+Drawing on previous work on structured GPs \citep{cunningham2008fast, gilboa2015scaling}, MSGP uses linear conjugate gradient descent as a method for evaluating $K_{X,X}^{-1}\Ty$ efficiently for Eq.~\ref{llgradient}.
+In addition, an efficient eigendecomposition that carries over to the SKI kernel for the remaining $\log\abs{K_{X,X}}$ term in Eq.~\ref{llgradient} has been noted previously \citep{wilson2014fast}.
 
@@
 First, we identify a block-Toeplitz structure induced by the LMC kernel evaluated on a uniformly spaced grid of inputs.
-Next, we show how an LMC kernel can be decomposed into two block diagonal components, one of which has structure similar to that of SLFM \cite{seeger2005semiparametric}.
+Next, we show how an LMC kernel can be decomposed into two block diagonal components, one of which has structure similar to that of SLFM \citep{seeger2005semiparametric}.
 Both of these structures coordinate for fast matrix-vector multiplication $K\Tz$ with the covariance matrix $K$ for any vector $\Tz$.
@@
 
-For low-dimensional inputs, the above contributions offer an asymptotic and practical runtime improvement in hyperparameter estimation while also expanding the feasible kernel families to any differentiable LMC kernels, relative to COGP (Tab.~\ref{asymp}) \cite{nguyen2014collaborative}.
+For low-dimensional inputs, the above contributions offer an asymptotic and practical runtime improvement in hyperparameter estimation while also expanding the feasible kernel families to any differentiable LMC kernels, relative to COGP (Tab.~\ref{asymp}) \citep{nguyen2014collaborative}.
 
@@
 Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
-Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as a gradient-only high-level optimization routine for $\mcL$ \cite{zeiler2012adadelta}.
+Critically, we cannot access $\mcL$ itself, only $\nabla\mcL$, so we choose AdaDelta as a gradient-only high-level optimization routine for $\mcL$ \citep{zeiler2012adadelta}.
 
@@
 \end{align}
-This approximation improves as the size of $K$ increases, so, as in other work \cite{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
+This approximation improves as the size of $K$ increases, so, as in other work \citep{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
 
@@
 Second, we use \textsc{minres} instead of LCG as the Krylov-subspace inversion method used to compute inverses from MVMs.
-Iterative \textsc{minres} solutions to numerically semidefinite matrices monotonically improve in practice, as opposed to LCG \cite{fong2012cg}.
+Iterative \textsc{minres} solutions to numerically semidefinite matrices monotonically improve in practice, as opposed to LCG \citep{fong2012cg}.
 This is essential in GP optimization, where the diagonal noise matrix $\bse$, iid for each output, shrinks throughout learning.
@@
 
-Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K})\sqrt{\kappa_2})$ \cite{raykar2007fast}.
+Every AdaDelta iteration (invoking Algorithm~\ref{alg:grad}) then takes total time $\tilde{O}(\mvm(\tilde{K})\sqrt{\kappa_2})$ \citep{raykar2007fast}.
 This analysis holds as long as the error in the gradients is fixed and we can compute MVMs with the matrix $\partial_{\theta_j}K$ for each $j$ at least as fast as $\mvm({\tilde{K}})$.
@@
 \For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
-\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot \partial_{\theta_j}(\tbr_i)$\Comment{$t$ approximates the trace term of Eq.~\ref{eq:trace}}
+\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot \partial_{\theta_j}(\tbr_i)$\Comment{Eq.~\ref{eq:trace}}
 \State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
@@
 On a grid of inputs with ${X'}={U'}$, the SKI interpolation becomes $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in reduces the number of MVMs necessary to iteratively find an inverse~\cite{chan1994circulant}.
+In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in reduces the number of MVMs necessary to iteratively find an inverse~\citep{chan1994circulant}.
 
@@
 \end{align*}
-where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \cite{seeger2005semiparametric}.
+where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \citep{seeger2005semiparametric}.
 Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
@@
 
-\begin{table}[!h]
+\begin{table*}[!h]
   \caption{
@@
 \end{sc}
-\end{table}
+\end{table*}
 
@@
 
-The predictive mean can be computed in $O(1)$ time by $K_{*,X}\bsa\approx W_{*,{U'}}K_{{U'},{U'}}\bsa$ \cite{msgp}.
+The predictive mean can be computed in $O(1)$ time by $K_{*,X}\bsa\approx W_{*,{U'}}K_{{U'},{U'}}\bsa$ \citep{msgp}.
 
@@
 We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance.
-The dataset includes ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} In LLGP, we set $Q=1,R=2$, as recommended for LMC models on this dataset~\cite{alvarez2010efficient}.
+The dataset includes ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} In LLGP, we set $Q=1,R=2$, as recommended for LMC models on this dataset~\citep{alvarez2010efficient}.
 COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP.
@@
 
-\begin{figure}[!h]
+\begin{figure*}[!h]
 \centering
@@
 \label{fx2007-graph}
-\end{figure}
+\end{figure*}
 
@@
 
-\begin{wrapfigure}{r}{0.5\textwidth}
-%\begin{figure}[!h]
+\begin{figure}[!h]
 \centering
-{\includegraphics[width=0.4\columnwidth]{m_time_nlpd.pdf}}
+{\includegraphics[width=0.8\columnwidth]{m_time_nlpd.pdf}}
 \caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run $10$ times.}
 \label{fig:llgpweather}
-%\end{figure}
-\end{wrapfigure}
+\end{figure}
 
@@
 
-In this paper, we present LLGP, which we show adapts and accelerates SKI \cite{kiss-gp} for the problem of multi-output GP regression.
+In this paper, we present LLGP, which we show adapts and accelerates SKI \citep{kiss-gp} for the problem of multi-output GP regression.
 LLGP exploits structure unique to LMC kernels,  enabling a parsimonious representation of the covariance matrix, and gradient computations in $\tilde{O}\pa{\sqrt{\kappa_2}(m+n)}$.
@@
 Future work would extend the inputs to accept multiple dimensions.
-This can be done without losing internal structure in the kernel, again so long as only $k_q$ are stationary \cite{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
+This can be done without losing internal structure in the kernel, again so long as only $k_q$ are stationary \citep{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
 The cubic interpolation $W$ requires a number of terms exponential in the dimension, so projection into lower dimensions estimated in a supervised manner would be essential.
 Another useful line for investigation would be more informed stopping heuristics. % cut if space needed
-Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \cite{cutajar2016preconditioning}.
+Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \citep{cutajar2016preconditioning}.
 

### 2017-10-05 paper/paper.tex
Updated paper to reflect multi-dimensional input changes

@@
 % Camera ready: \usepackage[accepted]{aistats2018}
+\newif\ifanonymized
+\anonymizedtrue
+% Camera ready: \anonymizedfalse
 
 \usepackage[round]{natbib}
-% Camera ready: \usepackage[final]{nips_2017}
-% currently, doesn't print NIPS
-
 \usepackage[utf8]{inputenc} % allow utf-8 input
@@
 \usepackage{lmodern}
-\usepackage{hyperref}       % hyperlinks
+\usepackage[hidelinks]{hyperref}       % hyperlinks
 \usepackage{url}            % simple URL typesetting
@@
 While model estimation can be performed efficiently for single-output GPs \citep{msgp}, these assume stationarity, but in the multi-output case the cross-covariance interaction is not stationary.
-We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and one-dimensional inputs.
-When applied to real time series data, we find our theoretical improvement relative to the current state of the art is realized with LLGP being generally an order of magnitude faster while improving or maintaining predictive accuracy.
+We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and low-dimensional inputs.
+When applied to real time series and temporal-spatial data, we find our theoretical improvement relative to the current state of the art is realized with LLGP reducing training time while improving or maintaining predictive accuracy.
 \end{abstract}
-
-% NOTES
-% 8 pages of content
-% footnotes after punctuation
-% no vertical rules in tables
-% \subsubsection*{Acknowledgments} only in final paper
 
@@
 GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.
-Multi-output GPs frequently appear in time-series contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \citep{osborne2008towards, alvarez2010efficient}.
+Multi-output GPs frequently appear in time-series and geostatistical contexts, such as the problem of imputing missing temperature readings for sensors in different locations or missing foreign exchange rates and commodity prices given rates and prices for other goods over time \citep{osborne2008towards, alvarez2010efficient}.
 Efficient model estimation would enable researchers to quickly explore large spaces of parameterizations to find an appropriate one for their task.
@@
 With zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
-Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by the marginalization $y_*|\Ty_X$ \citep{williams1996gaussian}.
+Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by conditioning $y_*|\Ty_X$ \citep{williams1996gaussian}.
 Predictive accuracy is sensitive to a particular parameterization of our kernel, and model estimation is performed by maximizing data log likelihood with respect to parameters $\bsth$ of $K$, $\mcL(\bsth)=\log p(\Ty_{X}|X,\bsth)$.
-Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$:
+Gradient-based optimization methods then require the gradient with respect to every parameter $\theta_j$ of $\bsth$.
+Fixing $\bsa=K_{X,X}^{-1}\Ty$:
 
 \begin{align}
-\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{X,X}\bsa -\frac{1}{2}\Tr\pa{K_{X,X}^{-1}\partial_{\theta_j}K_{X,X}};\;\; \bsa=K_{X,X}^{-1}\Ty.\label{llgradient}
+\partial_{\theta_j}\mcL = \frac{1}{2}\bsa^\top\partial_{\theta_j}K_{X,X}\bsa -\frac{1}{2}\Tr\pa{K_{X,X}^{-1}\partial_{\theta_j}K_{X,X}}.\label{llgradient}
 \end{align}
@@
 \begin{align*}
-X'=\set{(d, x)}{d\in[D],x\in X_d}\subset{\mcX'};\;\; n=\card{X'}.
+X'=\set{(d, \Tx)}{d\in[D],\Tx\in X_d}\subset{\mcX'};\;\; n=\card{X'}.
 \end{align*}
 
-An LMC kernel $K$ is of the form 
+An LMC kernel $K$ is of the form:
 \begin{align}
-K([i,\Tx_i],[j,\Tx_j])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx_i-\Tx_j})+\epsilon_i1_{i=j},\label{lmcpointwise}
+K([i,\Tx],[j,\Tz])=\sum_{q=1}^Qb_{ij}^{(q)}k_q(\norm{\Tx-\Tz}),\label{lmcpointwise}
 \end{align} 
-where $k_q:\R\rightarrow\R$ is a stationary kernel on $\mcX$.
+where $k_q:\R\rightarrow\R$ are stationary kernels on $\mcX$.
 Typically, the positive semi-definite (PSD) matrices $B_q\in\R^{D\times D}$ formed by $b_{ij}^{(q)}$ are parameterized as $A_qA_q^\top+\diag\bsk_q$, with $A_q\in\R^{D\times R_q},\bsk_q\in\R_+^D$ and $R_q$ a preset rank.
 Importantly, even though each $k_q$ is stationary, $K$ is only stationary on $\mcX'$ if $B_q$ is Toeplitz.
-In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx_i],[j,\Tx_j])\neq K([i+1,\Tx_i],[j+1,\Tx_j])$.
+In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx],[j,\Tz])\neq K([i+1,\Tx],[j+1,\Tz])$.
 
@@
 Such matrices arise naturally when we examine the covariance matrix induced by a stationary kernel $k$ applied to a one-dimensional grid of inputs.
-Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that:
-\[
-T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}.
-\]
-Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, which enables MVMs $C\mat{\Tx & \textbf{0}}^\top=\mat{T\Tx &\textbf{0}}^\top$ in $O(n\log n)$ time.
+Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that $T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}$.
+Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, in $O(n\log n)$ time.
 This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \citep{cunningham2008fast}.
+When applied to grids of more than one dimension, the resulting covariance becomes block-Toeplitz with Toeplitz blocks (BTTB) \citep{msgp}.
+Consider a two-dimensional $n_x\times n_y$ grid with separations $\Delta_x,\Delta_y$.
+For a fixed pair of $x_1,x_2$, this grid contains a one-dimensional subgrid over varying $y$ values.
+The pairwise covariances for a stationary kernel in this subgrid also exhibit Toeplitz structure, since we still have $(x_1-x_2,y_i-y_j)=(x_1-x_2,y_i+\Delta_y-y_{j}-\Delta_y)=(x_1-x_2,y_{i+1}-y_{j+1})$.
+Ordering points in lexicographic order, so that the covariance matrix $K$ has $n_x^2$ order-$n_y$ blocks $K_{x_ix_j}$ with pairwise covariances between varying $y$ values between a pair of fixed $x$ values, the previous sentence implies $K_{x_ix_j}$ are Toeplitz.
+By similar reasoning, since for any $y_1,y_2$, $(x_i-x_j,y_1-y_j)=(x_{i+1}-x_{j+1},y_1-y_2)$, we have $K_{x_ix_j}=K_{x_{i+1}x_{j+1}}$, thus the block structure of $K$ is itself Toeplitz, and hence $K$ is BTTB.
+For higher dimensions, one can imagine a recursive block-Toeplitz structure.
+BTTB matrices themselves admit $O(n\log n)$ runtime for MVMs, with $n$ being the total grid size, or $n_xn_y$ in the two dimensional case, and if they are symmetric they can be represented with only their top row as well.
+The MVM runtime constant scales exponentially in the input dimension, however, so this approach is only applicable for low-input-dimension problems.
 
@@
 
-First, we identify a block-Toeplitz structure induced by the LMC kernel evaluated on a uniformly spaced grid of inputs.
+First, we identify a BTTB structure induced by the LMC kernel evaluated on a grid.
 Next, we show how an LMC kernel can be decomposed into two block diagonal components, one of which has structure similar to that of SLFM \citep{seeger2005semiparametric}.
@@
 We show a modification to SKI is compatible with the piecewise-differentiable LMC kernel, which is only invariant to translation along the indexing subspace $\mcX$ (the subkernels $k_q$ are stationary).
-This partial stationarity induces the previously-identified block-Toeplitz structure of the LMC kernel and enables acceleration of GP estimation for non-uniform inputs.
+This partial stationarity induces the previously-identified BTTB structure of the LMC kernel and enables acceleration of GP estimation for non-uniform inputs.
 
 For low-dimensional inputs, the above contributions offer an asymptotic and practical runtime improvement in hyperparameter estimation while also expanding the feasible kernel families to any differentiable LMC kernels, relative to COGP (Tab.~\ref{asymp}) \citep{nguyen2014collaborative}.
+LLGP is also, to the author's knowledge, first in experimentally validating the viability of SKI for input dimensions higher than one, which addresses a large class of GP applications that stand to benefit from tractable joint-output modeling.
 
@@
 
-For a given $\bsth$, we construct an operator $\tilde{K}$ which approximates MVMs with the exact covariance matrix evaluated, which for brevity we overload as $K\triangleq K_{X',X'}$, so $K\Tz\approx \tilde{K}\Tz$.
+For a given $\bsth$, we construct an operator $\tilde{K}$ which approximates MVMs with the exact covariance matrix, which for brevity we overload as $K\triangleq K_{X',X'}$, so $K\Tz\approx \tilde{K}\Tz$.
 Using only the action of MVM with the covariance operator, we derive $\nabla\mcL(\bsth)$.
@@
 If gradient norms drop below a proportion of the running max norm more than a pre-set number of times, we terminate (Fig.~\ref{fx2007-iterations}).
+This heuristic is avoidable since it is possible to evaluate $\mcL$ with only MVMs \citep{han2015large}, but using the heuristic proved sufficient and results in a simpler gradient-only optimization routine.
 
@@
 
-When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise over ${U'}$  (Eq.~\ref{lmcgrid}) holds for Toeplitz matrices $K_q$ formed by the stationary kernels $k_q$ evaluated the shared interpolating points $U$ for all outputs:
+When LMC kernels are evaluated on a grid of points for each output, so $X_d=U$, the simultaneous covariance matrix equation without noise over ${U'}$  (Eq.~\ref{lmcgrid}) holds for BTTB matrices $K_q$ formed by the stationary kernels $k_q$ evaluated the shared interpolating points $U$ for all outputs:
 \begin{align}
@@
 \end{align}
-We cannot fold $\bse$ into the interpolated term $K_{{U'},{U'}}$ since it does not correspond to a differentiable kernel.
+We cannot fold the Gaussian noise $\bse$ into the interpolated term $K_{{U'},{U'}}$ since it does not correspond to a differentiable kernel.
 However, since $\bse$ is diagonal, it is efficient to represent and multiply with.
@@
 
-\subsubsection{\textsc{sum}: sum representation}
-
-In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
-At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a Toeplitz matrix of order $m$, represented using only the top row.
+% Sum is pretty useless in practice,
+% but a useful baseline for comparison between representations in the supplement.
+% move the explanation of sum representation to supplement, then
+% \subsubsection{Sum representation}
+% 
+% In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
+At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a BTTB matrix of order $m$, represented using only the top row.
 In turn, multiplication $K_{{U'},{U'}}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
-The Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast Toeplitz MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
+The Kronecker MVM $(B_q\otimes K_q)\Tz$ may be expressed as $D$ fast BTTB MVMs with $K_q$ and $m$ dense MVMs with $B_q$.
 In turn, assuming $D\ll m$, the runtime for each of the $Q$ terms is $O(Dm\log m)$.
 
-\subsubsection{\textsc{bt}: block-Toeplitz representation}
+\subsubsection{Block-Toeplitz representation}
 
@@
 \end{align*}
-On a one-dimensional grid $U$, these matrices are Toeplitz because they are linear combinations of Toeplitz matrices.
+On a grid $U$, these matrices are BTTB because they are linear combinations of BTTB matrices.
 \textsc{bt} requires $D^2$ $m$-sized rows to represent each $T_{ij}$.
-Then, using the usual block matrix multiplication, an MVM $K_{{U'},{U'}}\Tz$ takes $O(D^2m\log m)$ time.
+Then, using the usual block matrix multiplication, an MVM $K_{{U'},{U'}}\Tz$ takes $O(D^2m\log m)$ time since each inner block MVM is accelerated due to BTTB structure.
 
 On a grid of inputs with ${X'}={U'}$, the SKI interpolation becomes $W=I$.
-In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan block-Toeplitz preconditioner in reduces the number of MVMs necessary to iteratively find an inverse~\citep{chan1994circulant}.
+In this case, using \textsc{bt} alone leads to a faster algorithm---applying the Chan BTTB preconditioner reduces the number of MVMs necessary to find an inverse~\citep{chan1994circulant}.
 
-\subsubsection{\textsc{slfm}: SLFM representation}
+\subsubsection{SLFM representation}
 
@@
 where $\textbf{A}=\mat{ \textbf{a}_{q'}}_{q'}\otimes I_m$ with $\mat{ \textbf{a}_{q'}}_{q'}$ a matrix of horizontally stacked column vectors \citep{seeger2005semiparametric}.
-Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is Toeplitz.
-Thus, \textsc{slfm} represents $K_{{U'},{U'}}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a Toeplitz order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
+Next, we rearrange the remaining term $\sum_q\diag\bsk_q \otimes K_q$ as $\blockdiag_d(T_d)$, where $T_d=\sum_q \kappa_{qd}K_q$ is BTTB.
+Thus, \textsc{slfm} represents $K_{{U'},{U'}}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a BTTB order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
 
@@
 An algorithm that uses this condition to decide between representations can minimize runtime (Tab.~\ref{asymp}).
-We found that \textsc{sum} is efficient in practice for $Q=1$.
 
@@
   \caption{
+   Asymptotic Runtimes.
 For both LLGP and COGP, $m$ is a configurable parameter that increases up to $n$ to improve accuracy.
@@
 Computing $\nabla\mcL$ at $\bsth$ requires an up-front cost in addition to the per-hyperparameter cost for each $\theta_j\in\bsth$.
-Multiplicative log terms in $\kappa_2, m$ are hidden.
+Multiplicative log terms in $\kappa_2, m$ are hidden, as are exponential dependencies of the input dimension.
   }
@@
 In both cases, lower values represent better performance.
-We evaluated the performance of our flexible representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm for a grid of different kernels (see Supplement).\footnote{Hyperparameters for the stopping condition, code, and benchmarking scripts are available in
-\url{https://github.com/vlad17/runlmc}.} % camera ready
-%\texttt{<anonymous repository>}.}
-\subsection{Foreign exchange rate prediction (FX2007)}\label{fx2007-results}
+We evaluated the performance of our representations of the kernel, \textsc{bt} and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm over a variety of different kernels (see the supplement).\footnote{Hyperparameters, data, code, and benchmarking scripts are available in
+  \ifanonymized
+  \texttt{<anonymous repository>}.
+  \else
+  \url{https://github.com/vlad17/runlmc}.
+  \fi
+}
+
+\subsection{Foreign exchange rates (FX2007)}\label{fx2007-results}
 
 We replicate the medium-sized dataset from COGP as an application to evaluate LLGP performance.
-The dataset includes ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.\footnote{Data are from \url{http://fx.sauder.ubc.ca/data.html}} In LLGP, we set $Q=1,R=2$, as recommended for LMC models on this dataset~\citep{alvarez2010efficient}.
+The dataset includes ten foreign currency exchange rates---CAD, EUR, JPY, GBP, CHF, AUD, HKD, NZD, KRW, and MXN---and three precious metals---XAU, XAG, and XPT---implying that $D=13$.
+In LLGP, we set $Q=1,R=2$, as recommended for LMC models on this dataset~\citep{alvarez2010efficient}.
 COGP roughly corresponds to the the SLFM model, which has a total of 94 hyperparameters, compared to 53 for LLGP.
@@
 \begin{table}[!h]
-  \caption{Average predictive performance and training time over $10$ runs for LLGP and COGP on the FX2007 dataset.
+  \caption{Average Predictive Performance and Training Time Over $10$ Runs for LLGP and COGP on the FX2007 Dataset.
 Parenthesized values are standard error.
@@
 
-We use the default COGP parameters\footnote{\url{https://github.com/trungngv/cogp}, commit \texttt{3b07f621ff11}} LLGP was run with the same parameters as in FX2007, simulating the SLFM model.
-We tested LLGP models on $m=\{500, 1000\}$ interpolating points.
+We use the default COGP parameters.\footnote{\url{https://github.com/trungngv/cogp}} We tested LLGP models on from $500$ to $1000$ interpolating points.
 
 \begin{table}[!h]
-  \caption{Average predictive performance and training time over $10$ runs of LLGP and COGP on the weather dataset.
+  \caption{Average Predictive Performance and Training Time Over $10$ Runs of LLGP and COGP on the Weather Dataset.
 Parenthesized values are standard error.
@@
 \end{center}
+
 \end{table}
@@
 
+\subsection{NOAA dataset}\label{noaa-bench}
+
+TODO. NOAA dataset intro. Mention use of split kernel, what that is (Tbl.~\ref{noaa-tbl}). Also TODO: current numbers are all on Princeton machines.
+
+\begin{table}[!h]
+  \caption{Average Predictive Performance and Training Time Over $10$ Runs of LLGP and COGP (GPy?) on the NOAA Dataset.
+Parenthesized values are standard error.
+Both LLGP and COGP (GPy?) trained the SLFM model.}
+\label{noaa-tbl}
+\begin{center}
+  \begin{small}
+    \begin{sc}
+\begin{tabular}{lcc}\toprule
+Metric & LLGP & COGP (GPy?)\\
+\midrule
+seconds & $\textbf{0.00}$ ($\textbf{0.00}$) &  $\infty$ ($0.00$) \\
+SMSE & $\textbf{0.00}$ ($\textbf{0.00}$) & $\infty$ ($0.00$) \\
+NLPD & $\textbf{0.00}$ ($\textbf{0.00}$) & $\infty$ ($0.00$) \\
+\bottomrule
+\end{tabular}
+\end{sc}
+\end{small}
+\end{center}
+
+\end{table}
+
 \section{Conclusion}\label{conclusion}
@@
 
-Future work would extend the inputs to accept multiple dimensions.
-This can be done without losing internal structure in the kernel, again so long as only $k_q$ are stationary \citep{msgp}: Toeplitz covariance matrices become block-Toeplitz with Toeplitz-blocks (BTTB).
-The cubic interpolation $W$ requires a number of terms exponential in the dimension, so projection into lower dimensions estimated in a supervised manner would be essential.
-Another useful line for investigation would be more informed stopping heuristics. % cut if space needed
-Finally, an extension to non-Gaussian noise is also feasible in a matrix-free manner by following prior work \citep{cutajar2016preconditioning}.
+Future work could extend LLGP to accept large input dimensions, though most GP use cases are covered by low-dimensional inputs.
+Finally, an extension to non-Gaussian noise and use of LLGP as a preconditioner for fine-tuned exact GP models is also feasible in a manner following prior work \citep{cutajar2016preconditioning}.
+
+\ifanonymized
+
+\else
+\subsubsection*{Acknowledgments}
+(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof. Andrew Wilson.
+\fi
 

### 2017-10-05 data/noaa/clean.py
Added noaa data

@@ -1,0 +1,162 @@
+import datetime
+import numpy as np
+import os
+import pandas as pd
+import tempfile
+import tarfile
+from zipfile import ZipFile
+
+station_zip = 'wbanmasterlist.psv.zip'
+climate_tar = '2006_daily.tar.gz'
+
+assert station_zip in os.listdir(), station_zip
+assert climate_tar in os.listdir(), climate_tar
+
+with tempfile.TemporaryDirectory() as tmp:
+    print('expanding', station_zip, 'and', climate_tar, 'into', tmp)
+    with ZipFile(station_zip) as stations:
+        stations.extractall(tmp)
+    with tarfile.open(climate_tar) as climate:
+        climate.extractall(tmp)
+    climate_dir = os.path.join(tmp, '2006_daily')
+    stations_file = os.path.join(tmp, 'wbanmasterlist.psv')
+
+    print('reading in raw data')
+    stations = pd.read_csv(stations_file, sep='|', header=0)
+    climates = [os.path.join(climate_dir, f) for f in os.listdir(climate_dir)]
+    daily_dfs = [pd.read_csv(f, header=0, encoding='ISO-8859-1')
+                 for f in climates]
+    for df in daily_dfs:
+        df.columns = df.columns.str.strip()
+    climate = pd.concat(daily_dfs, ignore_index=True)
+
+print('cleaning climate data')
+climate['time'] = pd.to_datetime(climate['YearMonthDay'], format='%Y%m%d')
+climate['time'] = (climate['time'] - pd.to_datetime('2006-01-01')
+                   ) / datetime.timedelta(days=1)
+oldcols = ['Wban Number', 'Avg Temp', 'Avg Dew Pt',
+           'Pressure Avg Sea Level', 'Wind Avg Speed', 'time']
+newcols = ['wban', 'temp', 'dew', 'pressure', 'wind', 'time']
+climate = climate.rename(columns=dict(zip(oldcols, newcols)))
+climate = climate[newcols]
+print('climate NAs per column')
+print(len(climate) - climate.count())
+
+print('cleaning station data')
+oldcols = ['WBAN_ID', 'LOCATION']
+newcols = ['wban', 'loc']
+stations = stations.rename(columns=dict(zip(oldcols, newcols)))
+stations = stations[newcols]
+print('stations NAs per column')
+print(len(stations) - stations.count())
+
[…]

### 2017-10-05 benchmarks/asv/fx2007/fx2007.py
Upgraded to 16 threads / 15 trace terms (beefier machine available)

@@ -78,7 +78,7 @@ def main():
         cogp_runs = 10
         interpolation_points = None
         inducing_points = 100
-        nthreads = 4
+        nthreads = 16
 
     llgp_stats = Suite(runs, interpolation_points, nthreads).setup_cache()
     print('---> llgp Q1R2 m', interpolation_points, statsline(llgp_stats))

### 2017-10-05 data/noaa/clean.py
Updated noaa dataset to have float conversion

@@ -57,6 +57,13 @@ print('stations in climate missing location data', missing, 'of', tot)
 
 joined = climate.merge(stations, on='wban', how='left')
 tot = len(joined)
+
+
+numerics = ['temp', 'dew', 'pressure', 'wind', 'time']
+for name in numerics:
+    cleanstr = joined[name].astype('str').str.replace(r'[^\d\.]', '')
+    joined[name] = pd.to_numeric(cleanstr)
+
 joined = joined.dropna()
 print('left-joined climate/stations data has',
       len(joined), 'of', tot, 'clean rows')
@@ -114,7 +121,7 @@ print('# of bad lat/lon parses:', badparse)
 
 fname = 'noaa.csv'
 print('exporting cleaned noaa dataset to', fname)
-joined.to_csv(fname)
+joined.to_csv(fname, index=False)
 
 fname = 'stations.pdf'
 print('plotting stations with weather data on', fname)

### 2017-10-05 benchmarks/asv/fx2007/fx2007.py
Re-ran results on Cal machines

@@ -76,7 +76,7 @@ def main():
     else:
         runs = 10
         cogp_runs = 10
-        interpolation_points = None
+        interpolation_points = 238
         inducing_points = 100
         nthreads = 16
 

### 2017-10-06 benchmarks/asv/fx2007/fx2007.py
updated fx2007 results

@@ -76,7 +76,7 @@ def main():
     else:
         runs = 10
         cogp_runs = 10
-        interpolation_points = 238
+        interpolation_points = 234
         inducing_points = 100
         nthreads = 16
 

### 2017-10-06 runlmc/models/test_interpolated_llgp.py
reduced interpolated llgp test size

@@ -89,7 +89,7 @@ class LMCTestUtils:
     @staticmethod
     def _case_1d(input_dim):
         kernels = [RBF(inv_lengthscale=3)]
-        szs = [30]
+        szs = [10]
         coregs = [np.array([[1]])]
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
 
@@ -97,7 +97,7 @@ class LMCTestUtils:
     def _case_2d(input_dim):
         kernels = [RBF(inv_lengthscale=3),
                    RBF(inv_lengthscale=2)]
-        szs = [30, 40]
+        szs = [10, 15]
         coregs = [np.array(x).reshape(1, -1) for x in [[1, 2], [3, 4]]]
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
 
@@ -106,7 +106,7 @@ class LMCTestUtils:
         assert input_dim >= 2
         kernels = [RBF(inv_lengthscale=3, active_dims=split1),
                    RBF(inv_lengthscale=2, active_dims=split2)]
-        szs = [30, 40]
+        szs = [10, 15]
         coregs = [np.array(x).reshape(1, -1) for x in [[1, 2], [3, 4]]]
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
 
@@ -114,7 +114,7 @@ class LMCTestUtils:
     def _case_multirank(input_dim):
         kernels = [RBF(inv_lengthscale=3),
                    RBF(inv_lengthscale=2)]
-        szs = [30, 40]
+        szs = [10, 15]
         coregs = [np.array([[1, 2], [3, 4]]), np.array([[1, 1]])]
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
 
@@ -123,7 +123,7 @@ class LMCTestUtils:
         kernels = [RBF(inv_lengthscale=3),
                    RBF(inv_lengthscale=2),
                    RBF(inv_lengthscale=1)]
-        szs = [10, 12, 14, 12, 10]
+        szs = [6, 6, 6, 6, 6]
         coregs = [np.array(x).reshape(1, -1) for x in
                   [[1, 1, 1, 1, 2], [2, 1, 2, 1, 2], [-1, 1, -1, -1, -1]]]
         return ExactAnalogue(kernels, szs, coregs, indim=input_dim)
@@ -164,8 +164,9 @@ class LMCTestUtils:
         for (out_name, out), (in_name, in_dim) in itertools.product(
                 outs.items(), ins.items()):
[…]

### 2017-10-06 benchmarks/asv/weather/weather.py
increase number of weather runs

@@ -77,7 +77,7 @@ def main():
         inducing_points = 10
         nthreads = ''
     else:
-        runs = 10
+        runs = 100
         cogp_runs = 10
         interpolating_points = [500, 600, 700, 800, 900, 1000]
         inducing_points = 200

### 2017-10-08 benchmarks/asv/weather/weather.py
old 10 run value

@@ -77,7 +77,7 @@ def main():
         inducing_points = 10
         nthreads = ''
     else:
-        runs = 100
+        runs = 10
         cogp_runs = 10
         interpolating_points = [500, 600, 700, 800, 900, 1000]
         inducing_points = 200

### 2017-10-08 benchmarks/asv/weather/makepics.py
only run weather 500,1000 (too noisy for graph)

@@ -1,43 +1,0 @@
-if __name__ == '__main__':
-    import numpy as np
-    import matplotlib as mpl
-    mpl.use('Agg')
-    import matplotlib.pyplot as plt
-
-    import subprocess
-
-    cmd = ['grep', '-e', '--->', 'benchmarks/weather-out/stdout-weather.txt']
-    process = subprocess.Popen(cmd,
-                               stdout=subprocess.PIPE,
-                               stderr=subprocess.PIPE,
-                               universal_newlines=True)
-    out, err = process.communicate()
-    print(err)
-
-    lines = out.split('\n')
-
-    llgp = [line.split() for line in lines]
-    llgp = [line for line in llgp if line]
-    llgp, cogp = llgp[:-1], llgp[-1]
-
-    ms = [int(line[4]) for line in llgp]
-    times = [float(line[6]) for line in llgp]
-    se_times = [float(line[8][:-1]) for line in llgp]
-    smses = [float(line[10]) for line in llgp]
-    nlpds = [float(line[14]) for line in llgp]
-    se_nlpds = [float(line[16][:-1]) for line in llgp]
-
-    fig, ax1 = plt.subplots()
-    ax1.errorbar(ms, nlpds, yerr=se_nlpds, c='r', marker='s', ecolor='pink')
-    ax1.set_ylabel('NLPD', color='r')
-    ax1.set_xlabel(r'interpolating points $m$')
-    ax2 = ax1.twinx()
-    jittered_ms = [m - 10 for m in ms]
-    ax2.errorbar(jittered_ms, times, yerr=se_times, c='b', marker='o', ls=':', ecolor='lightblue')
-    ax2.set_ylabel('runtime (s)', color='b')
-    ax2.set_xlim([min(ms) - 30, max(ms) + 20])
-
-    fig.tight_layout()
-    plt.savefig('benchmarks/weather-out/m_time_nlpd.eps', format='eps', bbox_inches='tight')
-
-    print('saved m_time_nlpd.eps')

### 2017-10-08 benchmarks/asv/weather/run.sh

@@ -66,4 +66,3 @@ cd $REPOROOT
 
 OMP_NUM_THREADS=1 PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/benchmarks/asv/weather/weather.py $IS_VALIDATION $OUTFOLDER 2>&1 | tee $OUTFOLDER/stdout-weather.txt | egrep -e '--->' 
 
-python3 $REPOROOT/benchmarks/asv/weather/makepics.py

### 2017-10-08 benchmarks/asv/weather/weather.py

@@ -79,7 +79,7 @@ def main():
     else:
         runs = 10
         cogp_runs = 10
-        interpolating_points = [500, 600, 700, 800, 900, 1000]
+        interpolating_points = [500, 1000]
         inducing_points = 200
         nthreads = 16
 

### 2017-10-08 runlmc/models/interpolated_llgp.py

@@ -128,7 +128,7 @@ class InterpolatedLLGP(MultiGP):
     def __init__(self, Xs, Ys, normalize=True,  # pylint: disable=too-many-arguments,dangerous-default-value,too-many-locals
                  lo=None, hi=None, m=None, name='lmc',
                  metrics=False, prediction='on-the-fly', max_procs=None,
-                 trace_iterations=15,
+                 trace_iterations=15, tolerance=1e-4,
                  functional_kernel=None):
         super().__init__(Xs, Ys, normalize=normalize, name=name)
         self.update_model(False)

### 2017-10-08 benchmarks/benchlib/cogp_synth.m
initial synthetic data version

@@ -1,0 +1,69 @@
+% copied from various parts of COGP paper code
+% original repo: https://github.com/trungngv/cogp
+% repo with a argument count bug corrected:
+% https://github.com/vlad17/cogp
+
+% Uses similar assumptions to cogp_fx2007: datadir and runs required
+% datadir should go the csv directory
+% requires M, runs and datadir to be defined
+
+addpath(genpath([tempdir 'cogp']));
+format long;
+
+x = csvread([datadir 'xss.csv']);
+y = csvread([datadir 'yss.csv']);
+
+% test on upper quadrant only, for last output
+nlast = sum(~isnan(y(:, end)));
+sel = all(x>=0.5, 2);
+sel(1:end - nlast + 1) = 0;
+ytest = y;
+xtest = x;
+y(sel, end) = nan;
+
+rng(1234,'twister');
+
+[y,ymean,ystd] = standardize(y,[],[]);
+
+cf.covfunc_g  = 'covSEard';
+cf.lrate      = 1e-2;
+cf.momentum   = 0.9;
+cf.lrate_hyp  = 1e-5;
+cf.lrate_beta = 1e-4;
+cf.momentum_w = 0.9;
+cf.lrate_w    = 1e-6;
+cf.learn_z    = true;
+cf.momentum_z = 0.0;
+cf.lrate_z    = 1e-4;
+cf.maxiter = maxiter;
+cf.nbatch = 100; %2000;
+cf.beta = 1/0.1;
+cf.initz = 'random';
+cf.w = ones(size(y,2),2);
+cf.monitor_elbo = 100;
+cf.fix_first = false;
+Q = 2;
+
+smses = zeros(runs,1);
+nlpds = zeros(runs,1);
+times = zeros(runs,1);
+for r=1:runs
+  par.g = cell(Q,1);
+  tic;
+  [elbo,par] = slfm2_learn(x,y,M,par,cf);
+  times(r) = toc;
+  [mu,vaar,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest(sel,:),size(y,2));
+  mu = mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1);
+  fvar = vaar.*repmat(ystd.^2,size(mu,1),1);
+  smses(r) = mysmse(ytest(sel,end),mu(:,end),ymean(end));
+  nlpds(r) = mynlpd(ytest(sel,end),mu(:,end),fvar(:,end));  
+end
[…]

### 2017-10-08 benchmarks/benchlib/standard_tester.py

@@ -414,3 +414,62 @@ def latex_table(filename, cols, col_results):
     outdir = sys.argv[2]
     with open(outdir + '/' + filename, 'w') as f:
         f.write(begin + latex + end)
+
+
+def gen_random_k():
+    """Generates a Q=2 SLFM kernel with varying lengthscales for 5 outputs."""
+    ks = [RBF(name='rbf1', inv_lengthscale=5),
+          RBF(name='rbf2', inv_lengthscale=0.5)]
+    indeps = [RBF(name='indep{}'.format(i), inv_lengthscale=2)
+              for i in range(5)]
+    return FunctionalKernel(
+        D=5, lmc_kernels=[], lmc_ranks=[], slfm_kernels=ks, indep_gp=indeps)
+
+def cogp_synth(num_runs, inducing_pts, nthreads, nbatch, it):
+    _download_cogp()
+    benchmark_dir = os.path.join(
+        _download_own_data(), os.pardir, 'benchmarks', 'benchlib')
+    datadir = os.path.join(_download_own_data(), 'synth') + os.path.sep
+    # This runs the COGP code; only learning is timed
+    cmd = ['matlab', '-nojvm', '-r',
+           re.sub(r'[\s+]', '', """
+              datadir='{}';
+              maxiter={};
+              M={};
+              runs={};
+              maxNumCompThreads({});
+              cogp_synth;
+              exit""")
+           .format(datadir, it, inducing_pts, num_runs, nthreads)]
+    with open(TMP + '/out-synth-{}-{}'.format(nbatch, inducing_pts), 'w') as f:
+        f.write(' '.join(cmd))
+    process = subprocess.Popen(
+        cmd,
+        stdout=subprocess.PIPE,
+        stderr=subprocess.PIPE,
+        universal_newlines=True,
+        cwd=benchmark_dir,
+        env=env_no_omp())
+    mout, err = process.communicate()
+    print(err)
+    with open(TMP + '/out-{}'.format(num_runs), 'w') as f:
+        f.write(mout)
+
+    ending = mout[mout.find('mean/stderr times'):]
+    match = re.match('\D*([-+e\.\d]*)\s*([-+e\.\d]*)', ending)
+    m_time, se_time = float(match.groups()[0]), float(match.groups()[1])
+    ending = ending[ending.find('mean/stderr smses'):]
[…]

### 2017-10-08 data/synth/mkdata.py

@@ -1,0 +1,38 @@
+from runlmc.lmc.likelihood import ExactLMCLikelihood
+from runlmc.lmc.functional_kernel import FunctionalKernel
+from standard_tester import gen_random_k
+import numpy as np
+
+np.random.seed(1234)
+
+D = 5
+NperD = 100
+fk = gen_random_k()
+fk.coreg_vecs[0][:] = np.random.rand(5)
+fk.coreg_vecs[1][:] = np.random.rand(5)
+print('generated Q = 2 SLFM RBF kernel')
+
+Xs = np.random.rand(D, NperD, 2)
+print('generated', Xs.shape, 'inputs')
+
+fk.set_input_dim(2)
+K = ExactLMCLikelihood.kernel_from_indices(list(Xs), list(Xs), fk)
+K += np.diag(np.ones(len(K)) / 10)
+print('generated', K.shape, 'kernel')
+
+L = np.linalg.cholesky(K)
+Ys = L.dot(np.random.randn(len(K))).reshape(D, -1)
+print('generated', Ys.shape, 'sampled outputs')
+
+np.save('xss.npy', Xs)
+np.save('yss.npy', Ys)
+
+# cogp has a silly representation that's fixed-length
+np.savetxt('xss.csv', Xs.reshape(-1, 2), delimiter=",")
+cogp_ys = np.empty((D * NperD, D))
+cogp_ys[:] = np.nan
+for i in range(D):
+    start = i * NperD
+    end = (i + 1) * NperD
+    cogp_ys[start:end, i] = Ys[i]
+np.savetxt('yss.csv'.format(i), cogp_ys, delimiter=",")

### 2017-10-08 runlmc/models/interpolated_llgp.py
lowered minimum value to parallelize

@@ -164,7 +164,7 @@ class InterpolatedLLGP(MultiGP):
     def _generate_pool(self, max_procs):
         self._check_omp(max_procs)
         max_procs = cpu_count() if max_procs is None else max_procs
-        if max_procs == 1 or len(self.y) < 1000:
+        if max_procs == 1 or len(self.y) < 300:
             _LOG.info('InterpolatedLLGP (%d hyperparams) will run serially',
                       len(self.param_array))
             return None

### 2017-10-08 data/synth/mkdata.py
increase synthetic to 5000 size

@@ -6,7 +6,7 @@ import numpy as np
 np.random.seed(1234)
 
 D = 5
-NperD = 100
+NperD = 1000
 fk = gen_random_k()
 fk.coreg_vecs[0][:] = np.random.rand(5)
 fk.coreg_vecs[1][:] = np.random.rand(5)

### 2017-10-09 benchmarks/benchlib/standard_tester.py
synthetic dataset on 5k POINTS

@@ -147,6 +147,24 @@ def weather():
 
     return xss, yss, test_xss, test_yss, sensors
 
+def synth():
+    datadir = _download_own_data()
+    xss = np.load(os.path.join(datadir, 'synth', 'xss.npy'))
+    yss = np.load(os.path.join(datadir, 'synth', 'yss.npy'))
+    xss = list(xss)
+    yss = list(yss)
+
+    # the values for the last output in the upper-right quadrant of the unit
+    # square are the test values.
+    sel = np.all(xss[-1] >= 0.5, axis=1)
+    e = np.array([]).reshape(0, 2)
+    test_xss = [e] * 4 + [xss[-1][sel]]
+    e = e.reshape(0, 1)
+    test_yss = [e] * 4 + [yss[-1][sel]]
+    xss[-1] = xss[-1][~sel, :]
+    yss[-1] = yss[-1][~sel]
+    return xss, yss, test_xss, test_yss
+
 
 def toy_sinusoid():
     # Adapts the 2-output toy problem from
@@ -425,7 +443,14 @@ def gen_random_k():
     return FunctionalKernel(
         D=5, lmc_kernels=[], lmc_ranks=[], slfm_kernels=ks, indep_gp=indeps)
 
-def cogp_synth(num_runs, inducing_pts, nthreads, nbatch, it):
+def synth_gen():
+    """Kernel generators associated for gen_random_k()"""
+    return lambda: [], lambda: [], \
+        lambda: [RBF(name='rbf1'), RBF(name='rbf2')], \
+        lambda: [RBF(name='indep{}'.format(i)) for i in range(5)]
+
+
+def cogp_synth(num_runs, inducing_pts, nthreads, it):
     _download_cogp()
     benchmark_dir = os.path.join(
         _download_own_data(), os.pardir, 'benchmarks', 'benchlib')
@@ -441,7 +466,7 @@ def cogp_synth(num_runs, inducing_pts, nthreads, nbatch, it):
               cogp_synth;
               exit""")
            .format(datadir, it, inducing_pts, num_runs, nthreads)]
-    with open(TMP + '/out-synth-{}-{}'.format(nbatch, inducing_pts), 'w') as f:
+    with open(TMP + '/out-synth-{}'.format(inducing_pts), 'w') as f:
         f.write(' '.join(cmd))
     process = subprocess.Popen(
         cmd,
[…]

### 2017-10-09 benchmarks/synth/synth.py

@@ -1,0 +1,104 @@
+# compares to cogp on fake data
+
+import os
+os.environ['OMP_NUM_THREADS'] = '1'
+
+import numpy as np
+from standard_tester import *
+import pickle
+                                                                                                                                                                                                              
+import matplotlib as mpl                                                                                                                                                                                                         
+mpl.use('Agg')                                                                                                                                                                                                                   
+import matplotlib.pyplot as plt 
+
+def dump(x, f):
+    with open('out/' + f + '.pkl', 'wb') as handle:
+        pickle.dump(x, handle, pickle.HIGHEST_PROTOCOL)
+
+def main():
+    np.random.seed(1234)
+    activate_logs()
+
+    max_it = 100
+    interpolating_points = [50, 50]
+    nthreads = 16
+    inducing_points = 500 # like in COGP sarcos
+    iters = [50, 100, 150, 200, 250, 300, 350, 400, 450, 500]
+
+    kgen, rgen, slfmgen, indepgen = synth_gen()
+    
+    times = []
+    times, smses, nlpds = [], [], []
+    lmc = None
+    ctr = 0
+    timer = contexttimer.Timer()
+    xss, yss, test_xss, test_yss = synth()
+    fk = FunctionalKernel(
+        D=len(xss),
+        lmc_kernels=kgen(),
+        lmc_ranks=rgen(),
+        slfm_kernels=slfmgen(),
+        indep_gp=indepgen())
+
+    def perf_cb():
+        nonlocal times, smses, nlpds, lmc, ctr, timer
+        ctr += 1
+        if ctr % 10 != 0:
+            return
+
+        timer.__exit__(None, None, None)
+        np.save(TMP + 'lmc-iter-{}'.format(ctr), lmc.param_array)
+        pred_yss, pred_vss = lmc.predict(test_xss)
+        smses.append(smse(test_yss, pred_yss, yss))
+        nlpds.append(nlpd(test_yss, pred_yss, pred_vss))
+        prev = times[-1] if times else 0
+        times.append(timer.elapsed + prev)
+        print('[llgp] time {:.0f} smse {:.2f} nlpd {:.2f}'.format(
[…]

### 2017-10-09 benchmarks/benchlib/cogp_synth.m
updated comment

@@ -5,7 +5,7 @@
 
 % Uses similar assumptions to cogp_fx2007: datadir and runs required
 % datadir should go the csv directory
-% requires M, runs and datadir to be defined
+% requires maxiter, M, runs and datadir to be defined
 
 addpath(genpath([tempdir 'cogp']));
 format long;

### 2017-10-09 benchmarks/benchlib/cogp_synth.m
added synth benchmark

@@ -26,6 +26,7 @@ rng(1234,'twister');
 [y,ymean,ystd] = standardize(y,[],[]);
 
 cf.covfunc_g  = 'covSEard';
+cf.covfunc_h = 'covSEard';
 cf.lrate      = 1e-2;
 cf.momentum   = 0.9;
 cf.lrate_hyp  = 1e-5;
@@ -36,23 +37,24 @@ cf.learn_z    = true;
 cf.momentum_z = 0.0;
 cf.lrate_z    = 1e-4;
 cf.maxiter = maxiter;
-cf.nbatch = 2000;
+cf.nbatch = nbatch;
 cf.beta = 1/0.1;
 cf.initz = 'random';
 cf.w = ones(size(y,2),2);
-cf.monitor_elbo = 10;
-cf.fix_first = false;
+cf.monitor_elbo = 50;
+cf.fix_first = true;
 Q = 2;
 
 smses = zeros(runs,1);
+MM.g = M; MM.h = M;
 nlpds = zeros(runs,1);
 times = zeros(runs,1);
 for r=1:runs
-  par.g = cell(Q,1);
+  par.g = cell(Q,1); par.task = cell(size(y,2),1);
   tic;
-  [elbo,par] = slfm2_learn(x,y,M,par,cf);
+  [elbo,par] = slfm_learn(x,y,MM,par,cf);
   times(r) = toc;
-  [mu,vaar,mu_g,var_g] = slfm2_predict(cf.covfunc_g,par,xtest(sel,:),size(y,2));
+  [mu,vaar,mu_g,var_g] = slfm_predict(cf.covfunc_g, cf.covfunc_h,par,xtest(sel,:));
   mu = mu.*repmat(ystd,size(mu,1),1) + repmat(ymean,size(mu,1),1);
   fvar = vaar.*repmat(ystd.^2,size(mu,1),1);
   smses(r) = mysmse(ytest(sel,end),mu(:,end),ymean(end));

### 2017-10-09 benchmarks/benchlib/standard_tester.py

@@ -376,7 +376,8 @@ def cogp_weather(num_runs, M, nthreads):
     test_fx = ['cam', 'chi']
     cogp_mu = pd.read_csv(TMP + '/cogp-weather-mu{}{}'.format(num_runs, M),
                           header=None, names=test_fx)
-    cogp_var = pd.read_csv(TMP + '/cogp-weather-var{}{}'.format(num_runs, M))
+    cogp_var = pd.read_csv(TMP + '/cogp-weather-var{}{}'.format(num_runs, M),
+                          header=None, names=test_fx)
 
     stats = [(m_time, se_time), (m_smse, se_smse), (m_nlpd, se_nlpd)]
     return stats, cogp_mu, cogp_var
@@ -450,7 +451,7 @@ def synth_gen():
         lambda: [RBF(name='indep{}'.format(i)) for i in range(5)]
 
 
-def cogp_synth(num_runs, inducing_pts, nthreads, it):
+def cogp_synth(num_runs, inducing_pts, nthreads, nbatch, maxiter):
     _download_cogp()
     benchmark_dir = os.path.join(
         _download_own_data(), os.pardir, 'benchmarks', 'benchlib')
@@ -460,12 +461,13 @@ def cogp_synth(num_runs, inducing_pts, nthreads, it):
            re.sub(r'[\s+]', '', """
               datadir='{}';
               maxiter={};
+              nbatch={};
               M={};
               runs={};
               maxNumCompThreads({});
               cogp_synth;
               exit""")
-           .format(datadir, it, inducing_pts, num_runs, nthreads)]
+           .format(datadir, maxiter, nbatch, inducing_pts, num_runs, nthreads)]
     with open(TMP + '/out-synth-{}'.format(inducing_pts), 'w') as f:
         f.write(' '.join(cmd))
     process = subprocess.Popen(

### 2017-10-09 benchmarks/synth/synth.py

@@ -27,109 +27,36 @@ def main():
     np.random.seed(1234)
     activate_logs()
 
-    max_it = 30
-    interpolating_points = [100, 100]
+    llgp_runs = 3
+    cogp_runs = 3
+    cogp_max_it = 100 # like in COGP sarcos
+    interpolating_points = [25, 25]
     nthreads = 16
-    inducing_points = 1000  # like in COGP sarcos
-    iters = [50, 100, 150, 200, 250]
+    inducing_points = [500]
+    nbatches = [1000]
 
     kgen, rgen, slfmgen, indepgen = synth_gen()
-
-    times = []
-    times, smses, nlpds = [], [], []
-    lmc = None
-    ctr = 0
-    timer = contexttimer.Timer()
     xss, yss, test_xss, test_yss = synth()
-    fk = FunctionalKernel(
-        D=len(xss),
-        lmc_kernels=kgen(),
-        lmc_ranks=rgen(),
-        slfm_kernels=slfmgen(),
-        indep_gp=indepgen())
+    stats = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
+                test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
+                tolerance=1e-3) # reduce tol a bit b/c of large problem size
+    dump(stats, 'llgp_stats')
+    # stats = load('llgp_stats')
 
-    def perf_cb():
-        nonlocal times, smses, nlpds, lmc, ctr, timer
-        ctr += 1
-        if ctr % 2 != 0:
-            return
+    all_stats = [stats]
+    colnames = ['LLGP']
+    
+    for ind, nb in zip(inducing_points, nbatches):
+        cogp_stats = cogp_synth(
+            cogp_runs, ind, nthreads, nb, cogp_max_it)
+        dump(cogp_stats, 'cogp_stats-{}-{}'.format(ind, nb))
+        # cogp_stats = load('cogp_stats-{}-{}'.format(ind, nb))   
+        all_stats.append(cogp_stats)
+        colnames.append(r'\begin{tabular}{c}LLGP\\$m=' +
+                        str(ind) + ',n_b=' + str(nb) +
+                        r'$\end{tabular}')
 
-        timer.__exit__(None, None, None)
-        np.save(TMP + 'lmc-iter-{}'.format(ctr), lmc.param_array)
[…]

### 2017-10-09 benchmarks/pictures/fxmetrics.py
updated pictures to incl. weather

@@ -2,7 +2,6 @@
 # Computes explicit log likelihood and plots optimization trace
 
 import os
-os.environ['OMP_NUM_THREADS'] = '1'
 
 import matplotlib as mpl
 mpl.use('Agg')
@@ -33,7 +32,7 @@ print('running MINRES metrics')
 np.random.seed(1234)
 fk = FunctionalKernel(D=len(xss), lmc_kernels=ks, lmc_ranks=ranks)
 lmc_with_metrics = InterpolatedLLGP(
-    xss, yss, functional_kernel=fk, metrics=True)
+    xss, yss, functional_kernel=fk, metrics=True, max_procs=1)
 lmc_with_metrics.optimize(optimizer=AdaDelta(
     # Force full 35 iterations with ratio = 0
     verbosity=10, max_it=35, min_grad_ratio=0))
@@ -60,7 +59,7 @@ plt.xlabel('optimization iteration')
 plt.ylabel('inversion iterations')
 plt.legend(bbox_to_anchor=(.3, -0.15), loc=2)
 print('iterations.eps')
-plt.savefig(outdir + 'iterations.eps', format='eps', bbox_inches='tight')
+plt.savefig(outdir + 'iterations.pdf', format='pdf', bbox_inches='tight')
 plt.clf()
 
 fig, ax1 = plt.subplots()
@@ -76,6 +75,6 @@ ax2.plot(0.2 * div_rolled_max(lmc_with_metrics.metrics.grad_norms, 1),
 ax2.set_ylabel('grad norms', color='b')
 fig.tight_layout()
 plt.legend(bbox_to_anchor=(.3, -.07), loc=2)
-print('running_cutoff.eps')
-plt.savefig(outdir + 'running_cutoff.eps', format='eps', bbox_inches='tight')
+print('running_cutoff.pdf')
+plt.savefig(outdir + 'running_cutoff.pdf', format='pdf', bbox_inches='tight')
 plt.clf()

### 2017-10-09 benchmarks/pictures/pics.py

@@ -1,5 +1,8 @@
 # Runs 1-kernel LMC with a rank-2 RBF on fx2007
 # Compares to COGP and draws a diagram with each method's predictions.
+# Does same for SLFM model with m = 1000 on weather dataset.
+
+nthreads = '' # all cores
 
 import matplotlib as mpl
 mpl.use('Agg')
@@ -26,6 +29,8 @@ outdir = sys.argv[1] + '/'
 
 print('publishing results into out directory', outdir)
 
+print('FX2007 picture')
+
 # Nguyen 2014 COGP uses Q=2 R=1, but that is not LMC
 # Álvarez and Lawrence 2010 Convolved GP has R=4, sort of.
 # Álvarez and Lawrence 2010 find that vanilla LMC works best with Q=1 R=2
@@ -37,7 +42,7 @@ xss, yss, test_xss, test_yss, test_fx, cols = foreign_exchange_2007()
 
 np.random.seed(1234)
 fk = FunctionalKernel(D=len(xss), lmc_kernels=ks, lmc_ranks=ranks)
-lmc = InterpolatedLLGP(xss, yss, functional_kernel=fk)
+lmc = InterpolatedLLGP(xss, yss, functional_kernel=fk, max_procs=(nthreads if nthreads else None))
 opt = AdaDelta(verbosity=20, min_grad_ratio=0.2)
 print('training LLGP')
 with contexttimer.Timer() as t:
@@ -57,7 +62,7 @@ all_xs = np.arange(min(xs.min() for xs in xss), max(xs.max()
                                                     for xs in xss) + 1)
 test_ix = {col: list(cols).index(col) for col in test_fx}
 pred_xss = [all_xs if col in test_fx else np.array([]) for col in cols]
-lmc.prediction = 'exact'
+# lmc.prediction = 'exact'
 pred_yss, pred_vss = lmc.predict(pred_xss)
 pred_yss = {col: ys for col, ys in zip(cols, pred_yss)}
 pred_vss = {col: vs for col, vs in zip(cols, pred_vss)}
@@ -97,4 +102,76 @@ for col, ax in zip(test_fx, axs):
 
 print('fx2007graph.pdf')
 plt.savefig(outdir + 'fx2007graph.pdf', format='pdf', bbox_inches='tight')
+plt.clf()
+
+
+print('weather picture')
+
+test_fx = ['cam', 'chi']
[…]

### 2017-10-09 benchmarks/pictures/run.sh

@@ -1,8 +1,8 @@
 #!/bin/bash
-# Creates pictures from 1 run of fx2007 dataset example
+# Creates pictures from 1 run of fx2007 and weather dataset example
 
 USAGE="Usage: ./run.sh [--help]"
-EXPECTED_DIR="runlmc/benchmarks/picture-fx2007"
+EXPECTED_DIR="runlmc/benchmarks/pictures"
 HELP_STR="$USAGE
 
 Must be run from $EXPECTED_DIR.
@@ -13,8 +13,10 @@ the fx2007 paper and puts them into ./out.
 
 fx2007graph.pdf - graphical comparison of predictive mean+var of LLGP vs COGP
                   on the stocks being predicted.
-iterations.eps - number of iterations MINRES requires during course of optimization
-running_cutoff.eps - gradient norms (and cutoff) during course of optimization
+weather.pdf - graphical comparison of predictive mean+var of LLGP vs COGP
+                  on the weather being predicted
+iterations.pdf - number of iterations MINRES requires during course of optimization
+running_cutoff.pdf - gradient norms (and cutoff) during course of optimization
 
 stdout.txt - trace of execution stdout (this program)
 
@@ -61,6 +63,6 @@ REPOROOT=$(readlink -f "$PWD/../../../")
 
 cd $REPOROOT
 
-OMP_NUM_THREADS=1 PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/../$EXPECTED_DIR/fxpics.py $OUTFOLDER | tee $OUTFOLDER/stdout.txt
+OMP_NUM_THREADS=1 PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/../$EXPECTED_DIR/pics.py $OUTFOLDER | tee $OUTFOLDER/stdout.txt
 # metrics should use all threads (capped at 1 proc)
-PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/../$EXPECTED_DIR/fxmetrics.py $OUTFOLDER | tee -a $OUTFOLDER/stdout.txt
+OMP_NUM_THREADS=$(nproc) PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/../$EXPECTED_DIR/fxmetrics.py $OUTFOLDER | tee -a $OUTFOLDER/stdout.txt

### 2017-10-09 paper/copyresults.sh
updated copyresults script for new pdf pictures

@@ -7,13 +7,11 @@ function epsconvert {
     rm $1-unbounded.eps
 }
 
-echo 'iterations.eps -> iterations.pdf'
-cp ../benchmarks/picture-fx2007/out/iterations.eps .
-epsconvert iterations
+echo 'iterations.pdf'
+cp ../benchmarks/pictures/out/iterations.pdf .
 
-echo 'running_cutoff.eps -> running_cutoff.pdf'
-cp ../benchmarks/picture-fx2007/out/running_cutoff.eps .
-epsconvert running_cutoff
+echo 'running_cutoff.pdf'
+cp ../benchmarks/pictures/out/running_cutoff.pdf .
 
 echo 'relgrad_l2.eps -> relgrad_l2.pdf'
 cp ../benchmarks/grad-grid/out/relgrad_l2.eps .
@@ -28,7 +26,10 @@ cp ../benchmarks/grad-grid/out/relalpha_l2.eps .
 epsconvert relalpha_l2
 
 echo 'fx2007graph.pdf'
-cp ../benchmarks/picture-fx2007/out/fx2007graph.pdf .
+cp ../benchmarks/pictures/out/fx2007graph.pdf .
+
+echo 'weather.pdf'
+cp ../benchmarks/pictures/out/weather.pdf .
 
 echo 'representation-cmp/out/results.tex -> representation.tex'
 cp ../benchmarks/representation-cmp/out/results.tex representation.tex
@@ -38,3 +39,6 @@ cp ../benchmarks/fx2007-out/results_fx2007.tex .
 
 echo 'results_weather.tex'
 cp ../benchmarks/weather-out/results_weather.tex .
+
+echo 'results_synth.tex'
+cp ../benchmarks/synth/out/results_synth.tex .

### 2017-10-09 benchmarks/synth/run.sh
Added runfile for synth

@@ -1,0 +1,61 @@
+#!/bin/bash
+# Runs synthetic dataset example
+
+USAGE="Usage: ./run.sh [--help]"
+EXPECTED_DIR="runlmc/benchmarks/synth"
+HELP_STR="$USAGE
+
+Must be run from $EXPECTED_DIR.
+
+Runs only on local machine, but leverages all processessors avaiable in the 
+model evaluation phase.
+
+This runs the synth dataset for the SLFM model several times on several
+different interpolation point counts. Each setting has its timing, SMSE,
+and NLPD statistics outputted (mean and standard error).
+
+Then, we run COGP on the same dataset.
+
+Results are printed to a trace file, out/stdout-synth.txt
+Extracted results, printed as latex, out/results_synth.tex
+
+Flags
+    --help Print this help message.
+"
+
+base1=$(basename $PWD)
+base2=$(cd .. && basename $PWD)
+base3=$(cd .. && cd .. && basename $PWD)
+
+if [[ "$base3/$base2/$base1" != "$EXPECTED_DIR" ]]; then
+    echo "Must be run from $EXPECTED_DIR" >/dev/stderr
+    exit 1
+fi
+
+if [[ $# -gt 1 ]]; then
+    echo $USAGE >/dev/stderr
+    exit 1
+fi
+
+if [[ $# -eq 1 ]]; then
+    case $1 in
+        "--help")
+            printf "$HELP_STR"
+            exit 0
+            ;;
+        *)
+            echo $USAGE >/dev/stderr
+            exit 3
+            ;;
+    esac
+fi
+IS_VALIDATION=false
+
+REPOROOT=$(readlink -f "$PWD/../../../")
+cd out
+
+OUTFOLDER=$PWD
+
+cd $REPOROOT
+
+OMP_NUM_THREADS=1 PYTHONPATH="$REPOROOT:$REPOROOT/benchmarks/benchlib" python3 -u $REPOROOT/benchmarks/synth/synth.py $IS_VALIDATION $OUTFOLDER 2>&1 | tee $OUTFOLDER/stdout-synth.txt

### 2017-10-10 benchmarks/benchlib/cogp_weather.m
updated weather for COGP to use all of the dataset

@@ -17,11 +17,8 @@ chiy = csvread([datadir 'chiy.csv']);
 soty = csvread([datadir 'soty.csv']);
 
 x = csvread([datadir 'sotx.csv']);
-sel = x >= 10 & x <= 15;
 
-x = x(sel);
 y = [bray(:,4), camy(:,4), chiy(:,4), soty(:,4)];
-y = y(sel, :);
 y(y == -1) = nan; % missing data
 ytest = y;
 xtest = x;

### 2017-10-10 runlmc/models/interpolated_llgp.py
fix prediction-mode bug

@@ -384,7 +384,7 @@ class InterpolatedLLGP(MultiGP):
     def _var_predict_precompute(self, prediction_interpolants, _):
         nu = self._precomputed_nu()
         assert len(prediction_interpolants) == 1
-        W = next(prediction_interpolants.values())
+        W = list(prediction_interpolants.values()).pop()
         return W.dot(nu)
 
     def _var_predict_on_the_fly(self, _, Xs):

### 2017-10-10 benchmarks/synth/synth.py
added COGP grid search mode to synth.py

@@ -2,7 +2,7 @@
 
 import os
 os.environ['OMP_NUM_THREADS'] = '1'
-
+import sys
 import numpy as np
 from standard_tester import *
 import pickle
@@ -35,17 +35,24 @@ def main():
     inducing_points = [500]
     nbatches = [1000]
 
-    kgen, rgen, slfmgen, indepgen = synth_gen()
-    xss, yss, test_xss, test_yss = synth()
-    stats = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
-                test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
-                tolerance=1e-3) # reduce tol a bit b/c of large problem size
-    dump(stats, 'llgp_stats')
-    # stats = load('llgp_stats')
-
-    all_stats = [stats]
-    colnames = ['LLGP']
+    all_stats = []
+    colnames = []
     
+    if len(sys.argv) > 3: # COGP grid-search mode
+        nthreads = ''
+        inducing_points = [int(sys.argv[3])]
+        nbatches = [int(sys.argv[4])]
+    else:
+        kgen, rgen, slfmgen, indepgen = synth_gen()
+        xss, yss, test_xss, test_yss = synth()
+        stats = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
+                    test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
+                    tolerance=1e-3) # reduce tol a bit b/c of large problem size
+        dump(stats, 'llgp_stats')
+        # stats = load('llgp_stats')
+        all_stats = [stats]
+        colnames = ['LLGP']
+
     for ind, nb in zip(inducing_points, nbatches):
         cogp_stats = cogp_synth(
             cogp_runs, ind, nthreads, nb, cogp_max_it)

### 2017-10-10 benchmarks/synth/synth.py
cogp fine-tune mode x3

@@ -42,6 +42,8 @@ def main():
         nthreads = ''
         inducing_points = [int(sys.argv[3])]
         nbatches = [int(sys.argv[4])]
+        cogp_max_it = [int(sys.argv[5])]
+        cogp_runs = 1
     else:
         kgen, rgen, slfmgen, indepgen = synth_gen()
         xss, yss, test_xss, test_yss = synth()

### 2017-10-10 benchmarks/synth/synth.py
synth bug

@@ -42,7 +42,7 @@ def main():
         nthreads = ''
         inducing_points = [int(sys.argv[3])]
         nbatches = [int(sys.argv[4])]
-        cogp_max_it = [int(sys.argv[5])]
+        cogp_max_it = int(sys.argv[5])
         cogp_runs = 1
     else:
         kgen, rgen, slfmgen, indepgen = synth_gen()

### 2017-10-10 benchmarks/pictures/fxmetrics.py
updated pictures from benchmarks

@@ -58,7 +58,7 @@ plt.axhline(n, label='iterartion cutoff ($n={}$)'.format(n), c='r')
 plt.xlabel('optimization iteration')
 plt.ylabel('inversion iterations')
 plt.legend(bbox_to_anchor=(.3, -0.15), loc=2)
-print('iterations.eps')
+print('iterations.pdf')
 plt.savefig(outdir + 'iterations.pdf', format='pdf', bbox_inches='tight')
 plt.clf()
 

### 2017-10-10 benchmarks/synth/run.sh
re-ran synth.py

@@ -51,7 +51,7 @@ if [[ $# -eq 1 ]]; then
 fi
 IS_VALIDATION=false
 
-REPOROOT=$(readlink -f "$PWD/../../../")
+REPOROOT=$(readlink -f "$PWD/../../")
 cd out
 
 OUTFOLDER=$PWD

### 2017-10-10 benchmarks/synth/synth.py

@@ -14,12 +14,12 @@ import matplotlib.pyplot as plt
 
 
 def load(f):
-    with open('out/' + f + '.pkl', 'rb') as handle:
+    with open(sys.argv[2] + os.path.sep + f + '.pkl', 'rb') as handle:
         return pickle.load(handle)
 
 
 def dump(x, f):
-    with open('out/' + f + '.pkl', 'wb') as handle:
+    with open(sys.argv[2] + os.path.sep + f + '.pkl', 'wb') as handle:
         pickle.dump(x, handle, pickle.HIGHEST_PROTOCOL)
 
 
@@ -29,7 +29,7 @@ def main():
 
     llgp_runs = 3
     cogp_runs = 3
-    cogp_max_it = 100 # like in COGP sarcos
+    cogp_max_it = [1000]
     interpolating_points = [25, 25]
     nthreads = 16
     inducing_points = [500]
@@ -42,7 +42,7 @@ def main():
         nthreads = ''
         inducing_points = [int(sys.argv[3])]
         nbatches = [int(sys.argv[4])]
-        cogp_max_it = int(sys.argv[5])
+        cogp_max_it = [int(sys.argv[5])]
         cogp_runs = 1
     else:
         kgen, rgen, slfmgen, indepgen = synth_gen()
@@ -55,14 +55,14 @@ def main():
         all_stats = [stats]
         colnames = ['LLGP']
 
-    for ind, nb in zip(inducing_points, nbatches):
+    for ind, nb, mi in zip(inducing_points, nbatches, cogp_max_it):
         cogp_stats = cogp_synth(
-            cogp_runs, ind, nthreads, nb, cogp_max_it)
+            cogp_runs, ind, nthreads, nb, mi)
         dump(cogp_stats, 'cogp_stats-{}-{}'.format(ind, nb))
         # cogp_stats = load('cogp_stats-{}-{}'.format(ind, nb))   
         all_stats.append(cogp_stats)
-        colnames.append(r'\begin{tabular}{c}LLGP\\$m=' +
-                        str(ind) + ',n_b=' + str(nb) +
+        colnames.append(r'\begin{tabular}{c}COGP\\$m=' +
+                        str(ind) + '\\n_b=' + str(nb) +
                         r'$\end{tabular}')
 
     latex_table('results_synth.tex', colnames, all_stats)

### 2017-10-10 benchmarks/synth/synth.py
moved new results to paper/, fixed synth.tex result bug

@@ -37,8 +37,8 @@ def main():
 
     all_stats = []
     colnames = []
-    
-    if len(sys.argv) > 3: # COGP grid-search mode
+
+    if len(sys.argv) > 3:  # COGP grid-search mode
         nthreads = ''
         inducing_points = [int(sys.argv[3])]
         nbatches = [int(sys.argv[4])]
@@ -48,8 +48,8 @@ def main():
         kgen, rgen, slfmgen, indepgen = synth_gen()
         xss, yss, test_xss, test_yss = synth()
         stats = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
-                    test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
-                    tolerance=1e-3) # reduce tol a bit b/c of large problem size
+                             test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
+                             tolerance=1e-3)  # reduce tol a bit b/c of large problem size
         dump(stats, 'llgp_stats')
         # stats = load('llgp_stats')
         all_stats = [stats]
@@ -59,13 +59,14 @@ def main():
         cogp_stats = cogp_synth(
             cogp_runs, ind, nthreads, nb, mi)
         dump(cogp_stats, 'cogp_stats-{}-{}'.format(ind, nb))
-        # cogp_stats = load('cogp_stats-{}-{}'.format(ind, nb))   
+        # cogp_stats = load('cogp_stats-{}-{}'.format(ind, nb))
         all_stats.append(cogp_stats)
         colnames.append(r'\begin{tabular}{c}COGP\\$m=' +
-                        str(ind) + '\\n_b=' + str(nb) +
+                        str(ind) + r'$\\ $n_b=' + str(nb) +
                         r'$\end{tabular}')
 
     latex_table('results_synth.tex', colnames, all_stats)
+
 
 if __name__ == '__main__':
     main()

### 2017-10-10 benchmarks/benchlib/bench.py
fix bench.py for new StochasticDerivService api

@@ -211,7 +211,7 @@ def run_kernel_benchmark(
     n_it = 10
     metrics = None
     with closing(Pool(processes=cpu_count())) as pool:
-        sds = StochasticDerivService(metrics, pool, n_it)
+        sds = StochasticDerivService(metrics, pool, n_it, 1e-4)
         with contexttimer.Timer() as t:
             grid_kernel, _ = gen_grid_kernel(
                 fkern, grid_dists, interpolants, list(map(len, Xs)))

### 2017-10-12 benchmarks/benchlib/bench.py
parallel slurm wrapper

@@ -3,6 +3,9 @@
 
 # pylint: skip-file
 
+import os
+os.environ['OMP_NUM_THREADS'] = '1'
+
 import sys
 from contextlib import closing
 from multiprocessing import Pool, cpu_count
@@ -19,7 +22,6 @@ from runlmc.approx.iterative import Iterative
 from runlmc.kern.rbf import RBF
 from runlmc.kern.matern32 import Matern32
 from runlmc.kern.std_periodic import StdPeriodic
-from runlmc.models.interpolated_llgp import InterpolatedLLGP
 from runlmc.lmc.stochastic_deriv import StochasticDerivService
 from runlmc.lmc.functional_kernel import FunctionalKernel
 from runlmc.lmc.grid_kernel import *

### 2017-10-12 benchmarks/benchlib/slurm-wrapper.sh

@@ -1,6 +1,7 @@
 #!/bin/bash
 # Internal slurm wrapper when it's missing.
 # Only mimics slurm array-launching functionality.
+# parallelizes across all cores (so tasks should be 1-core only)
 
 if which sbatch >/dev/null 2>/dev/null; then
     jobid=$(sbatch --parsable "$@")
@@ -13,13 +14,9 @@ else
     array=$(grep "#SBATCH --array=" $script | head -1)
     array=$(echo $array | cut -c17- | tr '-' ' ')
     for i in $(seq $array); do
-        echo "Running array job $i"
         outfile="slurm-out-$i.txt"
         errfile="slurm-err-$i.txt"
-        SLURM_ARRAY_TASK_ID=$i $script "$@" >$outfile 2>$errfile
-    done
+        echo "echo Running array job $i && \
+              SLURM_ARRAY_TASK_ID=$i $script \"$@\" >$outfile 2>$errfile"
+    done | xargs --max-procs=$(nproc) --replace /bin/bash -c "{}"
 fi
-
-    
-        
-

### 2017-10-12 benchmarks/synth/synth.py
Ran cogp+ to longer time

@@ -28,11 +28,11 @@ def main():
     activate_logs()
 
     llgp_runs = 3
-    cogp_runs = 1
-    cogp_max_it = [100, 100]
+    cogp_runs = 3
+    cogp_max_it = [100, 1000]
     interpolating_points = [25, 25]
     nthreads = 16
-    inducing_points = [500, 5000]
+    inducing_points = [500, 500]
     nbatches = [1000, 1000]
     cogp_colnames = ['COGP', 'COGP+']
 

### 2017-10-12 paper/paper.tex

@@
 \begin{document}
+
+\ifanonymized
+
 \twocolumn[
-
 \aistatstitle{Large Linear Multi-output Gaussian Process Learning}
-
-\aistatsauthor{   Vladimir Feinberg \And Li-Fang Cheng \And  {Kai Li} \And {Barbara E Engelhardt} }
-
+\aistatsauthor{ Anonymous Author }
+\aistatsaddress{ Anonymous Institution }
+]
+\else
+\twocolumn[
+\aistatstitle{Large Linear Multi-output Gaussian Process Learning}
+\aistatsauthor{   {Vladimir Feinberg} \And {Li-Fang Cheng} \And  {Kai Li} \And {Barbara E Engelhardt} }
 \aistatsaddress{ UC Berkeley \And Princeton University \And Princeton University \And Princeton University }
 ]
+\fi
 
 \begin{abstract}
+
 Gaussian processes, or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \citep{alvarez2012kernels}.
@@
 We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and low-dimensional inputs.
-When applied to real time series and temporal-spatial data, we find our theoretical improvement relative to the current state of the art is realized with LLGP reducing training time while improving or maintaining predictive accuracy.
+When applied to synthetic two-dimensional input and real time series data, we find our theoretical improvement relative to the current state of the art is realized with LLGP reducing training time while improving or maintaining predictive accuracy.
+By using a direct likelihood approximation rather than a variational one predictive variance estimates are significantly improved.
 \end{abstract}
@@
 The pairwise covariances for a stationary kernel in this subgrid also exhibit Toeplitz structure, since we still have $(x_1-x_2,y_i-y_j)=(x_1-x_2,y_i+\Delta_y-y_{j}-\Delta_y)=(x_1-x_2,y_{i+1}-y_{j+1})$.
-Ordering points in lexicographic order, so that the covariance matrix $K$ has $n_x^2$ order-$n_y$ blocks $K_{x_ix_j}$ with pairwise covariances between varying $y$ values between a pair of fixed $x$ values, the previous sentence implies $K_{x_ix_j}$ are Toeplitz.
+Ordering points in lexicographic order, so that the covariance matrix $K$ has $n_x^2$ order-$n_y$ blocks $K_{x_ix_j}$ with pairwise covariances between varying $y$ values between a pair of fixed $x$ values, the previous sentence implies $\{K_{x_ix_j}\}_{ij}$ are Toeplitz.
 By similar reasoning, since for any $y_1,y_2$, $(x_i-x_j,y_1-y_j)=(x_{i+1}-x_{j+1},y_1-y_2)$, we have $K_{x_ix_j}=K_{x_{i+1}x_{j+1}}$, thus the block structure of $K$ is itself Toeplitz, and hence $K$ is BTTB.
@@
 BTTB matrices themselves admit $O(n\log n)$ runtime for MVMs, with $n$ being the total grid size, or $n_xn_y$ in the two dimensional case, and if they are symmetric they can be represented with only their top row as well.
-The MVM runtime constant scales exponentially in the input dimension, however, so this approach is only applicable for low-input-dimension problems.
+The MVM runtime constant scales exponentially in the input dimension, however, so this approach is only applicable to low-input-dimension problems.
 
@@
 Moreover, SLFM and COGP models include an independent GP for each output, represented in LMC as additional kernels $\idx{k_d}{d=1}{D}$, where $A_d=0$ and $\bsk_d=\Te_d\in\R^D$.
-COGP uses its shared structure to derive efficient expressions for variational inference for parameter estimation.
+COGP uses its shared structure to derive efficient expressions for variational inference (VI) for parameter estimation.
 
@@
 
-\begin{figure}[!h]
+\begin{figure}[!ht]
 \begin{center}
@@
 \begin{algorithm}[!ht]
-  \caption{Compute an approximation of $\nabla L$.
+  \caption{Compute an approximation of $\nabla \mcL$.
 Assume \textsc{minres} is the inversion routine.
@@
 
-% Sum is pretty useless in practice,
-% but a useful baseline for comparison between representations in the supplement.
-% move the explanation of sum representation to supplement, then
-% \subsubsection{Sum representation}
-% 
-% In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
+% Can probably move this subsection to supplement if necessary
+\subsubsection{Sum representation}
+
+In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
 At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a BTTB matrix of order $m$, represented using only the top row.
@@
 
-\begin{table*}[!h]
+\begin{table*}[!ht]
   \caption{
@@
 
+% TODO rerun graphs with axis labels
 \section{Results}
@@
 In both cases, lower values represent better performance.
-We evaluated the performance of our representations of the kernel, \textsc{bt} and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm over a variety of different kernels (see the supplement).\footnote{Hyperparameters, data, code, and benchmarking scripts are available in
+We evaluated the performance of our representations of the kernel, \textsc{sum}, \textsc{bt}, and \textsc{slfm}, by computing exact gradients using the standard Cholesky algorithm over a variety of different kernels (see the supplement).\footnote{Hyperparameters, data, code, and benchmarking scripts are available in
   \ifanonymized
@@
 }
+
+Predictive performance on held-out data with SMSE and NLPD offers an apples-to-apples comparison with COGP, which was itself proposed on the basis of these two metrics.
+Training data log likelihood would unfairly favor LLGP, which optimizes $\mcL$ directly.
+Note that predicting each output's mean results in a baseline SMSE of $1$.
+
+Finally, we attempted to run hyperparameter optimization with both an exact GP and a variational DTC approximation as provided by the GPy package, but both runtime and predictive performance were already an order of magnitude worse than both LLGP and COGP on our smallest dataset from Sec.~\ref{fx2007-results} \citep{gpy2014}.
+
+\subsection{Synthetic dataset}\label{synth-bench}
+
+First, we evaluate raw learning performance on a synthetic dataset.
+We fix the GP index as $\R^2$ and generate a fixed SLFM RBF kernel with $Q=2$, fixed lengthscales, and covariance hyperparameters $\textbf{a}_1,\textbf{a}_2$.
+We set the output dimension to be $D=5$, so this synthetic dataset might resemble a geospatial GP model for subterranean mineral density, where the various minerals would be different outputs.
+Sampling $n\approx 50000$ GP values from the unit square, we hold out approximately $2500$ of them, corresponding to the values for the final output in the upper-right quadrant of the sampling square.
+
+We consider the problem of estimating the pre-fixed GP hyperparameters (starting from randomly-initialized ones) with LLGP and COGP.
+We evaluate the fit based on imputation performance on the held-out values (Tbl.~\ref{tbl:synth}).
+For COGP, we use hyperparameter settings applied to the Sarcos dataset from the COGP paper, a dataset of approximately the same size, which have the number of inducing points $m=500$.
+However, COGP failed to have above-baseline SMSE performance with learned inducing points, even on a range of $m$ up to $5000$ and various learning rates.
+Using fixed inducing points allowed for moderate improvement in COGP, which we used for comparison.
+For LLGP, we use a grid of size $m=25\times 25=625$ on the unit square with no learning rate tuning.
+LLGP was able to estimate more predictive hyperparameter values in about the same amount of time it took COGP to learn significantly worse values in terms of prediction.
+
+\begin{table}[!ht]
+\caption{Predictive Performance versus Training Time Tradeoffs on the Synthetic Dataset.
+We evaluate the learned LLGP model with $m=625$.
+COGP was evaluated with $m=500$, which were used on a similar-sized dataset from the COGP paper, and increasing $m$ did not improve performance.
+Since COGP does not provide a terminating condition for its optimization, we also show its performance when permitted to train longer, labelled COGP+.
+All trials were run 3 times, with parenthesized values representing standard error shown below.}
+\label{tbl:synth}
+\begin{center}
+  \begin{small}
+    \begin{sc}
+\input{results_synth.tex}
+\end{sc}
+\end{small}
+\end{center}
+
+\end{table}
 
@@
 The test points include 50 contiguous points extracted from each of the CAD, JPY, and AUD exchanges.
-For this application, LLGP uses $m=n/D=238$ interpolating points.
+For this application, LLGP uses $m=n/D=234$ interpolating points.
 We used the COGP settings from the paper.
@@
 
-\begin{table}[!h]
+\begin{table}[!ht]
   \caption{Average Predictive Performance and Training Time Over $10$ Runs for LLGP and COGP on the FX2007 Dataset.
 Parenthesized values are standard error.
-LLGP was run with LMC set to $Q=1$, $R=2$, and $238$ interpolating points.
+LLGP was run with LMC set to $Q=1$, $R=2$, and $234$ interpolating points.
 COGP used a $Q=2$ kernel with $100$ inducing points.}
@@
 
-\begin{figure*}[!h]
+\begin{figure*}[!ht]
 \centering
@@
 
-We use the default COGP parameters.\footnote{\url{https://github.com/trungngv/cogp}} We tested LLGP models on from $500$ to $1000$ interpolating points.
+We use the default COGP parameters.\footnote{\url{https://github.com/trungngv/cogp}} We tested LLGP models on $500$ and $1000$ interpolating points.
 
-\begin{table}[!h]
+\begin{table}[!ht]
   \caption{Average Predictive Performance and Training Time Over $10$ Runs of LLGP and COGP on the Weather Dataset.
@@
 
-LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tab.~\ref{weather-tbl}).
-Varying the number of interpolating points $m$ from $500$ to $1000$ demonstrates the runtime versus NLPD tradeoff (Fig.~\ref{fig:llgpweather}).
+LLGP performed slightly worse than COGP in SMSE, but both NLPD and runtime indicate significant improvements (Tab.~\ref{weather-tbl}, Fig.~\ref{weather-fig}).
+Varying the number of interpolating points $m$ from $500$ to $1000$ demonstrates the runtime versus NLPD tradeoff.
 While NLPD improvement diminishes as $m$ increases, LLGP still improves upon COGP for a wide range of $m$ by an order of magnitude in runtime and almost two orders of magnitude in NLPD.
 
-\begin{figure}[!h]
+\begin{figure*}[!ht]
 \centering
-{\includegraphics[width=0.8\columnwidth]{m_time_nlpd.pdf}}
-\caption{Average and standard error of NLPD and runtime of the SLFM model on LLGP across over varying interpolating points. Every setting was run $10$ times.}
-\label{fig:llgpweather}
-\end{figure}
-
-\subsection{NOAA dataset}\label{noaa-bench}
-
-TODO. NOAA dataset intro. Mention use of split kernel, what that is (Tbl.~\ref{noaa-tbl}). Also TODO: current numbers are all on Princeton machines.
-
-\begin{table}[!h]
-  \caption{Average Predictive Performance and Training Time Over $10$ Runs of LLGP and COGP (GPy?) on the NOAA Dataset.
-Parenthesized values are standard error.
-Both LLGP and COGP (GPy?) trained the SLFM model.}
-\label{noaa-tbl}
-\begin{center}
-  \begin{small}
-    \begin{sc}
-\begin{tabular}{lcc}\toprule
-Metric & LLGP & COGP (GPy?)\\
-\midrule
-seconds & $\textbf{0.00}$ ($\textbf{0.00}$) &  $\infty$ ($0.00$) \\
-SMSE & $\textbf{0.00}$ ($\textbf{0.00}$) & $\infty$ ($0.00$) \\
-NLPD & $\textbf{0.00}$ ($\textbf{0.00}$) & $\infty$ ($0.00$) \\
-\bottomrule
-\end{tabular}
-\end{sc}
-\end{small}
-\end{center}
-
-\end{table}
+{\includegraphics[width=\textwidth]{weather.pdf}}
+\caption{Test outputs for the Weather dataset.
+COGP mean is black, with 95\% confidence intervals shaded in grey.
+LLGP mean is a solid red curve, with light green 95\% confidence intervals.
+Magenta points are in the training set, while blue ones are in the test set.
+Like Fig.~\ref{fx2007-graph}, the training run was not cherry-picked.}
+\label{weather-fig}
+\end{figure*}
 
@@
 We have shown on several datasets that this can be done in a way that is faster and leads to more accurate results than variational approximations.
-Because LLGP scales well with increases in $m$, capturing complex interactions in the covariance with an accurate interpolation is cheap, as demonstrated by performance on both the FX2007 and weather datasets (Fig.~\ref{fx2007-graph}, Tab.~\ref{fx2007-tbl}, Tab.~\ref{weather-tbl}).
+Because LLGP scales well with increases in $m$, capturing complex interactions in the covariance with an accurate interpolation is cheap, as demonstrated by performance on a variety of datasets (Tab.~\ref{tbl:synth}, Tab.~\ref{fx2007-tbl}, Tab.~\ref{weather-tbl}).
 
@@
 \subsubsection*{Acknowledgments}
-(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof. Andrew Wilson.
+(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof. Andrew Wilson, Prof. John Cunningham.
 \fi

### 2017-10-12 benchmarks/pictures/pics.py
updated pictures with axes labels

@@ -67,7 +67,10 @@ pred_yss = {col: ys for col, ys in zip(cols, pred_yss)}
 pred_vss = {col: vs for col, vs in zip(cols, pred_vss)}
 
 _, axs = plt.subplots(ncols=3, figsize=(16, 4))
-for col, ax in zip(test_fx, axs):
+for i, (col, ax) in enumerate(zip(test_fx, axs)):
+    if i == 0:
+        ax.set_ylabel('USD : foreign currency rate')
+    ax.set_xlabel('time (days)')
 
     # Prediction on entire domain for COGP
     ax.plot(all_xs, cogp_mu[col], c='black', ls='-')
@@ -140,7 +143,10 @@ pred_yss = {col: ys for col, ys in zip(cols, pred_yss)}
 pred_vss = {col: vs for col, vs in zip(cols, pred_vss)}
 
 _, axs = plt.subplots(ncols=2, figsize=(16, 4))
-for col, ax in zip(test_fx, axs):
+for i, (col, ax) in enumerate(zip(test_fx, axs)):
+    if i == 0:
+        ax.set_ylabel('temperature (celsius)')
+    ax.set_xlabel('time (days)')
     # Prediction on entire domain for COGP
     ax.plot(all_xs[sel], cogp_mu[col].values[sel], c='black', ls='-')
     sd = np.sqrt(cogp_var[col].values[sel])

### 2017-10-12 paper/supplement.tex
Float Barrier in Supplement
Made paper arxiv-compatible

@@
 \documentclass{article}
+\usepackage{aistats2018}
+% Camera ready: \usepackage[accepted]{aistats2018}
+\newif\ifanonymized
+\anonymizedtrue
+% Camera ready: \anonymizedfalse
 
-\PassOptionsToPackage{numbers, compress}{natbib}
-% Camera ready: \usepackage[final]{nips_2017}
-\usepackage{nips_2017}
-
+\usepackage[round]{natbib}
 \usepackage[utf8]{inputenc} % allow utf-8 input
 \usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{hyperref}       % hyperlinks
+\usepackage{lmodern}
+\usepackage[hidelinks]{hyperref}       % hyperlinks
 \usepackage{url}            % simple URL typesetting
@@
 \usepackage{booktabs}
-\usepackage[section]{placeins}
 
@@
 
-\title{Supplementary Material for Large Linear Multi-output Gaussian Process Learning for Time Series}
-
-\author{
-  Vladimir Feinberg, Li-Fang Cheng, {Barbara Engelhardt}, {Kai Li}\\
-  Department of Computer Science\\
-  Princeton University\\
-  Princeton, NJ 08544 \\
-  \texttt{\{vyf,lifangc,bee,li\}@princeton.edu} \\
-}
-
 \begin{document}
 
-\maketitle
+\ifanonymized
+
+\twocolumn[
+\aistatstitle{Supplementary Material for Large Linear Multi-output Gaussian Process Learning}
+\aistatsauthor{ Anonymous Author }
+\aistatsaddress{ Anonymous Institution }
+]
+\else
+\twocolumn[
+\aistatstitle{Supplementary Material for Large Linear Multi-output Gaussian Process Learning}
+\aistatsauthor{   {Vladimir Feinberg} \And {Li-Fang Cheng} \And  {Kai Li} \And {Barbara E Engelhardt} }
+\aistatsaddress{ UC Berkeley \And Princeton University \And Princeton University \And Princeton University }
+]
+\fi
 
@@
 
-LLGP was implemented in Python 3 from the Anaconda, which offered an Intel MKL-linked scipy \cite{scipy}.
-The code made heavy use of other packages, namely climin \cite{climin}, GPy \cite{gpy}, and paramz \cite{paramz}.
-Code and benchmarks are available at \texttt{<anonymous repository>}.
+LLGP was implemented in Python 3 from the Anaconda, which offered an Intel MKL-linked scipy \citep{scipy}.
+The code made heavy use of other packages, namely climin \citep{climin}, GPy \citep{gpy}, and paramz \citep{paramz}.
+Code and benchmarks are available at
+\ifanonymized
+\texttt{<anonymous repository>}.
+\else
+\url{https://github.com/vlad17/runlmc}.
+\fi
 
-Application of our approach to all replication studies was carried out on a large server in a multiprogramming environment: CentOS 6.5 with 80 Intel(R) Xeon(R) CPU E7-4870 @ 2.40GHz.
-The representation evaluation benchmarks were done at once on a cluster of machines running CentOS 5.2-5.9 with Intel(R) Xeon(R) CPU E5430 @ 2.66GHz, where these jobs ran on a single thread per CPU.
+Application of our approach to all replication studies was carried out on a large server in a multiprogramming environment: Ubuntu 16.04.3 LTS with 48 Intel(R) Xeon(R) CPU E5-2670 v3 @ 2.30GHz.
+All experiments in the main report were allotted up to 16 threads (both for COGP MATLAB computation threads and parallel Monte Carlo in the case of LLGP).
+All experiments in the supplement (below, Sec.~\ref{empirical-rep}) were allotted only 1 thread each.
 
@@
 Next, we construct a random LMC kernel by sampling entries of each $A_q$ from a standard normal distribution truncated to the unit interval, $\bsk_q$ from an inverse gamma with unit shape and scale, and independent noise $\bse$ for every output from an inverse gamma with unit scale and mean $\epsilon$.
-Inputs and outputs were independently generated from $\Unif[0,1]$ for benchmarking.
+Inputs and outputs were independently generated from $\Unif[0,1]$ for representation evaluation benchmarking.
 
@@
 For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
-\begin{table}[!h]
+\begin{table*}[!ht]
   \caption{The runtime in seconds for solving $K\Tx=\Ty$ for a random kernel $K$ constructed as in Sec.~\ref{empirical-rep} using \textsc{minres} for each of the kernel representations.
@@
 \end{center}
-\end{table}
+\end{table*}
 
@@
 The relative error in the gradients is generally low for smooth kernels that induce diagonally dominant covariance matrices (Fig.~\ref{fig:relgrad}).
-Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions (Fig.~\ref{fig:rellog}.
+Kernels such as the single Mat\'{e}rn or periodic kernel with noise on the order of $10^{-4}$ lead to less accurate gradients, owing to poor \textsc{minres} convergence in the inversions (Fig.~\ref{fig:rellog}).
 We showed in the main paper that the stochastic gradients suffice for optimization in practical examples.
 
-\begin{figure}[!h]
+\begin{figure*}[!ht]
   \centering
-{\includegraphics[width=\textwidth]{relgrad_l2.eps}}
+{\includegraphics[width=\textwidth]{relgrad_l2.pdf}}
 \caption{Negative logarithm of the relative error from the LLGP gradient construction to the exact log likelihood gradient.
@@
 \label{fig:relgrad}
-\end{figure}
+\end{figure*}
 
-\begin{figure}[!h]
+\begin{figure*}[!ht]
 \centering
-{\includegraphics[width=\textwidth]{relalpha_l2.eps}}
+{\includegraphics[width=\textwidth]{relalpha_l2.pdf}}
 \caption{Negative logarithm of the relative error in $K^{-1}\Ty$, using \textsc{minres} compared to the Cholesky solution.
@@
 \label{fig:rellog}
-\end{figure}
+\end{figure*}
 
 \bibliographystyle{unsrtnat}
-\bibliography{paper}
+\bibliography{supplement}
 

### 2017-10-13 benchmarks/pictures/pics.py
updated pictures to include legend

@@ -73,34 +73,39 @@ for i, (col, ax) in enumerate(zip(test_fx, axs)):
     ax.set_xlabel('time (days)')
 
     # Prediction on entire domain for COGP
-    ax.plot(all_xs, cogp_mu[col], c='black', ls='-')
+    ax.plot(all_xs, cogp_mu[col], c='black', ls='-', label='COGP mean')
     sd = np.sqrt(cogp_var[col])
     top = cogp_mu[col] + 2 * sd
     bot = cogp_mu[col] - 2 * sd
-    ax.fill_between(all_xs, bot, top, facecolor='grey', alpha=0.2)
+    ax.fill_between(all_xs, bot, top, facecolor='grey', alpha=0.2, label='COGP CI')
 
     # Prediction for LLGP
-    ax.plot(all_xs, pred_yss[col], c='red')
+    ax.plot(all_xs, pred_yss[col], c='red', label='LLGP mean')
     sd = np.sqrt(pred_vss[col])
     top = pred_yss[col] + 2 * sd
     bot = pred_yss[col] - 2 * sd
-    ax.fill_between(all_xs, bot, top, facecolor='green', alpha=0.3)
+    ax.fill_between(all_xs, bot, top, facecolor='green', alpha=0.3, label='LLGP CI')
 
     # Actual holdout
     marker_size = 5
     test_xs = test_xss[test_ix[col]]
     test_ys = test_yss[test_ix[col]]
     ax.scatter(test_xs, test_ys, c='blue',
-               edgecolors='none', s=marker_size, zorder=11)
+               edgecolors='none', s=marker_size, zorder=11, label='holdout')
 
     # Rest of image (training)
     rest_xs = xss[test_ix[col]]
     rest_ys = yss[test_ix[col]]
     ax.scatter(rest_xs, rest_ys, c='magenta',
-               edgecolors='none', s=marker_size, zorder=10)
+               edgecolors='none', s=marker_size, zorder=10, label='training')
 
     ax.set_xlim([0, 250])
     ax.set_title('output {} (95%)'.format(col))
+
+    h, l = ax.get_legend_handles_labels()
+
+plt.figlegend(h, l, loc="upper center", borderaxespad=0., ncol=6)
+    
 
 print('fx2007graph.pdf')
 plt.savefig(outdir + 'fx2007graph.pdf', format='pdf', bbox_inches='tight')
@@ -148,31 +153,34 @@ for i, (col, ax) in enumerate(zip(test_fx, axs)):
[…]

### 2017-10-13 paper/paper.tex
updated paper

@@
 \usepackage{wrapfig}
+\usepackage{color}
+
+\usepackage{afterpage}
 
@@
 
-Gaussian processes, or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \citep{alvarez2012kernels}.
+Gaussian processes (GPs), or distributions over arbitrary functions in a continuous domain, can be generalized to the multi-output case: a linear model of coregionalization (LMC) is one approach \citep{alvarez2012kernels}.
 LMCs estimate and exploit correlations across the multiple outputs.
 While model estimation can be performed efficiently for single-output GPs \citep{msgp}, these assume stationarity, but in the multi-output case the cross-covariance interaction is not stationary.
-We propose Large Linear GPs (LLGPs), which circumvent the need for stationarity by using LMC's structure, enabling optimization of GP hyperparameters for multi-dimensional outputs and low-dimensional inputs.
-When applied to synthetic two-dimensional input and real time series data, we find our theoretical improvement relative to the current state of the art is realized with LLGP reducing training time while improving or maintaining predictive accuracy.
-By using a direct likelihood approximation rather than a variational one predictive variance estimates are significantly improved.
+We propose Large Linear GP (LLGP), which circumvents the need for stationarity by inducing structure in the LMC kernel through a common grid of inputs shared between outputs, enabling optimization of GP hyperparameters for multi-dimensional outputs and low-dimensional inputs.
+When applied to synthetic two-dimensional and real time series data, we find our theoretical improvement relative to the current solutions for multi-output GPs is realized with LLGP reducing training time while improving or maintaining predictive mean accuracy.
+Moreover, by using a direct likelihood approximation rather than a variational one, model confidence estimates are significantly improved.
 \end{abstract}
@@
 
-Gaussian processes (GPs) are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \citep{williams1996gaussian}.
+GPs are a nonlinear regression method that capture function smoothness across inputs through a response covariance function \citep{williams1996gaussian}.
 GPs extend to multi-output regression, where the objective is to build a probabilistic regression model over vector-valued observations by identifying latent cross-output processes.
@@
 The polynomial dependence on $m$ is still cubic: the matrix inversion underlying the state-of-the-art multi-output GP estimation method ignores LMC's structure.
-Here, we exploit this structure to avoid direct matrix inversion.
+Namely, the cross-covariance between two outputs is determined by a linear combination of stationary subkernels.
+On a grid of inputs, each subkernel induces matrix structure, so viewing the LMC kernel as a linear combination of structured matrices we can avoid direct matrix inversion.
 
@@
 
-A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian \citep{williams1996gaussian}.
-With zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$.
+A GP is a set of random variables (RVs) $\{y_\Tx\}_\Tx$ indexed by $\Tx\in\mcX$, with the property that, for any finite collection $X=\idx{\Tx_i}{i=1}{n}$ of $\mcX$, the RVs are jointly Gaussian with zero mean without loss of generality and a prespecified covariance $K:\mcX^2\rightarrow\R$, $\Ty_X\sim N\pa{\bsz, K_{X, X}}$, where $(\Ty_X)_i=y_{\Tx_i}$ and $(K_{X,X})_{ij}=K(\Tx_i,\Tx_j)$ \citep{williams1996gaussian}.
 Given observations of $\Ty_X$, inference at a single point $*\in\mcX$ of an $\R$-valued RV $y_*$ is performed by conditioning $y_*|\Ty_X$ \citep{williams1996gaussian}.
@@
 Ordering points in lexicographic order, so that the covariance matrix $K$ has $n_x^2$ order-$n_y$ blocks $K_{x_ix_j}$ with pairwise covariances between varying $y$ values between a pair of fixed $x$ values, the previous sentence implies $\{K_{x_ix_j}\}_{ij}$ are Toeplitz.
-By similar reasoning, since for any $y_1,y_2$, $(x_i-x_j,y_1-y_j)=(x_{i+1}-x_{j+1},y_1-y_2)$, we have $K_{x_ix_j}=K_{x_{i+1}x_{j+1}}$, thus the block structure of $K$ is itself Toeplitz, and hence $K$ is BTTB.
+By similar reasoning, since for any $y_1,y_2$, $(x_i-x_j,y_1-y_2)=(x_{i+1}-x_{j+1},y_1-y_2)$, we have $K_{x_ix_j}=K_{x_{i+1}x_{j+1}}$, thus the block structure of $K$ is itself Toeplitz, and hence $K$ is BTTB (Fig.~\ref{fig:bttb}).
 For higher dimensions, one can imagine a recursive block-Toeplitz structure.
@@
 The MVM runtime constant scales exponentially in the input dimension, however, so this approach is only applicable to low-input-dimension problems.
+
+\begin{figure}[!ht]
+\begin{center}
+$$
+\begin{array}{c}
+K_{i\Delta_x\times Y,j\Delta_x\times Y}=\mat{{\color{red} k\pa{\abs{i-j}\Delta_x} }&{\color{blue}\ddots} \\
+{\color{blue}\ddots} &{ \color{red} k\pa{\abs{i-j}\Delta_x}} } \\
+K_U=\mat{
+ {\color{red} K_{0\times Y} } & {\color{blue} K_{0\times Y,\Delta_x\times Y}}& {\color{green}\ddots }\\
+{\color{blue} K_{\Delta_x\times Y, 0\times Y}} & {\color{red} K_{\Delta_x\times Y}} & {\color{blue} K_{0\times Y, \Delta_x\times Y}} \\
+{\color{green} \ddots } & {\color{blue}  K_{2\Delta_x\times Y, \Delta_x\times Y}} & {\color{red} K_{\Delta_x\times Y}}
+}
+\end{array}
+$$
+\end{center}
+\caption{An illustration of a stationary kernel $K((a, b), (c, d))=k\pa{\norm{(a-c, b-d)}}$ evaluated on a two-dimensional input grid $U=X\times Y =\{x_i\}\times \{y_j\}$ with separations $\Delta_x,\Delta_y$, resulting in BTTB structure (here, with $\card{X}=3$).
+Identical matrices are colored the same.
+We use the shorthand $K_Z=K_{Z,Z}$.}
+\label{fig:bttb}
+\end{figure}
 
@@
 This approach is agnostic to kernel stationarity, so one may use inducing points for all outputs $T'\subset \mcX'$, with the model equal to the exact GP model when $T'={X'}$ \citep{alvarez2010efficient}.
-Computationally, these approaches resemble making rank-$m$ approximations to the $n\times n$ covariance matrix.
+Computationally, these approaches resemble making rank-$\card{T}$ approximations to the $n\times n$ covariance matrix.
 
@@
 The simultaneous interpolation $W_{X,U}\in\R^{n\times m}$ then yields the SKI approximation: $K_{X,X}\approx W_{X,U}K_{U,U}W_{U,X}^\top$.
-$W$ has only $4^dn$ nonzero entries, with $\mcX=\R^d$.
+$W$ has only $4^Pn$ nonzero entries, with $\mcX=\R^P$.
 Even without relying on structure, SKI reduces the representation of $K_{X,X}$ to an $m$-rank matrix.
@@
 
-\begin{figure}[!ht]
+\begin{figure*}[!ht]
 \begin{center}
-  \subfigure[]{\includegraphics[width=0.4\columnwidth]{iterations.pdf}}
-  \subfigure[]{\includegraphics[width=0.4\columnwidth]{running_cutoff.pdf}}
+  \subfigure[]{\includegraphics[width=0.4\textwidth]{iterations.pdf}}
+  \subfigure[]{\includegraphics[width=0.4\textwidth]{running_cutoff.pdf}}
 \end{center}
@@
 \label{fx2007-iterations}
-\end{figure}
+\end{figure*}
 
@@
 
-% TODO rerun graphs with axis labels
 \section{Results}
@@
 Training data log likelihood would unfairly favor LLGP, which optimizes $\mcL$ directly.
-Note that predicting each output's mean results in a baseline SMSE of $1$.
+Note that predicting a constant value equal to each output's holdout mean results in a baseline SMSE of $1$.
 
@@
 We evaluate the fit based on imputation performance on the held-out values (Tbl.~\ref{tbl:synth}).
-For COGP, we use hyperparameter settings applied to the Sarcos dataset from the COGP paper, a dataset of approximately the same size, which have the number of inducing points $m=500$.
+For COGP, we use hyperparameter settings applied to the Sarcos dataset from the COGP paper, a dataset of approximately the same size, which has the number of inducing points $m=500$.
 However, COGP failed to have above-baseline SMSE performance with learned inducing points, even on a range of $m$ up to $5000$ and various learning rates.
@@
 {\includegraphics[width=\textwidth]{fx2007graph.pdf}}
+% TODO add legend
 \caption{Test outputs for the FX2007 dataset.
@@
 
+\begin{figure*}[!ht]
+\centering
+{\includegraphics[width=\textwidth]{weather.pdf}}
+\caption{Test outputs for the Weather dataset.
+COGP mean is black, with 95\% confidence intervals shaded in grey.
+LLGP mean is a solid red curve, with light green 95\% confidence intervals.
+Magenta points are in the training set, while blue ones are in the test set.
+Like Fig.~\ref{fx2007-graph}, the training run was not cherry-picked.}
+\label{weather-fig}
+\end{figure*}
+
 Next, we replicate results from a weather dataset, a large time series used to validate COGP.
@@
 
-\begin{table}[!ht]
+\begin{table}[!htbp]
   \caption{Average Predictive Performance and Training Time Over $10$ Runs of LLGP and COGP on the Weather Dataset.
@@
 
-\begin{figure*}[!ht]
-\centering
-{\includegraphics[width=\textwidth]{weather.pdf}}
-\caption{Test outputs for the Weather dataset.
-COGP mean is black, with 95\% confidence intervals shaded in grey.
-LLGP mean is a solid red curve, with light green 95\% confidence intervals.
-Magenta points are in the training set, while blue ones are in the test set.
-Like Fig.~\ref{fx2007-graph}, the training run was not cherry-picked.}
-\label{weather-fig}
-\end{figure*}
-
 \section{Conclusion}\label{conclusion}
@@
 
+\pagebreak
 \ifanonymized
@@
 \subsubsection*{Acknowledgments}
-(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof. Andrew Wilson, Prof. John Cunningham.
+(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof.Ion Stoica, Prof. Andrew Wilson, Prof. John Cunningham.
 \fi

### 2017-10-13 paper/paper.tex
a minor paper bug

@@
  {\color{red} K_{0\times Y} } & {\color{blue} K_{0\times Y,\Delta_x\times Y}}& {\color{green}\ddots }\\
-{\color{blue} K_{\Delta_x\times Y, 0\times Y}} & {\color{red} K_{\Delta_x\times Y}} & {\color{blue} K_{0\times Y, \Delta_x\times Y}} \\
-{\color{green} \ddots } & {\color{blue}  K_{2\Delta_x\times Y, \Delta_x\times Y}} & {\color{red} K_{\Delta_x\times Y}}
+{\color{blue} K_{\Delta_x\times Y, 0\times Y}} & {\color{red} K_{\Delta_x\times Y}} & {\color{blue} K_{\Delta_x\times Y, 2\Delta_x\times Y}} \\
+{\color{green} \ddots } & {\color{blue}  K_{2\Delta_x\times Y, \Delta_x\times Y}} & {\color{red} K_{2\Delta_x\times Y}}
 }

### 2017-11-26 paper/paper.tex
added aistats review comments (pt 1)

@@
 \fi
+
+% TODO: drop to 8 pages
+% TODO: add cov matrix to supplement
+% TODO: add particular IGP, CGP numbers?
 
@@
 In practice, where we wish to capture covariance across outputs as a $D^2$-dimensional latent process, $B_q$ is not Toeplitz, so $K([i,\Tx],[j,\Tz])\neq K([i+1,\Tx],[j+1,\Tz])$.
+Thus, LLGP allows for arbitrary PSD $B_q$.
 
@@
 Since the difference in adjacent inputs $t_{i+1}-t_{i}$ is the same for all $i$, we have the Toeplitz property that $T_{(i+1)(j+1)}=k(\abs{t_{i+1} -t_{j+1}})=k(\abs{t_i-t_j})=T_{ij}$.
-Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, in $O(n\log n)$ time.
+Furthermore, we can embed $T$ in the upper-left corner of a circulant matrix $C$ of twice its size, permitting MVM in $O(n\log n)$ time.
 This approach has been used for fast inference in single-output GP time series with uniformly spaced inputs \citep{cunningham2008fast}.
@@
 \end{align}
-This approximation improves as the size of $K$ increases, so, as in other work \citep{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
+As in other work \citep{cutajar2016preconditioning}, we let $\tbr\sim\Unif\{\pm 1\}$ and take a fixed number $N$ samples from $\tbr$.
+We refer the reader to existing literature regarding the convergence of the Hutchinson trace estimator \citep{avron2011randomized}.
 
@@
 \caption{Trace of (a) the number of MVMs that \textsc{minres} must perform to invert $K^{-1}\Ty$ and (b) $\mcL, \norm{\nabla\mcL}$ given $\bsth$ at each optimization iteration for a GP applied to the dataset in Sec.~\ref{fx2007-results}.
-In (b), in green, we have 20\% of the rolling maximum $\infty$-norm of previous gradients.
+In (b), in green, we have 20\% of the roolling maximum $\infty$-norm of previous gradients.
 }
@@
 Assume \textsc{minres} is the inversion routine.
-We also assume we have access to linear operators $\partial_{\theta_j}$, representing matrices $\partial_{\theta_j}\tilde{K}$.} \label{alg:grad}
+We also assume we have access to linear operators $\partial_{\theta_j}\tilde{K}$.
+Note that matrix structure and, in turn, MVM efficiency in $\tilde{K}$ is preserved under pointwise differentiation.} \label{alg:grad}
 \begin{algorithmic}[1]
-  \Procedure{LLGP}{$\tilde{K}$, $\Ty$, $N$, $\{\partial_{\theta_j}\}$}
+  \Procedure{LLGP}{$\tilde{K}$, $\Ty$, $N$, $\{\partial_{\theta_j}\tilde K\}_j$}
   \State $R\gets\idx{\tbr_i}{i=1}{N}$, sampling $\tbr\sim\Unif\{\pm 1\}$.
@@
 \For{$\theta_j$ in $\bsth$}\Comment{Compute $\partial_{\theta_j}\mcL$}
-\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot \partial_{\theta_j}(\tbr_i)$\Comment{Eq.~\ref{eq:trace}}
-\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot \tilde{K}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
+\State $t\gets \frac{1}{N}\sum_{i=1}^{N}\pa{K^{-1}\tbr_i}\cdot \partial_{\theta_j}\tilde{K}(\tbr_i)$\Comment{Eq.~\ref{eq:trace}}
+\State $g_j\gets \frac{1}{2}\pa{K^{-1}\Ty}\cdot\partial_{\theta_j} \tilde{K}\pa{K^{-1}\Ty}-\frac{1}{2}t$\Comment{Eq.~\ref{llgradient}}
 \EndFor
@@
 We consider different representations of $K_{{U'},{U'}}$ (Eq.~\ref{lmcski}) to reduce the memory and runtime requirements for performing the multiplication $K_{{U'},{U'}}\Tz$ in the following sections.
+For a given $D,Q,R$, asymptotic complexities for MVMs are described for each of the representations below.
+By choosing the representation with the smallest complexity, we can minimize runtime in a variety of scenarios (Tab.~\ref{asymp}).\footnote{The supplement further discusses automatic selection of representation in practice.}
 
@@
 In \textsc{sum}, we represent $K_{{U'},{U'}}$ with a $Q$-length list.
-At each index $q$, $B_q$ is a dense matrix of order $D$ and $K_q$ is a BTTB matrix of order $m$, represented using only the top row.
+At each index $q$, $B_q$ is a dense, unstructured matrix of order $D$ and $K_q$ is a BTTB matrix of order $m$, represented using only the top row.
 In turn, multiplication $K_{{U'},{U'}}\Tz$ is performed by multiplying each matrix in the list with $\Tz$ and summing the results.
@@
 Thus, \textsc{slfm} represents $K_{{U'},{U'}}$ as the sum of two block diagonal matrices of block order $QR$ and $D$, where each block is a BTTB order $m$ matrix; thus, MVMs run in $O((QR + D)m\log m)$.
-
-Note that \textsc{bt} and \textsc{slfm} each have a faster run time than the other depending on whether $D^2>QR$.
-An algorithm that uses this condition to decide between representations can minimize runtime (Tab.~\ref{asymp}).
 
@@
 LLGP was run with LMC set to $Q=1$, $R=2$, and $234$ interpolating points.
-COGP used a $Q=2$ kernel with $100$ inducing points.}
+COGP used a $Q=2$ kernel with $100$ inducing points.
+\cite{nguyen2014collaborative} also evaluate independent and convolved GPs (IGPs, CGPs) on FX2007 in Tbl.~2, but LLGP beats the reported figures.}
 \label{fx2007-tbl}
@@
 {\includegraphics[width=\textwidth]{fx2007graph.pdf}}
-% TODO add legend
+
 \caption{Test outputs for the FX2007 dataset.
@@
 Both LLGP and COGP trained the SLFM model.
-We show LLGP with $500$ and $1000$ interpolating points and COGP with $200$ inducing points.}
+We show LLGP with $500$ and $1000$ interpolating points and COGP with $200$ inducing points.
+Again, analogous reported results for all methods (COGP, IGP, CGP) in \cite{nguyen2014collaborative} Tbl.~3 are beaten.}
 \label{weather-tbl}

### 2017-11-26 paper/supplement.tex
updated supplement, new images in paper/
added aistats review comments (pt 1)

@@
 
-LLGP was implemented in Python 3 from the Anaconda, which offered an Intel MKL-linked scipy \citep{scipy}.
+LLGP was implemented in Python 3 from Anaconda, which offered an Intel MKL-linked scipy \citep{scipy}.
 The code made heavy use of other packages, namely climin \citep{climin}, GPy \citep{gpy}, and paramz \citep{paramz}.
@@
 For inverting systems, all computations were carried out until the residual $\ell_2$ norm was at most $10^{-4}$.
+Altogether, this validates a rule based on asymptotic complexity, where \textsc{sum} is chosen for $Q=1$, \textsc{bt} when $QR<D^2$, and \textsc{slfm} otherwise.
 \begin{table*}[!ht]

### 2017-11-26 benchmarks/benchlib/standard_tester.py
plot synth matrix

@@ -147,6 +147,7 @@ def weather():
 
     return xss, yss, test_xss, test_yss, sensors
 
+
 def synth():
     datadir = _download_own_data()
     xss = np.load(os.path.join(datadir, 'synth', 'xss.npy'))
@@ -235,6 +236,9 @@ def nlpd(test_yss, pred_yss, pred_vss):
 def bench_runlmc(num_runs, m, xss, yss, test_xss, test_yss,
                  kgen, rgen, slfmgen, indepgen, optimizer_opts, **kwargs):
     times, smses, nlpds = [], [], []
+    return_lmc = 'return_lmc' in kwargs
+    if return_lmc:
+        del kwargs['return_lmc']
     for i in range(num_runs):
         fk = FunctionalKernel(
             D=len(xss),
@@ -257,6 +261,8 @@ def bench_runlmc(num_runs, m, xss, yss, test_xss, test_yss,
 
     points = [times, smses, nlpds]
     stats = [(np.mean(x), np.std(x) / np.sqrt(len(x))) for x in points]
+    if return_lmc:
+        return lmc
     return stats
 
 
@@ -377,7 +383,7 @@ def cogp_weather(num_runs, M, nthreads):
     cogp_mu = pd.read_csv(TMP + '/cogp-weather-mu{}{}'.format(num_runs, M),
                           header=None, names=test_fx)
     cogp_var = pd.read_csv(TMP + '/cogp-weather-var{}{}'.format(num_runs, M),
-                          header=None, names=test_fx)
+                           header=None, names=test_fx)
 
     stats = [(m_time, se_time), (m_smse, se_smse), (m_nlpd, se_nlpd)]
     return stats, cogp_mu, cogp_var
@@ -443,6 +449,7 @@ def gen_random_k():
               for i in range(5)]
     return FunctionalKernel(
         D=5, lmc_kernels=[], lmc_ranks=[], slfm_kernels=ks, indep_gp=indeps)
+
 
 def synth_gen():
     """Kernel generators associated for gen_random_k()"""

### 2017-11-26 benchmarks/synth/synth.py

@@ -47,12 +47,20 @@ def main():
         cogp_runs = 1
         cogp_colnames = ['COGP']
     else:
+        import h5py
         kgen, rgen, slfmgen, indepgen = synth_gen()
         xss, yss, test_xss, test_yss = synth()
-        stats = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
-                             test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
-                             tolerance=1e-3)  # reduce tol a bit b/c of large problem size
+        stats, lmc = bench_runlmc(llgp_runs, interpolating_points, xss, yss, test_xss,
+                                  test_yss, kgen, rgen, slfmgen, indepgen, {}, max_procs=nthreads,
+                                  # reduce tol a bit b/c of large problem size
+                                  tolerance=1e-3, return_lmc=True)
+        approx = lmc.kernel.K.as_numpy()
+        exact = lmc.K()
+        with h5py.File('llgp-mats.h5', 'w') as f:
+            f.create_dataset('approx', data=approx)
+            f.create_dataset('exact', data=exact)
         dump(stats, 'llgp_stats')
+        #
         # stats = load('llgp_stats')
         all_stats = [stats]
         colnames = ['LLGP']

### 2021-09-06 paper/paper.tex
final template gen

--- 2017-10-13 paper/paper.tex
@@
 \documentclass[twoside]{article}
-\usepackage{aistats2018}
+\usepackage[arxiv]{aistats2018}
 % Camera ready: \usepackage[accepted]{aistats2018}
 \newif\ifanonymized
-\anonymizedtrue
+\anonymizedfalse
 % Camera ready: \anonymizedfalse
@@
 \subsubsection*{Acknowledgments}
-(TODO) Thanks to Princeton, Cal, Prof. Joseph Gonzalez, Prof.Ion Stoica, Prof. Andrew Wilson, Prof. John Cunningham.
+The authors would like to thank Princeton University and University of California, Berkeley, for providing the computational resources necessary to evaluate our method.
+Further, the authors owe thanks to Professors Joseph Gonzalez, Ion Stoica, Andrew Wilson, and John Cunningham for reviewing drafts of the paper and offering insightful advice about directions to explore.
 \fi

### 2021-09-06 arxiv-tar.sh

@@ -1,3 +1,6 @@
 #!/bin/bash
 
+mkdir -p paper/anc
+cp paper/supplement.pdf paper/anc
 tar --exclude="paper/*.eps" --exclude="paper/.gitignore" --exclude="*.sh" --exclude="paper/*.out" --exclude="paper/supplement*" --exclude="paper/paper.pdf" -czvf paper.tgz paper/
+rm -rf paper/anc
